---
title: "Mathematics You Need for PK/PD"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Work with exponential growth and decay functions
- Use logarithms for linearization and half-life calculations
- Understand ordinary differential equations (ODEs) in PK contexts
- Perform dimensional analysis and unit checking in R
- Interpret mathematical parameters biologically

---

## Exponential Functions: The Heart of PK

### Why Exponentials?

Most PK processes follow first-order kinetics — the rate of change is proportional to the amount present. This leads naturally to exponential functions:

$$\frac{dC}{dt} = -k \cdot C \quad \Rightarrow \quad C(t) = C_0 \cdot e^{-kt}$$

where $C$ is concentration, $k$ is the rate constant, and $t$ is time.

```{r}
library(ggplot2)

# Comparing different elimination rate constants
time <- seq(0, 24, by = 0.1)
C0 <- 100

df <- data.frame(
  time = rep(time, 4),
  conc = c(C0 * exp(-0.1 * time),  # slow elimination
           C0 * exp(-0.3 * time),  # moderate
           C0 * exp(-0.7 * time),  # fast
           C0 * exp(-1.5 * time)), # very fast
  k = factor(rep(c("k=0.1", "k=0.3", "k=0.7", "k=1.5"), each = length(time)))
)

ggplot(df, aes(time, conc, color = k)) +
  geom_line(linewidth = 1) +
  labs(title = "Exponential Decay: Effect of Rate Constant",
       x = "Time (hours)", y = "Concentration (mg/L)",
       color = "Elimination Rate") +
  scale_color_manual(values = c("#0B6E6B", "#2E8B7E", "#C8963E", "#C44536")) +
  theme_minimal()
```

---

## Logarithms and Half-Life

### The Power of Log Transformation

Taking the natural log of the exponential decay equation gives a linear relationship:

$$\ln C(t) = \ln C_0 - k \cdot t$$

This means a plot of $\ln C$ vs $t$ is a straight line with slope $-k$. This is the basis for **terminal slope estimation** in PK.

```{r}
# Log-linear plot for half-life determination
time <- seq(0, 24, by = 0.5)
C0 <- 100
k <- 0.3
conc <- C0 * exp(-k * time)

df <- data.frame(time, conc, log_conc = log(conc))

# Find half-life from data
t_half <- log(2) / k

p1 <- ggplot(df, aes(time, conc)) +
  geom_point(color = "#0B6E6B") +
  geom_line(color = "#0B6E6B") +
  labs(title = "Linear Scale", x = "Time (h)", y = "Concentration (mg/L)") +
  theme_minimal()

p2 <- ggplot(df, aes(time, log_conc)) +
  geom_point(color = "#C44536") +
  geom_smooth(method = "lm", se = TRUE, color = "#C44536") +
  labs(title = "Log-Linear Scale",
       subtitle = paste0("Slope = -k = ", -k, ", Half-life = ", round(t_half, 2), " h"),
       x = "Time (h)", y = "ln(Concentration)") +
  theme_minimal()

library(patchwork)
p1 + p2
```

### Half-Life Formula

$$t_{1/2} = \frac{\ln(2)}{k}$$

This is one of the most important relationships in PK. The half-life tells you how long it takes for the drug concentration to fall by 50%.

```{r}
# How many half-lives to reach steady state or complete elimination?
n_halflives <- 1:7
pct_eliminated <- (1 - 0.5^n_halflives) * 100

data.frame(
  HalfLives = n_halflives,
  Pct_Eliminated = round(pct_eliminated, 1),
  Pct_Remaining = round(100 - pct_eliminated, 1)
)
```

::: {.key-box}
### Key Takeaway

- After **3.3 half-lives**: ~90% eliminated
- After **5 half-lives**: ~97% eliminated (commonly used for steady state)
- After **7 half-lives**: >99% eliminated
:::

---

## Ordinary Differential Equations in PK

### What Is a Differential Equation?

A differential equation describes how a quantity changes over time. In PK, we use ODEs to model how drug amounts change in compartments.

The simplest PK model (one-compartment IV bolus):

$$\frac{dA}{dt} = -k \cdot A$$

where $A$ is the amount of drug. This says: "the rate of change of drug amount is proportional to the amount present."

### Solving ODEs in R

```{r}
# Solving ODEs with rxode2
library(rxode2)

# Define a one-compartment IV bolus model
mod1 <- rxode2({
  d/dt(central) <- -ke * central
  conc <- central / V
})

# Simulate
params <- c(ke = 0.3, V = 10)
inits  <- c(central = 100)  # 100 mg dose
times  <- seq(0, 24, by = 0.1)
ev     <- et(times)

sim <- rxSolve(mod1, params, ev, inits = inits)

plot(sim$time, sim$conc, type = "l", col = "#0B6E6B", lwd = 2,
     xlab = "Time (h)", ylab = "Concentration (mg/L)",
     main = "ODE Solution: One-Compartment IV Bolus")
```

### The Analytical vs Numerical Approach

For simple models, we can write the analytical solution directly: $C(t) = C_0 e^{-kt}$. For complex models with nonlinearities or multiple compartments, we solve ODEs numerically. R handles both approaches.

---

## Units and Dimensional Analysis

### Why Units Matter

In PK/PD, incorrect units are a leading source of errors. Always track your units:

| Parameter | Common Unit | Alternative |
|-----------|-------------|-------------|
| Dose | mg | mg/kg |
| Concentration | mg/L | μg/mL, ng/mL |
| Volume of distribution | L | L/kg |
| Clearance | L/h | mL/min |
| Half-life | h | min |
| AUC | mg·h/L | μg·h/mL |

```{r}
# Dimensional checking in R
check_units <- function(dose_mg, conc_mg_L, V_L) {
  # V = dose / C0
  computed_V <- dose_mg / conc_mg_L
  cat("Dose:", dose_mg, "mg\n")
  cat("C0:", conc_mg_L, "mg/L\n")
  cat("Computed V:", computed_V, "L\n")
  cat("Expected V:", V_L, "L\n")
  cat("Match:", ifelse(abs(computed_V - V_L) < 1e-6, "✓", "✗ MISMATCH - check units!\n"))
}

check_units(dose_mg = 500, conc_mg_L = 25, V_L = 20)
```

---

## Scaling Between Species: Allometry

Allometric scaling is used to extrapolate PK parameters between species based on body weight:

$$Y = a \cdot W^b$$

where $Y$ is the PK parameter, $W$ is body weight, and $b$ is the allometric exponent.

```{r}
# Allometric scaling example
body_weights <- c(0.02, 0.25, 2.5, 10, 70)  # kg: mouse, rat, rabbit, dog, human
clearance <- 2.5 * body_weights^0.75         # allometric scaling for CL

df_scale <- data.frame(
  Species = c("Mouse", "Rat", "Rabbit", "Dog", "Human"),
  Weight_kg = body_weights,
  CL_L_h = round(clearance, 2)
)

ggplot(df_scale, aes(Weight_kg, CL_L_h, label = Species)) +
  geom_point(color = "#C44536", size = 3) +
  geom_line(color = "#0B6E6B") +
  geom_text(vjust = -1, size = 3.5) +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Allometric Scaling of Clearance",
       subtitle = "CL = 2.5 × W^0.75",
       x = "Body Weight (kg, log scale)", y = "Clearance (L/h, log scale)") +
  theme_minimal()
```

---

::: {.exercise-box}
### Exercises

1. A drug has a half-life of 4 hours. What is its elimination rate constant $k$?
2. After how many half-lives will 95% of a drug be eliminated?
3. Convert a clearance of 120 mL/min to L/h.
4. A drug follows $C(t) = 50 \cdot e^{-0.2t}$. Find $C_0$ and $t_{1/2}$.
5. Write the ODE for a one-compartment model with IV infusion at rate $R$.
:::

::: {.mistakes-box}
### Common Mistakes

- **Confusing rate constant $k$ and half-life**: $k = \ln(2) / t_{1/2}$, not $1 / t_{1/2}$
- **Using log10 instead of ln for half-life**: $t_{1/2} = \ln(2)/k$ (natural log, not log10)
- **Mixing units**: mg vs μg is a 1000× error
- **Assuming all processes are first-order**: Some drugs show saturable (Michaelis–Menten) kinetics
:::

::: {.key-box}
### Key Takeaway

PK mathematics is built on exponentials, logarithms, and differential equations. Master the relationship $C(t) = C_0 e^{-kt}$, the log-linear transformation, and the half-life formula — these appear in every chapter that follows. Always check your units.
:::

