---
title: "R Fundamentals and Workflow for PK/PD"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Set up a complete R environment for PK/PD analysis
- Understand R data structures through a pharmacometric lens
- Master the "Tidyverse" workflow for data cleaning and transformation
- Handle common PK/PD data hurdles (BLQ values, units, and pivoting)
- Build sophisticated, faceted `ggplot2` visualizations
- Organize reproducible PK/PD analysis projects

---

## Setting Up Your R Environment

### Installing R and RStudio

1. Download R from [cran.r-project.org](https://cran.r-project.org)
2. Download RStudio (or VS Code with the R extension).
3. **The PK/PD Ecosystem:** Unlike other fields, PK/PD in R relies on a specific set of high-performance packages.

```{r}
#| eval: false
install.packages(c(
  "tidyverse",  # Data manipulation & plotting (dplyr, ggplot2, tidyr)
  "PKNCA",      # Industrial-strength Noncompartmental Analysis
  "rxode2",     # Fast ODE simulation
  "nlmixr2",    # Nonlinear Mixed Effects Modeling
  "mrgsolve",   # Simulation for clinical trial design
  "patchwork",  # Combining multiple plots
  "janitor",    # Cleaning messy column names
  "here"        # Robust file paths
))
```

### Recommended Project Structure

```
PKPD_Project/
├── data/            # Raw and processed data files
├── R/               # Custom R functions
├── scripts/         # Analysis scripts
├── output/          # Generated tables and figures
├── reports/         # Quarto/R Markdown reports
├── renv/            # Package version management (optional)
└── PKPD_Project.Rproj
```

---

## Importing PK/PD Data

### Common Data Formats in PK/PD

PK data typically has a "long" structure with columns for subject ID, time, concentration, and dose information.

```{r}
library(dplyr)
library(readr)
library(ggplot2)

# Simulate a typical PK dataset
set.seed(123)
n_subjects <- 6
times <- c(0, 0.25, 0.5, 1, 2, 4, 8, 12, 24)

pk_data <- expand.grid(
  ID = 1:n_subjects,
  TIME = times
) %>%
  mutate(
    DOSE = 100,
    CONC = (DOSE / 10) * exp(-0.3 * TIME) * rlnorm(n(), meanlog = 0, sdlog = 0.2),
    .by = ID
  )

# View the first few rows
head(pk_data, 12)
```

### Importing from Files

```{r}
#| eval: false
# Import CSV
pk_data <- read_csv("data/pk_concentrations.csv")

# Import Excel
library(readxl)
pk_data <- read_excel("data/pk_data.xlsx", sheet = "Concentrations")

# Import NONMEM-style data
pk_data <- read_table("data/pk_data.txt")
```

### Data Cleaning and Reshaping

```{r}
# Check for missing values
colSums(is.na(pk_data))

# Convert columns to correct types
pk_data <- pk_data %>%
  mutate(
    ID   = as.factor(ID),
    TIME = as.numeric(TIME),
    CONC = as.numeric(CONC)
  )

# Reshape wide → long (if needed)
pk_wide <- pk_data %>%
  tidyr::pivot_wider(names_from = TIME, values_from = CONC,
                     names_prefix = "T_")

# Reshape long → wide
pk_long <- pk_wide %>%
  tidyr::pivot_longer(cols = starts_with("T_"),
                      names_to = "TIME",
                      values_to = "CONC",
                      names_prefix = "T_") %>%
  mutate(TIME = as.numeric(TIME))
```

---

## Plotting Concentration-Time Profiles

### Individual Profiles

```{r}
# Individual concentration-time plots
ggplot(pk_data, aes(TIME, CONC, group = ID, color = ID)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.8) +
  labs(title = "Individual Concentration-Time Profiles",
       subtitle = "IV Bolus, 100 mg dose",
       x = "Time (hours)", y = "Concentration (mg/L)",
       color = "Subject") +
  theme_minimal()
```

### Mean ± SD Profile

```{r}
# Summary statistics and mean ± SD plot
pk_summary <- pk_data %>%
  summarise(
    Mean = mean(CONC),
    SD   = sd(CONC),
    N    = n(),
    .by = TIME
  )

ggplot(pk_summary, aes(TIME, Mean)) +
  geom_ribbon(aes(ymin = Mean - SD, ymax = Mean + SD),
              fill = "#0B6E6B", alpha = 0.2) +
  geom_point(color = "#0B6E6B", size = 2) +
  geom_line(color = "#0B6E6B", linewidth = 1) +
  labs(title = "Mean Concentration-Time Profile (± SD)",
       x = "Time (hours)", y = "Concentration (mg/L)") +
  theme_minimal()
```

### Semi-Log Plot (Standard PK View)

```{r}
ggplot(pk_data, aes(TIME, CONC, group = ID, color = ID)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(title = "Semi-Log Concentration-Time Profiles",
       x = "Time (hours)", y = "Concentration (mg/L, log scale)",
       color = "Subject") +
  theme_minimal()
```

---

## Plotting Dose-Response Data

```{r}
# Simulate dose-response data
doses <- c(0.1, 0.3, 1, 3, 10, 30, 100, 300)
n_per_dose <- 5

set.seed(456)
dr_data <- expand.grid(DOSE = doses, REP = 1:n_per_dose) %>%
  mutate(
    Emax = 100,
    EC50 = 10,
    EFFECT = (Emax * DOSE) / (EC50 + DOSE) + rnorm(n(), 0, 5)
  )

ggplot(dr_data, aes(DOSE, EFFECT)) +
  geom_point(color = "#C44536", alpha = 0.6, size = 2) +
  scale_x_log10() +
  labs(title = "Dose-Response Data",
       x = "Dose (mg, log scale)", y = "Effect (% of maximum)") +
  theme_minimal()
```

---

## Reproducible Scripts and Version Control

### Best Practices for PK/PD Scripts

1. **Start every script with a header** describing purpose, author, and date
2. **Load all packages at the top**
3. **Use relative paths** (not absolute paths like `C:/Users/...`)
4. **Comment your code** — explain *why*, not *what*
5. **Separate data import, analysis, and visualization**
6. **Use version control** (Git) for all scripts

```{r}
# Example script header
# ============================================================
# Script: 01_nca_analysis.R
# Purpose: Noncompartmental analysis of Study PK-2024-001
# Author:  Your Name
# Date:    2024-06-01
# Packages: PKNCA, dplyr, ggplot2
# Input:   data/raw/study001_conc.csv
# Output:  output/nca_parameters.csv, output/figures/conc_time.pdf
# ============================================================
```

---

## Creating a Reproducible PK/PD Workflow

```{r}
# A minimal reproducible PK analysis workflow

# 1. Load libraries
library(PKNCA)
library(dplyr)
library(ggplot2)

# 2. Import data
# pk_data <- read_csv("data/pk_data.csv")

# 3. Set up NCA intervals
# my_intervals <- data.frame(start = 0, end = 24, ..., cmax = TRUE, ...)

# 4. Calculate NCA parameters
# nca_results <- pk.nca(my_data, my_intervals)

# 5. Summarize and export
# nca_summary <- summary(nca_results)
# write_csv(nca_summary, "output/nca_results.csv")

# 6. Visualize
# plot(nca_results)
```

---

## Handling Missing Data in PK/PD

Missing concentration values are common in clinical PK studies. The appropriate handling strategy depends on **why** the data are missing.

### The MCAR/MAR/MNAR Framework

| Mechanism | Definition | Test/Clue | Action |
|---|---|---|---|
| **MCAR** (Missing Completely at Random) | Missingness unrelated to any variable | No pattern in missing vs observed | Mean/median imputation OK |
| **MAR** (Missing at Random) | Missingness explained by observed covariates | Pattern linked to recorded variables (e.g., more missing at later times) | Regression imputation, multiple imputation (`mice`), or MLE-based modeling |
| **MNAR** (Missing Not at Random) | Missingness depends on the unobserved value itself (e.g., samples below LLOQ) | Plausible mechanism: high CL patients drop out, BLQ samples | Selection/pattern-mixture models + sensitivity analysis; **no method is unbiased** |

```{r}
# Example: Identify and handle missing data patterns
library(dplyr)

# Simulate PK data with different missingness types
set.seed(42)
pk_missing <- data.frame(
  ID = rep(1:10, each = 8),
  TIME = rep(c(0.5, 1, 2, 4, 6, 8, 12, 24), 10),
  CONC = 25 * exp(-0.2 * rep(c(0.5, 1, 2, 4, 6, 8, 12, 24), 10)) * 
          exp(rnorm(80, 0, 0.12))
)

# Introduce MCAR: random 10% missing
mcar_idx <- sample(1:80, 8)
pk_missing$CONC_MCAR <- pk_missing$CONC
pk_missing$CONC_MCAR[mcar_idx] <- NA

# Introduce MAR: more missing at later time points
mar_prob <- 0.1 * pk_missing$TIME / 24
mar_idx <- which(runif(80) < mar_prob)
pk_missing$CONC_MAR <- pk_missing$CONC
pk_missing$CONC_MAR[mar_idx] <- NA

# Summarize missingness
missing_summary <- pk_missing %>%
  summarise(
    MCAR_missing = sum(is.na(CONC_MCAR)),
    MAR_missing  = sum(is.na(CONC_MAR)),
    .by = TIME
  )

cat("Missing by time point:\n")
print(missing_summary)

# Simple imputation: last observation carried forward (LOCF)
pk_missing <- pk_missing %>%
  group_by(ID) %>%
  tidyr::fill(CONC_MCAR, CONC_MAR, .direction = "down") %>%
  ungroup()
```

::: {.key-box}
### BLQ (Below Limit of Quantification) Handling

BLQ values are a special case of **MNAR** — the value is missing *because* it's low. Common approaches:

- **LLOQ/2 substitution** — set BLQ = LLOQ/2 (simple, common in NCA)
- **M3 method** — likelihood-based approach that treats BLQ as left-censored observations
- **Imputation** — multiple imputation with left-censoring constraint

For NCA, LLOQ/2 is standard. For population PK, likelihood-based methods (M3) are preferred as they correctly propagate uncertainty.
:::

---

::: {.exercise-box}
### Exercises

1. Create a simulated PK dataset with 10 subjects, 8 time points, and 2 dose groups. Plot individual and mean profiles.
2. Import a CSV file of your choice and reshape it from wide to long format.
3. Create a semi-log plot with proper axis labels and provide a caption about what the linear terminal phase indicates.
4. Write a reusable R function that calculates $k_e$ from two concentration-time points.
5. Set up a project directory structure following the recommended layout.
:::

::: {.mistakes-box}
### Common Mistakes

- **Using absolute file paths** — scripts break when shared or moved
- **Not checking for BLOQ (Below Limit of Quantification) values** — they need special handling
- **Plotting concentration on linear scale** — always check the semi-log view
- **Forgetting to convert time units** (hours → minutes, days → hours)
- **Overwriting raw data** — always keep original files unchanged
:::

::: {.key-box}
### Key Takeaway

A well-organized R environment with proper project structure, version control, and reproducible scripts is the foundation of reliable PK/PD analysis. Master data import, cleaning, and visualization in R before moving to modeling.
:::

