---
title: "Descriptive PK and Noncompartmental Analysis"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Structure PK data for noncompartmental analysis
- Compute AUC using the trapezoidal rule
- Estimate $C_{max}$, $T_{max}$, terminal half-life, and other NCA parameters
- Calculate clearance, volume of distribution, and bioavailability
- Apply NCA to both IV and extravascular data
- Understand the strengths and limitations of NCA

---

## What Is Noncompartmental Analysis?

**Noncompartmental analysis (NCA)** estimates PK parameters directly from concentration-time data without assuming a specific compartmental model. It relies on:

1. **The trapezoidal rule** for AUC computation
2. **Log-linear regression** of the terminal phase for $k_e$ and $t_{1/2}$
3. **Moment analysis** for mean residence time (MRT)

NCA is the most commonly used method in early-phase clinical trials and bioequivalence studies because it makes minimal assumptions.

---

## Data Structure for NCA

A proper PK dataset for NCA requires:

```{r}
library(dplyr)
library(ggplot2)
library(PKNCA)

# Example: Single subject, single dose, IV bolus
pk_iv <- data.frame(
  TIME = c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 24),
  CONC = c(12.3, 9.8, 8.1, 6.4, 5.7, 4.9, 3.8, 3.0, 1.9, 1.2, 0.5, 0.1)
)

# Required: TIME and CONC columns
# Optional but recommended: ID (subject), DOSE, ROUTE, and grouping variables

ggplot(pk_iv, aes(TIME, CONC)) +
  geom_point(size = 3, color = "#0B6E6B") +
  geom_line(color = "#0B6E6B") +
  labs(title = "Concentration-Time Profile for NCA",
       x = "Time (hours)", y = "Concentration (mg/L)") +
  theme_minimal()
```

---

## Computing AUC: The Trapezoidal Rule

### Linear Trapezoidal Rule

The area under the curve between two time points is approximated as a trapezoid:

$$AUC_{t_1 \to t_2} = \frac{(C_1 + C_2)}{2} \cdot (t_2 - t_1)$$

Total AUC from 0 to last time point:

$$AUC_{0 \to t_{last}} = \sum_{i=1}^{n-1} \frac{(C_i + C_{i+1})}{2} \cdot (t_{i+1} - t_i)$$

```{r}
# Manual AUC calculation using the trapezoidal rule
calc_auc_linear <- function(time, conc) {
  n <- length(time)
  auc <- 0
  for (i in 1:(n - 1)) {
    auc <- auc + (conc[i] + conc[i + 1]) / 2 * (time[i + 1] - time[i])
  }
  return(auc)
}

AUC_last <- calc_auc_linear(pk_iv$TIME, pk_iv$CONC)
cat("AUC(0 to last):", round(AUC_last, 3), "mg·h/L\n")
```

### Extrapolation to Infinity

$$AUC_{0 \to \infty} = AUC_{0 \to t_{last}} + \frac{C_{last}}{k_e}$$

where $k_e$ is the terminal elimination rate constant.

---

## Estimating the Terminal Elimination Rate

### Log-Linear Regression

The terminal phase is identified by selecting the last 3–5 points on the log-linear plot that form a straight line.

```{r}
# Identify terminal phase and compute k_e
pk_iv <- pk_iv %>%
  mutate(log_conc = log(CONC))

# Use last 4 points for terminal slope
terminal_idx <- (nrow(pk_iv) - 3):nrow(pk_iv)
terminal_data <- pk_iv[terminal_idx, ]

# Linear regression on log-transformed concentrations
lm_terminal <- lm(log_conc ~ TIME, data = terminal_data)
k_e <- -coef(lm_terminal)[2]
t_half <- log(2) / k_e

cat("Elimination rate constant (ke):", round(k_e, 4), "1/h\n")
cat("Half-life (t1/2):", round(t_half, 2), "hours\n")
```

```{r}
# Visualize terminal slope estimation
ggplot(pk_iv, aes(TIME, CONC)) +
  geom_point(size = 2) +
  geom_line() +
  scale_y_log10() +
  # Highlight terminal phase points
  geom_point(data = terminal_data, aes(TIME, CONC),
             color = "#C44536", size = 3) +
  labs(title = "Terminal Phase Identification",
       subtitle = paste0("Red points used for ke estimation: ke = ", round(k_e, 3), " 1/h"),
       x = "Time (hours)", y = "Concentration (mg/L, log scale)") +
  theme_minimal()
```

---

## Computing Key NCA Parameters

| Parameter | Formula | Interpretation |
|-----------|---------|---------------|
| $C_{max}$ | Maximum observed concentration | Peak exposure |
| $T_{max}$ | Time of $C_{max}$ | Absorption rate indicator |
| $AUC_{0 \to \infty}$ | $AUC_{last} + C_{last}/k_e$ | Total drug exposure |
| $t_{1/2}$ | $\ln(2)/k_e$ | Elimination half-life |
| $CL$ (IV) | $Dose / AUC_{0 \to \infty}$ | Volume cleared per unit time |
| $V_d$ (IV) | $Dose / (k_e \cdot AUC_{0 \to \infty})$ | Apparent distribution volume |
| $MRT$ | $AUMC / AUC$ | Mean residence time |

```{r}
# Complete NCA function
calculate_nca <- function(time, conc, dose, route = "IV") {
  # AUC to last
  auc_last <- calc_auc_linear(time, conc)

  # Terminal slope (last 4 points)
  n <- length(time)
  terminal_idx <- max(1, n - 3):n
  lm_term <- lm(log(conc[terminal_idx]) ~ time[terminal_idx])
  ke <- -coef(lm_term)[2]
  t_half <- log(2) / ke

  # Extrapolated AUC
  auc_inf <- auc_last + conc[n] / ke

  # Cmax, Tmax
  cmax <- max(conc)
  tmax <- time[which.max(conc)]

  # CL, Vd
  CL <- dose / auc_inf
  Vd <- CL / ke

  # AUMC and MRT
  auc_moment <- function(time, conc) {
    sum(diff(time) * (time[-1] * conc[-1] + time[-length(time)] * conc[-length(time)]) / 2)
  }
  aumc_last <- auc_moment(time, conc)
  aumc_inf <- aumc_last + time[n] * conc[n] / ke + conc[n] / (ke^2)
  MRT <- aumc_inf / auc_inf

  data.frame(
    Parameter = c("Cmax (mg/L)", "Tmax (h)", "AUC_last (mg·h/L)",
                  "AUC_inf (mg·h/L)", "ke (1/h)", "t_half (h)",
                  "CL (L/h)", "Vd (L)", "MRT (h)"),
    Value = round(c(cmax, tmax, auc_last, auc_inf, ke, t_half, CL, Vd, MRT), 3)
  )
}

nca_results <- calculate_nca(pk_iv$TIME, pk_iv$CONC, dose = 100)
knitr::kable(nca_results, caption = "NCA Parameters for IV Bolus")
```

---

## Extravascular Dosing

For oral or other extravascular administration, we also need:

- **Bioavailability ($F$)**: Fraction of dose reaching systemic circulation
  $$F = \frac{AUC_{oral} \cdot D_{IV}}{AUC_{IV} \cdot D_{oral}}$$

- **Absorption rate constant ($k_a$)**: Estimated from the absorption phase

```{r}
# Simulate oral PK data
set.seed(42)
time_oral <- seq(0, 24, by = 0.25)
ka <- 1.0    # absorption rate constant
ke <- 0.3    # elimination rate constant
F  <- 0.8    # bioavailability
D  <- 100    # dose (mg)
V  <- 10     # volume (L)

conc_oral <- (F * D * ka) / (V * (ka - ke)) * (exp(-ke * time_oral) - exp(-ka * time_oral))
# Add some noise
conc_oral <- conc_oral + rnorm(length(conc_oral), 0, 0.05)
conc_oral[conc_oral < 0] <- 0

pk_oral <- data.frame(TIME = time_oral, CONC = conc_oral)

# Sample sparse time points for a realistic dataset
sample_times <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 24)
pk_oral_sparse <- pk_oral[pk_oral$TIME %in% sample_times, ]

ggplot(pk_oral, aes(TIME, CONC)) +
  geom_line(color = "#0B6E6B", alpha = 0.3) +
  geom_point(data = pk_oral_sparse, size = 3, color = "#C44536") +
  geom_line(data = pk_oral_sparse, color = "#C44536", linewidth = 1) +
  labs(title = "Oral Dosing: Concentration-Time Profile",
       subtitle = "Red points = typical sparse sampling schedule",
       x = "Time (hours)", y = "Concentration (mg/L)") +
  theme_minimal()
```

---

## NCA Using the PKNCA Package

```{r}
# PKNCA workflow (standard approach)
library(PKNCA)

# Prepare data
pk_oral_sparse <- pk_oral_sparse %>%
  mutate(Subject = 1)

# One row per dose (single dose at time 0)
dose_data <- data.frame(
  Subject   = 1,
  Dose      = 100,
  Dose_Time = 0
)

# Set up intervals for NCA
my_intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE,
  aucint.last = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, cl.obs = TRUE,
  vz.obs = TRUE
)

# Create PKNCAconc and PKNCAdose objects
conc_obj <- PKNCAconc(pk_oral_sparse, CONC ~ TIME | Subject)
dose_obj <- PKNCAdose(dose_data, Dose ~ Dose_Time | Subject)

# Combine and compute
data_obj <- PKNCAdata(conc_obj, dose_obj, intervals = my_intervals)
results <- pk.nca(data_obj)

# View results
summary(results)
```

---

## Strengths and Limitations of NCA

::: {.theory-box}
### Strengths
- Minimal assumptions about the underlying model
- Standardized, widely accepted in regulatory submissions
- Simple to compute and interpret
- Robust across diverse drugs and dosing routes

### Limitations
- Cannot predict concentrations at untested times or doses
- Sensitive to sampling schedule (miss the peak → biased $C_{max}$)
- Terminal phase identification can be subjective
- No mechanistic insight into drug behavior
- Limited for sparse data or complex dosing regimens
:::

::: {.exercise-box}
### Exercises

1. Calculate the AUC using both linear and log-linear trapezoidal rules. When would you prefer one over the other?
2. A drug has $C_{last}$ = 2.5 mg/L at 12 h, and the terminal slope gives $k_e$ = 0.15 1/h. What is $AUC_{0 \to \infty}$ if $AUC_{0 \to 12}$ = 85 mg·h/L?
3. Compute the relative bioavailability of an oral formulation ($AUC$ = 120 mg·h/L, dose = 50 mg) compared to IV ($AUC$ = 200 mg·h/L, dose = 25 mg).
4. How does the choice of terminal phase points affect $t_{1/2}$ and $AUC_{0 \to \infty}$? Demonstrate with the dataset.
:::

::: {.mistakes-box}
### Common Mistakes

- **Using too few points for terminal slope** — need at least 3 points after $C_{max}$
- **Including distribution-phase points in the terminal slope** — inflates $k_e$ and underestimates half-life
- **Extrapolating >20% of total AUC** — if $AUC_{extrap}$ > 20% of $AUC_{inf}$, half-life is unreliable
- **Using linear scale to identify terminal phase** — always use semi-log plot
- **Forgetting to account for BLOQ values** — they should be handled consistently
:::

::: {.key-box}
### Key Takeaway

NCA is the workhorse of early-phase PK analysis. Master the trapezoidal rule, terminal slope estimation, and the calculation of $CL$, $V_d$, and $t_{1/2}$. Always validate your terminal phase selection with semi-log plots and check that extrapolated AUC is ≤20% of total.
:::

