---
title: "The Two-Compartment and Multi-Compartment Models"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Understand why multi-compartment models are necessary
- Distinguish distribution and elimination phases
- Simulate and fit two-compartment models in R
- Interpret micro- and macro-constants
- Compare one- and two-compartment models
- Understand identifiability in multi-compartment models

---

## Why Multi-Compartment Models?

After IV bolus dosing, many drugs show a **biphasic decline** in concentration: a rapid initial drop (distribution phase) followed by a slower decline (elimination phase). A one-compartment model cannot capture this behavior.

```{r}
library(rxode2)
library(dplyr)
library(ggplot2)

# Compare one- vs two-compartment profiles
times <- seq(0, 24, by = 0.05)

# One-compartment
mod_1cmt <- rxode2({
  d/dt(central) <- -ke * central
  conc <- central / V
})

# Two-compartment
mod_2cmt <- rxode2({
  d/dt(central)    <- -k10 * central - k12 * central + k21 * peripheral
  d/dt(peripheral) <-  k12 * central - k21 * peripheral
  conc <- central / V1
})

params_1 <- c(ke = 0.3, V = 15)
params_2 <- c(k10 = 0.3, k12 = 1.5, k21 = 0.8, V1 = 5)

event <- et(time = 0, amt = 100)

sim1 <- rxSolve(mod_1cmt, params_1, event %>% et(times))
sim2 <- rxSolve(mod_2cmt, params_2, event %>% et(times))

df <- bind_rows(
  as.data.frame(sim1) %>% mutate(Model = "One-Compartment"),
  as.data.frame(sim2) %>% mutate(Model = "Two-Compartment")
)

ggplot(df, aes(time, conc, color = Model)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(title = "One- vs Two-Compartment Models",
       subtitle = "Note the biphasic decline in the two-compartment model",
       x = "Time (hours)", y = "Concentration (mg/L, log scale)") +
  scale_color_manual(values = c("#0B6E6B", "#C44536")) +
  theme_minimal()
```

---

## Two-Compartment Model Structure

### Model Equations

$$\frac{dA_1}{dt} = -k_{10}A_1 - k_{12}A_1 + k_{21}A_2$$
$$\frac{dA_2}{dt} = k_{12}A_1 - k_{21}A_2$$
$$C_1(t) = \frac{A_1(t)}{V_1}$$

where:
- $A_1$, $A_2$: amounts in central and peripheral compartments
- $V_1$: volume of central compartment
- $k_{10}$: elimination rate constant from central compartment
- $k_{12}$: transfer rate from central to peripheral
- $k_{21}$: transfer rate from peripheral to central

### Analytical Solution

$$C(t) = A \cdot e^{-\alpha t} + B \cdot e^{-\beta t}$$

where $\alpha$ (distribution) and $\beta$ (elimination) are hybrid rate constants ($\alpha > \beta$), and $A$ and $B$ are hybrid coefficients.

```{r}
# Visualize the analytical solution components
alpha <- 2.8
beta  <- 0.25
A_coef <- 12
B_coef <- 8

times <- seq(0, 24, by = 0.05)
C_distribution <- A_coef * exp(-alpha * times)
C_elimination  <- B_coef * exp(-beta * times)
C_total <- C_distribution + C_elimination

df_comp <- data.frame(
  time = rep(times, 3),
  conc = c(C_total, C_distribution, C_elimination),
  Component = rep(c("Total", "Distribution (α)", "Elimination (β)"), each = length(times))
)

ggplot(df_comp, aes(time, conc, color = Component, linetype = Component)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(title = "Two-Compartment Model: Analytical Components",
       subtitle = paste0("C(t) = ", round(A_coef, 1), "·e^{-", alpha, "t} + ",
                         round(B_coef, 1), "·e^{-", beta, "t}"),
       x = "Time (hours)", y = "Concentration (mg/L, log scale)") +
  scale_color_manual(values = c("#0B6E6B", "#D4A843", "#C44536")) +
  theme_minimal()
```

---

## Micro- and Macro-Constants

| Parameter Type | Symbols | Meaning |
|----------------|---------|---------|
| **Micro-constants** | $k_{10}, k_{12}, k_{21}$ | Rate constants between compartments |
| **Macro-constants** | $A, B, \alpha, \beta$ | Hybrid coefficients and exponents |

### Conversion Between Micro and Macro

$$\alpha + \beta = k_{10} + k_{12} + k_{21}$$
$$\alpha \cdot \beta = k_{10} \cdot k_{21}$$
$$A = \frac{D(\alpha - k_{21})}{V_1(\alpha - \beta)}, \quad B = \frac{D(k_{21} - \beta)}{V_1(\alpha - \beta)}$$

```{r}
# Convert micro to macro constants
micro_to_macro <- function(k10, k12, k21, V1, dose) {
  sum_k <- k10 + k12 + k21
  prod_k <- k10 * k21

  # Alpha and beta are roots of: r^2 - sum_k * r + prod_k = 0
  disc <- sqrt(sum_k^2 - 4 * prod_k)
  alpha <- (sum_k + disc) / 2
  beta  <- (sum_k - disc) / 2

  A <- dose * (alpha - k21) / (V1 * (alpha - beta))
  B <- dose * (k21 - beta) / (V1 * (alpha - beta))

  c(alpha = alpha, beta = beta, A = A, B = B)
}

macro <- micro_to_macro(k10 = 0.3, k12 = 1.5, k21 = 0.8, V1 = 5, dose = 100)
macro
```

---

## Fitting a Two-Compartment Model

### Using nls() with the Analytical Solution

```{r}
# Simulate noisy two-compartment data
set.seed(42)
sample_times <- c(0.05, 0.1, 0.25, 0.5, 1, 2, 4, 6, 8, 12, 24)
true_macro <- micro_to_macro(0.3, 1.5, 0.8, 5, 100)

true_conc <- true_macro["A"] * exp(-true_macro["alpha"] * sample_times) +
             true_macro["B"] * exp(-true_macro["beta"] * sample_times)

obs_2cmt <- data.frame(
  time = sample_times,
  conc = true_conc * rlnorm(length(sample_times), 0, 0.1)
)

# Fit two-compartment model
fit_2cmt <- nls(conc ~ A * exp(-alpha * time) + B * exp(-beta * time),
                data = obs_2cmt,
                start = list(A = 15, alpha = 2.0, B = 10, beta = 0.2),
                lower = c(0, 0, 0, 0),
                algorithm = "port")

summary(fit_2cmt)
```

```{r}
# Plot observed vs fitted
coefs <- coef(fit_2cmt)
pred_times <- seq(0, 24, by = 0.05)
pred_conc <- coefs["A"] * exp(-coefs["alpha"] * pred_times) +
             coefs["B"] * exp(-coefs["beta"] * pred_times)

ggplot(obs_2cmt, aes(time, conc)) +
  geom_point(size = 3, color = "#C44536") +
  geom_line(data = data.frame(time = pred_times, conc = pred_conc),
            aes(time, conc), color = "#0B6E6B", linewidth = 1.2) +
  scale_y_log10() +
  labs(title = "Fitted Two-Compartment Model",
       x = "Time (hours)", y = "Concentration (mg/L, log scale)") +
  theme_minimal()
```

---

## Model Comparison: One vs Two Compartments

```{r}
# Fit one-compartment model for comparison
fit_1cmt <- nls(conc ~ C0 * exp(-ke * time),
                data = obs_2cmt,
                start = list(C0 = 20, ke = 0.3))

# Compare using AIC
aic_1cmt <- AIC(fit_1cmt)
aic_2cmt <- AIC(fit_2cmt)

comparison <- data.frame(
  Model = c("One-Compartment", "Two-Compartment"),
  AIC = round(c(aic_1cmt, aic_2cmt), 1),
  Parameters = c(2, 4)
)
knitr::kable(comparison, caption = "Model Comparison via AIC")
```

---

## Identifiability and Parameter Interpretation

::: {.theory-box}
### When One-Compartment Models Are Not Enough

- Distribution phase is clinically relevant (e.g., rapid onset drugs)
- Drug shows clear biphasic decline on semi-log plot
- Peripheral compartment has physiological meaning (e.g., tissue binding)
- Accurate prediction of early time points is needed
:::

### Common Issues

- **Parameter identifiability**: With sparse data, 4 parameters ($A, B, \alpha, \beta$) may not be uniquely estimable
- **Correlation**: $A$ and $\alpha$ are often highly correlated
- **Initial estimates**: Good starting values are critical; use graphical methods (method of residuals)

```{r}
# Method of residuals for initial estimates
# Step 1: Fit terminal phase to get B and beta
terminal_idx <- 6:length(sample_times)  # Choose terminal points
lm_term <- lm(log(obs_2cmt$conc[terminal_idx]) ~ obs_2cmt$time[terminal_idx])
B_init <- exp(coef(lm_term)[1])
beta_init <- -coef(lm_term)[2]

# Step 2: Calculate residuals and fit distribution phase
terminal_pred <- B_init * exp(-beta_init * sample_times)
residuals <- obs_2cmt$conc - terminal_pred
residuals <- residuals[residuals > 0]

cat("Initial estimates from method of residuals:\n")
cat("B =", round(B_init, 2), ", beta =", round(beta_init, 3), "\n")
```

---

::: {.exercise-box}
### Exercises

1. Simulate a two-compartment model with $k_{10}$ = 0.2, $k_{12}$ = 2.0, $k_{21}$ = 1.0, $V_1$ = 8 L. Plot on semi-log scale.
2. Convert the micro-constants from Exercise 1 to macro-constants ($A, B, \alpha, \beta$).
3. Fit a two-compartment model to the simulated data. How well do you recover the true parameters?
4. When would you choose a two-compartment model over a one-compartment model? Use AIC to compare.
5. Explain what happens to the concentration-time profile when $k_{12}$ is very large compared to $k_{21}$.
:::

::: {.mistakes-box}
### Common Mistakes

- **Over-parameterization**: A two-compartment model with sparse data may fit noise, not signal
- **Misidentifying phases**: The initial rapid decline may be due to distribution, not elimination
- **Using one-compartment model when distribution matters**: e.g., for drugs with rapid onset of action
- **Poor initial estimates**: nls() is sensitive to starting values; use method of residuals
:::

::: {.key-box}
### Key Takeaway

Multi-compartment models capture the distribution phase that one-compartment models miss. The two-compartment model with parameters ($A, B, \alpha, \beta$) or ($k_{10}, k_{12}, k_{21}, V_1$) is the most common structural model in PK. Always check semi-log plots for biphasic behavior before choosing your model.
:::

