---
title: "Absorption, Bioavailability, and Routes of Administration"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Compare PK profiles across different routes of administration
- Understand and compute absolute and relative bioavailability
- Model absorption with lag time, zero-order, and first-order processes
- Analyze modified-release formulations
- Select appropriate absorption models based on route and formulation

---

## Routes of Administration and PK Implications

### Comparison of Routes

| Route | Absorption Process | Key Feature |
|-------|-------------------|-------------|
| **Intravenous (IV)** | None (direct to blood) | 100% bioavailability, immediate |
| **Oral (PO)** | GI tract → portal vein → liver → systemic | First-pass metabolism, variable $F$ |
| **Subcutaneous (SC)** | Diffusion from tissue | Slower than IV, may have flip-flop |
| **Intramuscular (IM)** | Diffusion from muscle | Faster than SC, affected by blood flow |
| **Inhalation** | Lung epithelium | Rapid absorption, local + systemic |
| **Transdermal** | Skin diffusion | Very slow, zero-order kinetics |

---

## Bioavailability

### Absolute Bioavailability

$$F = \frac{AUC_{extravascular} \cdot Dose_{IV}}{AUC_{IV} \cdot Dose_{extravascular}}$$

$F$ represents the fraction of the administered dose that reaches systemic circulation.

### Relative Bioavailability

$$F_{rel} = \frac{AUC_{test} \cdot Dose_{reference}}{AUC_{reference} \cdot Dose_{test}}$$

Used to compare two formulations (e.g., tablet vs capsule).

```{r}
library(rxode2)
library(dplyr)
library(ggplot2)

# Simulate IV and oral profiles to compute bioavailability
mod_1cmt <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - ke * central
  conc <- central / V
})

times <- seq(0, 24, by = 0.1)

# IV dose
event_iv <- et(time = 0, amt = 50)  # central compartment
sim_iv <- rxSolve(rxode2({
  d/dt(central) <- -ke * central
  conc <- central / V
}), c(ke = 0.3, V = 10), event_iv %>% et(times))

# Oral dose
event_po <- et(time = 0, amt = 100)
sim_po <- rxSolve(mod_1cmt, c(ka = 1.0, ke = 0.3, V = 10), event_po %>% et(times))

# Compute AUC
auc_iv <- sum(diff(times) * (sim_iv$conc[-1] + sim_iv$conc[-length(sim_iv$conc)]) / 2)
auc_po <- sum(diff(times) * (sim_po$conc[-1] + sim_po$conc[-length(sim_po$conc)]) / 2)

F_abs <- (auc_po * 50) / (auc_iv * 100)
cat("Absolute Bioavailability (F):", round(F_abs * 100, 1), "%\n")
```

---

## First-Pass Metabolism

Drugs absorbed from the GI tract pass through the liver before reaching systemic circulation. The fraction escaping first-pass metabolism is:

$$F = f_{abs} \cdot (1 - E_H)$$

where $f_{abs}$ is the fraction absorbed from the gut and $E_H$ is the hepatic extraction ratio.

```{r}
# Bioavailability vs hepatic extraction ratio
E_H <- seq(0, 1, by = 0.01)
F_for_EH <- 1 - E_H

ggplot(data.frame(E_H, F_for_EH), aes(E_H, F_for_EH)) +
  geom_line(color = "#C44536", linewidth = 1.2) +
  labs(title = "Effect of Hepatic Extraction on Bioavailability",
       subtitle = "Assuming complete absorption (f_abs = 1)",
       x = "Hepatic Extraction Ratio (E_H)", y = "Bioavailability (F)") +
  theme_minimal()
```

---

## Absorption Lag Time

Some formulations have a delay before absorption begins (e.g., gastric emptying, tablet disintegration):

$$C(t) = \frac{F \cdot D \cdot k_a}{V \cdot (k_a - k_e)} \cdot (e^{-k_e(t - t_{lag})} - e^{-k_a(t - t_{lag})}), \quad t \geq t_{lag}$$

```{r}
# Model with absorption lag time
mod_lag <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - ke * central
  conc <- central / V

  # Parameter constraint
  alag(depot) <- tlag  # lag time for depot compartment
})

# Simulate with and without lag
event_po <- et(time = 0, amt = 100)
times <- seq(0, 24, by = 0.1)

sim_no_lag <- rxSolve(mod_1cmt, c(ka = 1.0, ke = 0.3, V = 10), event_po %>% et(times))
sim_lag    <- rxSolve(mod_lag, c(ka = 1.0, ke = 0.3, V = 10, tlag = 0.5), event_po %>% et(times))

df_lag <- bind_rows(
  as.data.frame(sim_no_lag) %>% mutate(Type = "No Lag"),
  as.data.frame(sim_lag) %>% mutate(Type = "t_lag = 0.5 h")
)

ggplot(df_lag, aes(time, conc, color = Type)) +
  geom_line(linewidth = 1) +
  labs(title = "Effect of Absorption Lag Time",
       x = "Time (hours)", y = "Concentration (mg/L)") +
  scale_color_manual(values = c("#0B6E6B", "#C44536")) +
  theme_minimal()
```

---

## Modified-Release Formulations

### Types of Modified Release

- **Extended release (ER/XR)**: Slower absorption → lower $C_{max}$, longer apparent $t_{1/2}$
- **Delayed release (DR)**: Release begins after a delay (e.g., enteric coating)
- **Pulsatile release**: Drug released in pulses at specific times

```{r}
# Model zero-order absorption (constant rate over duration D)
mod_zero <- rxode2({
  d/dt(depot)   <- -ka * depot  # but with zero-order into depot
  d/dt(central) <-  ka * depot - ke * central
  conc <- central / V
})

# Simulate extended release using a slow ka
event_er <- et(time = 0, amt = 100)
times <- seq(0, 48, by = 0.1)

sim_ir <- rxSolve(mod_1cmt, c(ka = 2.0, ke = 0.3, V = 10), event_er %>% et(times))
sim_er <- rxSolve(mod_1cmt, c(ka = 0.2, ke = 0.3, V = 10), event_er %>% et(times))

df_mr <- bind_rows(
  as.data.frame(sim_ir) %>% mutate(Formulation = "Immediate Release (ka=2.0)"),
  as.data.frame(sim_er) %>% mutate(Formulation = "Extended Release (ka=0.2)")
)

ggplot(df_mr, aes(time, conc, color = Formulation)) +
  geom_line(linewidth = 1) +
  labs(title = "Immediate vs Extended Release",
       x = "Time (hours)", y = "Concentration (mg/L)") +
  scale_color_manual(values = c("#C44536", "#0B6E6B")) +
  theme_minimal()
```

---

## Subcutaneous and Intramuscular Absorption

For SC/IM routes, absorption often follows first-order kinetics, but the rate can be slower than elimination (flip-flop):

```{r}
# Compare SC absorption rates
ka_values <- c(0.1, 0.3, 0.8, 2.0)
times <- seq(0, 48, by = 0.1)

sims_sc <- lapply(ka_values, function(ka_val) {
  sim <- rxSolve(mod_1cmt, c(ka = ka_val, ke = 0.3, V = 10), et(time = 0, amt = 100) %>% et(times))
  as.data.frame(sim) %>% mutate(ka = factor(paste0("ka = ", ka_val)))
})

df_sc <- bind_rows(sims_sc)

ggplot(df_sc, aes(time, conc, color = ka)) +
  geom_line(linewidth = 1) +
  labs(title = "Effect of Absorption Rate on SC Profile",
       subtitle = "ke = 0.3 (fixed); note flip-flop when ka < ke",
       x = "Time (hours)", y = "Concentration (mg/L)",
       color = "Absorption Rate") +
  theme_minimal()
```

---

## How Route Affects Modeling Choices

| Route | Recommended Absorption Model | Key Considerations |
|-------|------------------------------|-------------------|
| IV bolus | None | Instantaneous distribution |
| IV infusion | Zero-order input | Duration, rate |
| Oral (IR) | First-order ± lag | $T_{max}$, flip-flop risk |
| Oral (MR) | Sequential zero/first-order | Deconvolution may be needed |
| SC/IM | First-order | Often flip-flop |
| Transdermal | Zero-order (patch) | Constant rate until removal |

::: {.exercise-box}
### Exercises

1. Compute the absolute bioavailability of an oral tablet ($AUC$ = 85 mg·h/L, dose = 200 mg) given IV $AUC$ = 150 mg·h/L for a 50 mg IV dose.
2. Simulate an oral profile with $t_{lag}$ = 1 h. How does $T_{max}$ change compared to no lag?
3. An extended-release formulation has $k_a$ = 0.15 1/h while the IR has $k_a$ = 1.5 1/h. Compare $C_{max}$, $T_{max}$, and fluctuation.
4. Why might a SC injection show flip-flop kinetics? How would you confirm this?
:::

::: {.mistakes-box}
### Common Mistakes

- **Assuming 100% bioavailability for oral drugs** — always verify with IV data
- **Ignoring first-pass metabolism** — especially for high-extraction drugs
- **Not accounting for lag time** — can bias $k_a$ and $T_{max}$ estimates
- **Using the same absorption model for all routes** — SC/IM often need different treatment
- **Confusing absolute and relative bioavailability**
:::

::: {.key-box}
### Key Takeaway

The route of administration determines the absorption model you need. Master the concepts of bioavailability ($F$), first-pass metabolism, lag time, and how absorption rate interacts with elimination rate. Always check for flip-flop kinetics with extravascular data.
:::

