---
title: "Pharmacodynamics and Dose-Response"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Distinguish direct and indirect pharmacodynamic effects
- Differentiate graded and quantal responses
- Fit Emax, sigmoid Emax, linear, and log-linear PD models in R
- Interpret $E_{max}$, $EC_{50}$, and Hill coefficient
- Understand baseline effect, maximal response, and variability

---

## What Is Pharmacodynamics?

Pharmacodynamics describes the relationship between drug concentration (or dose) and the magnitude of the pharmacological effect. The fundamental question is:

> *Given a certain drug concentration at the site of action, how large is the effect?*

---

## The Emax Model

### Model Equation

$$E = E_0 + \frac{E_{max} \cdot C}{EC_{50} + C}$$

where:
- $E$: Observed effect
- $E_0$: Baseline effect (at $C = 0$)
- $E_{max}$: Maximum drug-attributable effect
- $EC_{50}$: Concentration producing 50% of $E_{max}$
- $C$: Drug concentration

```{r}
library(ggplot2)
library(dplyr)

# Visualize the Emax model
conc <- 10^seq(-2, 3, length.out = 200)
Emax <- -80   # e.g., 80% reduction (negative Emax for inhibition)
E0   <- 100   # baseline = 100%
EC50 <- 10

effect <- E0 + (Emax * conc) / (EC50 + conc)

df_emax <- data.frame(conc, effect)

ggplot(df_emax, aes(conc, effect)) +
  geom_line(color = "#C44536", linewidth = 1.2) +
  geom_vline(xintercept = EC50, linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = E0 + Emax, linetype = "dashed", color = "gray50") +
  annotate("text", x = EC50 * 2, y = E0 + 0.5 * Emax,
           label = paste0("EC50 = ", EC50), hjust = 0) +
  scale_x_log10() +
  labs(title = "Emax Model",
       subtitle = paste0("E0 = ", E0, ", Emax = ", Emax, ", EC50 = ", EC50),
       x = "Concentration (log scale)", y = "Effect") +
  theme_minimal()
```

---

## The Sigmoid Emax Model

### Model Equation

$$E = E_0 + \frac{E_{max} \cdot C^{\gamma}}{EC_{50}^{\gamma} + C^{\gamma}}$$

The Hill coefficient ($\gamma$) controls the steepness of the curve:
- $\gamma = 1$: Standard Emax
- $\gamma > 1$: Steeper (more switch-like)
- $\gamma < 1$: Shallower (more gradual)

```{r}
# Compare different Hill coefficients
conc <- 10^seq(-2, 3, length.out = 200)
E0 <- 0
Emax <- 100
EC50 <- 10
gamma_values <- c(0.5, 1, 2, 5)

df_sigmoid <- expand.grid(conc = conc, gamma = gamma_values) %>%
  mutate(
    effect = E0 + (Emax * conc^gamma) / (EC50^gamma + conc^gamma),
    gamma = factor(paste0("γ = ", gamma))
  )

ggplot(df_sigmoid, aes(conc, effect, color = gamma)) +
  geom_line(linewidth = 1) +
  scale_x_log10() +
  labs(title = "Sigmoid Emax Model: Effect of Hill Coefficient",
       x = "Concentration (log scale)", y = "Effect (% of maximum)",
       color = "Hill Coefficient") +
  scale_color_manual(values = c("#D4A843", "#C44536", "#0B6E6B", "#1B365D")) +
  theme_minimal()
```

### Interpreting the Hill Coefficient

| $\gamma$ | Curve Shape | Biological Interpretation |
|----------|-------------|--------------------------|
| < 1 | Shallow, gradual | Multiple binding sites with negative cooperativity |
| = 1 | Standard hyperbola | Simple 1:1 binding |
| > 1 | Steep, switch-like | Positive cooperativity, signal amplification |
| >> 1 | Nearly step function | Ultrasensitive systems |

---

## Linear and Log-Linear Models

### Linear Model

$$E = E_0 + S \cdot C$$

Used when the therapeutic range is well below $EC_{50}$.

### Log-Linear Model

$$E = m \cdot \ln(C) + b$$

Common in the 20–80% effect range on a log scale.

```{r}
# Compare Emax, linear, and log-linear in therapeutic range
conc_range <- seq(1, 50, length.out = 100)
EC50 <- 30
Emax <- 80

# Emax
eff_emax <- (Emax * conc_range) / (EC50 + conc_range)
# Linear approximation (slope at C=0 is Emax/EC50)
eff_lin <- (Emax / EC50) * conc_range
# Log-linear (fitted in mid-range)
eff_loglin <- 20 + 30 * log10(conc_range)

df_models <- data.frame(
  conc = rep(conc_range, 3),
  effect = c(eff_emax, eff_lin, eff_loglin),
  Model = rep(c("Emax", "Linear", "Log-Linear"), each = length(conc_range))
)

ggplot(df_models, aes(conc, effect, color = Model)) +
  geom_line(linewidth = 1) +
  labs(title = "PD Model Comparison",
       x = "Concentration", y = "Effect") +
  scale_color_manual(values = c("#C44536", "#0B6E6B", "#D4A843")) +
  theme_minimal()
```

---

## Fitting PD Models in R

```{r}
# Simulate dose-response data
set.seed(123)
doses <- c(0.1, 0.3, 1, 3, 10, 30, 100, 300)
n_rep <- 4

pd_data <- expand.grid(DOSE = doses, REP = 1:n_rep) %>%
  mutate(
    Emax = 90, EC50 = 10, E0 = 5,
    Effect_true = E0 + (Emax * DOSE) / (EC50 + DOSE),
    Effect_obs  = Effect_true + rnorm(n(), 0, 3)
  )

# Fit Emax model
fit_emax <- nls(Effect_obs ~ E0 + (Emax * DOSE) / (EC50 + DOSE),
                data = pd_data,
                start = list(E0 = 0, Emax = 80, EC50 = 5))
summary(fit_emax)
```

```{r}
# Fit sigmoid Emax model
fit_sig <- nls(Effect_obs ~ E0 + (Emax * DOSE^gamma) / (EC50^gamma + DOSE^gamma),
               data = pd_data,
               start = list(E0 = 0, Emax = 80, EC50 = 10, gamma = 1))
summary(fit_sig)
```

```{r}
# Plot fits
dose_pred <- 10^seq(-2, 3, length.out = 200)
pred_emax <- predict(fit_emax, newdata = data.frame(DOSE = dose_pred))
pred_sig  <- predict(fit_sig, newdata = data.frame(DOSE = dose_pred))

df_pred <- bind_rows(
  data.frame(DOSE = dose_pred, Effect = pred_emax, Model = "Emax"),
  data.frame(DOSE = dose_pred, Effect = pred_sig,  Model = "Sigmoid Emax")
)

ggplot(pd_data, aes(DOSE, Effect_obs)) +
  geom_point(size = 2, alpha = 0.6, color = "gray40") +
  geom_line(data = df_pred, aes(DOSE, Effect, color = Model), linewidth = 1.2) +
  scale_x_log10() +
  labs(title = "Fitted PD Models",
       x = "Dose (log scale)", y = "Effect") +
  scale_color_manual(values = c("#0B6E6B", "#C44536")) +
  theme_minimal()
```

---

## Graded vs Quantal Responses

| Response Type | Description | Example | Analysis |
|---------------|-------------|---------|----------|
| **Graded** | Continuous effect magnitude | Blood pressure reduction (mmHg) | Emax, linear models |
| **Quantal** | Binary (yes/no) outcome | Death/survival, response/no response | Logistic regression, $ED_{50}$ |

```{r}
# Quantal dose-response: logistic regression
set.seed(456)
doses_q <- rep(c(1, 3, 10, 30, 100), each = 20)
# Probability of response = Emax model with variability
logit_p <- -2 + 1.5 * log10(doses_q)
p <- plogis(logit_p)
response <- rbinom(length(doses_q), 1, p)

quantal_data <- data.frame(DOSE = doses_q, Response = response) %>%
  group_by(DOSE) %>%
  summarise(N = n(), Responders = sum(Response), .groups = "drop") %>%
  mutate(Fraction = Responders / N)

# Fit logistic regression
fit_logit <- glm(cbind(Responders, N - Responders) ~ log10(DOSE),
                 family = binomial, data = quantal_data)

# Predict for smooth curve
dose_pred <- 10^seq(0, 2.5, length.out = 100)
pred_p <- predict(fit_logit, newdata = data.frame(DOSE = dose_pred),
                  type = "response")

ggplot(quantal_data, aes(DOSE, Fraction)) +
  geom_point(size = 3, color = "#C44536") +
  geom_line(data = data.frame(DOSE = dose_pred, Fraction = pred_p),
            color = "#0B6E6B", linewidth = 1.2) +
  scale_x_log10() +
  labs(title = "Quantal Dose-Response",
       x = "Dose (log scale)", y = "Fraction Responding") +
  theme_minimal()
```

---

## Receptor Theory: The Mechanistic Basis of PD Curves

The empirical Emax and sigmoid Emax models describe concentration-effect data well, but they don't explain *why* the curves have that shape. Receptor theory bridges this gap by deriving PD curves from the fundamental chemistry of drug-receptor binding.

### Mass-Action Binding

The simplest model: a drug molecule ($L$) binds reversibly to a receptor ($R$):

$$R + L \underset{k_{-1}}{\stackrel{k_1}{\rightleftharpoons}} RL$$

At equilibrium, the dissociation constant $K_d$ is:

$$K_d = \frac{k_{-1}}{k_1} = \frac{[R][L]}{[RL]}$$

The fraction of receptors occupied (**occupancy**, $\theta$) follows the Langmuir isotherm:

$$\theta = \frac{[RL]}{[R]_{total}} = \frac{[L]}{K_d + [L]}$$

This is exactly the **same functional form as the Emax model** — the Emax curve arises naturally from simple 1:1 binding. $EC_{50}$ corresponds to $K_d$, and $E_{max}$ corresponds to the maximal response when all receptors are occupied.

```{r}
# Demonstrate receptor occupancy as a function of concentration
L <- 10^seq(-2, 3, length.out = 200)
Kd <- 3.5

# Occupancy (Langmuir isotherm) vs Emax model
occupancy <- L / (Kd + L)
emax_equiv <- L / (Kd + L) * 100  # scaled to 100%

df_receptor <- data.frame(
  Concentration = rep(L, 2),
  Value = c(occupancy, emax_equiv / 100),
  Curve = rep(c("Occupancy (θ)", "Emax (scaled)"), each = length(L))
)

ggplot(df_receptor, aes(Concentration, Value, color = Curve)) +
  geom_line(linewidth = 1.2) +
  scale_x_log10() +
  geom_vline(xintercept = Kd, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = Kd * 1.5, y = 0.5, label = paste0("Kd = ", Kd), hjust = 0) +
  labs(title = "Receptor Occupancy = Emax Model",
       subtitle = "The Emax curve is the Langmuir binding isotherm",
       x = "[L] (log scale)", y = "Fraction", color = "") +
  scale_color_manual(values = c("#0B6E6B", "#C44536")) +
  theme_minimal()
```

### From Occupancy to Effect: The Operational Model of Agonism

Not all occupied receptors produce the same effect. The **operational model of agonism** (Black & Leff, 1983) accounts for the transduction efficiency between binding and response:

$$E = \frac{E_{max} \cdot \theta^n}{\theta^n + K_A^n}$$

Substituting $\theta = [L]/(K_d + [L])$ gives a function that reduces to the sigmoid Emax model under most conditions. This explains why the Hill coefficient ($\gamma$) can differ from 1 even with simple 1:1 binding — it reflects **signal amplification** downstream of the receptor.

::: {.theory-box}
### The Chain: Binding → Occupancy → Effect

$$\text{Mass action: } R+L \rightleftharpoons RL \quad\rightarrow\quad \theta = \frac{[L]}{K_d+[L]}$$

$$\text{Operational model: } E = E_{max} \cdot \frac{\theta^n}{\theta^n + K_A^n}$$

$$\text{Empirical (Hill): } E = E_{max} \cdot \frac{[L]^\gamma}{EC_{50}^\gamma + [L]^\gamma}$$

These three equations form a nested hierarchy — each is a special case or generalization of the one above.
:::

---

## Agonism, Antagonism, and Allosteric Modulation

### Agonists

An **agonist** binds to a receptor and produces a response. The potency is reflected in $EC_{50}$ (or $K_d$), and the efficacy in $E_{max}$.

### Competitive Antagonism

A **competitive antagonist** binds to the same site as the agonist but produces no response. It shifts the dose-response curve to the right without reducing $E_{max}$:

$$E = E_{max} \cdot \frac{[A]}{[A] + EC_{50} \cdot (1 + [B]/K_B)}$$

where $[A]$ is agonist concentration, $[B]$ is antagonist concentration, and $K_B$ is the antagonist's dissociation constant.

```{r}
# Competitive antagonism: rightward shift without Emax reduction
A <- 10^seq(-2, 3, length.out = 200)
EC50 <- 8; Emax <- 100
B_conc <- c(0, 2, 8, 25)  # antagonist concentrations
KB <- 4                      # antagonist Kd

df_antag <- expand.grid(A = A, B = B_conc) %>%
  mutate(
    EC50_app = EC50 * (1 + B / KB),
    Effect = Emax * A / (EC50_app + A),
    Antagonist = factor(paste0("[B] = ", B))
  )

ggplot(df_antag, aes(A, Effect, color = Antagonist)) +
  geom_line(linewidth = 1) +
  scale_x_log10() +
  labs(title = "Competitive Antagonism",
       subtitle = "Rightward shift in dose-response curve; Emax preserved",
       x = "[Agonist] (log scale)", y = "Effect (% of maximum)") +
  scale_color_manual(values = c("#0B6E6B", "#D4A843", "#C44536", "#1B365D")) +
  theme_minimal()
```

### Non-Competitive Antagonism

A **non-competitive antagonist** binds to an allosteric site and reduces the maximal response without shifting $EC_{50}$ (in the pure case), or both reduces $E_{max}$ and shifts $EC_{50}$:

$$E = \frac{E_{max}}{1 + [B]/K_B} \cdot \frac{[A]}{[A] + EC_{50}}$$

### Allosteric Modulation

**Allosteric modulators** bind at a site distinct from the orthosteric (agonist) site and change receptor affinity ($\alpha$) and/or efficacy ($\beta$):

- **Positive allosteric modulator (PAM)**: increases agonist affinity or efficacy
- **Negative allosteric modulator (NAM)**: decreases agonist affinity or efficacy

::: {.key-box}
### Clinical Significance

Understanding antagonism mechanisms is essential for:
- **Reversal agents**: Naloxone (competitive antagonist) reverses opioid overdose
- **DDI prediction**: Co-administered antagonists shift the effective dose
- **Therapeutic window**: Irreversible antagonists can cause prolonged effects
:::

---

## Variability in PD

Sources of variability in PD:

- **Inter-individual**: Different patients, different responses
- **Intra-individual**: Same patient, different occasions
- **Measurement error**: Assay variability
- **Tolerance/desensitization**: Response changes over time

```{r}
# Visualize inter-individual variability in PD
set.seed(789)
n_subjects <- 20
conc <- 10^seq(-1, 3, length.out = 100)

# Each subject has a different EC50
pd_pop <- expand.grid(conc = conc, ID = 1:n_subjects) %>%
  mutate(
    EC50_ind = rlnorm(n(), log(10), 0.5),  # log-normal variability in EC50
    Emax_ind = rnorm(n(), 80, 10),          # normal variability in Emax
    Effect   = (Emax_ind * conc) / (EC50_ind + conc)
  )

ggplot(pd_pop, aes(conc, Effect, group = ID)) +
  geom_line(alpha = 0.3, color = "#0B6E6B") +
  scale_x_log10() +
  labs(title = "Inter-Individual Variability in PD",
       subtitle = "Each line = one subject; EC50 varies log-normally",
       x = "Concentration (log scale)", y = "Effect") +
  theme_minimal()
```

---

::: {.exercise-box}
### Exercises

1. Fit an Emax model to simulated data. Report $E_{max}$, $EC_{50}$, and their standard errors.
2. How does the Hill coefficient change the shape of the sigmoid Emax curve? Plot for $\gamma$ = 0.5, 1, 2, 4.
3. Generate quantal dose-response data and estimate $ED_{50}$ using logistic regression.
4. A drug has $E_{max}$ = -60 (inhibition), $EC_{50}$ = 5 ng/mL, and $\gamma$ = 1. What concentration is needed for 80% of maximum effect?
5. Compare the AIC of Emax, sigmoid Emax, and linear models for a given dataset. Which is preferred?
:::

::: {.mistakes-box}
### Common Mistakes

- **Using linear models when the drug approaches $E_{max}$** — linear extrapolation overestimates effect
- **Ignoring baseline ($E_0$)**: Always measure and model the baseline
- **Confusing $EC_{50}$ with potency**: Lower $EC_{50}$ means more potent
- **Forcing $\gamma = 1$ without checking**: Test whether the sigmoid model improves fit
- **Extrapolating far beyond the observed concentration range**
:::

::: {.key-box}
### Key Takeaway

The Emax and sigmoid Emax models are the workhorses of PD analysis. Understand the meaning of $E_{max}$, $EC_{50}$, and $\gamma$. Always plot your data before fitting, consider baseline effects, and be aware that PD variability can be substantial between individuals.
:::

