---
title: "PK/PD Link Models"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Understand why concentration and effect are often misaligned in time
- Build and simulate direct link, effect compartment, and indirect response models
- Identify and interpret hysteresis loops
- Estimate PK/PD link parameters in R
- Choose the appropriate link model for your data

---

## The PK/PD Disconnect

In many cases, the time course of drug effect does not parallel the time course of plasma concentration. This happens because:

1. The site of action is not in plasma (effect compartment delay)
2. The physiological response is indirect (signal transduction takes time)
3. The drug affects the production or loss of a response mediator

This leads to **hysteresis** — different concentrations produce different effects depending on whether concentration is rising or falling.

```{r}
library(rxode2)
library(ggplot2)
library(dplyr)

# Simulate PK with a delayed PD effect to demonstrate hysteresis
mod_pk <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - ke * central
  conc <- central / V
})

times <- seq(0, 24, by = 0.1)
params <- c(ka = 1.5, ke = 0.3, V = 10)
event <- et(time = 0, amt = 100)

sim_pk <- rxSolve(mod_pk, params, event %>% et(times))
pk_df <- as.data.frame(sim_pk)

# Add a delayed effect (simple shift)
pk_df <- pk_df %>%
  mutate(Effect = 100 * conc / (5 + conc))  # no delay

# Simulate with effect compartment delay
mod_link <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - ke * central
  d/dt(effect)  <- ke0 * (central/V - effect)
  conc <- central / V
  Ce   <- effect
})

params_link <- c(ka = 1.5, ke = 0.3, V = 10, ke0 = 0.4)
sim_link <- rxSolve(mod_link, params_link, event %>% et(times))
link_df <- as.data.frame(sim_link) %>%
  mutate(Effect_delayed = 100 * Ce / (5 + Ce))

# Hysteresis plot
ggplot(link_df, aes(conc, Effect_delayed)) +
  geom_path(color = "#0B6E6B", linewidth = 1) +
  geom_point(data = link_df[seq(1, nrow(link_df), by = 5), ],
             aes(conc, Effect_delayed), size = 1, alpha = 0.5) +
  labs(title = "Hysteresis Loop: Concentration vs Effect",
       subtitle = "Counter-clockwise hysteresis indicates effect compartment delay",
       x = "Plasma Concentration (mg/L)", y = "Effect") +
  theme_minimal()
```

---

## Direct Link Models

The simplest PK/PD model assumes the effect is directly related to plasma concentration without delay:

$$E(t) = f(C_p(t))$$

This works when the site of action equilibrates rapidly with plasma.

```{r}
# Direct link: Emax model driven by PK
mod_direct <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - ke * central
  conc <- central / V
  effect <- E0 + (Emax * conc) / (EC50 + conc)
})

params_dir <- c(ka = 1.5, ke = 0.3, V = 10,
                E0 = 5, Emax = 90, EC50 = 3)

sim_dir <- rxSolve(mod_direct, params_dir, event %>% et(times))

dir_df <- as.data.frame(sim_dir)

ggplot(dir_df, aes(time)) +
  geom_line(aes(y = conc, color = "Concentration"), linewidth = 1) +
  geom_line(aes(y = effect, color = "Effect"), linewidth = 1) +
  scale_y_continuous(
    name = "Concentration (mg/L)",
    sec.axis = sec_axis(~., name = "Effect")
  ) +
  labs(title = "Direct Link PK/PD Model",
       x = "Time (hours)", color = "") +
  scale_color_manual(values = c("#C44536", "#0B6E6B")) +
  theme_minimal()
```

---

## Effect Compartment Model

When there is a delay between plasma concentration and effect, we add an **effect compartment**:

$$\frac{dC_e}{dt} = k_{e0} \cdot (C_p - C_e)$$

where $k_{e0}$ is the equilibration rate constant between plasma and effect site.

```{r}
# Effect compartment model
mod_effect <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - ke * central
  d/dt(effect_c) <- ke0 * (central/V - effect_c)  # effect compartment
  conc <- central / V
  Ce   <- effect_c
  effect <- E0 + (Emax * Ce) / (EC50 + Ce)
})

params_ec <- c(ka = 1.5, ke = 0.3, V = 10,
               ke0 = 0.5, E0 = 5, Emax = 90, EC50 = 3)

sim_ec <- rxSolve(mod_effect, params_ec, event %>% et(times))
ec_df <- as.data.frame(sim_ec)

# Compare plasma concentration and effect-site concentration
ggplot(ec_df, aes(time)) +
  geom_line(aes(y = conc, color = "Plasma (Cp)"), linewidth = 1) +
  geom_line(aes(y = Ce, color = "Effect Site (Ce)"), linewidth = 1) +
  geom_line(aes(y = effect / 10, color = "Effect (÷10)"), linewidth = 1, linetype = "dashed") +
  labs(title = "Effect Compartment Model",
       subtitle = paste0("ke0 = ", params_ec["ke0"], " 1/h"),
       x = "Time (hours)", y = "Concentration (mg/L)", color = "") +
  scale_color_manual(values = c("#C44536", "#0B6E6B", "#D4A843")) +
  theme_minimal()
```

### Effect of $k_{e0}$ on Hysteresis

- **Large $k_{e0}$** → rapid equilibration → minimal hysteresis (approaches direct link)
- **Small $k_{e0}$** → slow equilibration → pronounced hysteresis

```{r}
# Compare different ke0 values
ke0_values <- c(0.1, 0.5, 2.0, 10)
hysteresis_data <- lapply(ke0_values, function(k) {
  params_k <- c(ka = 1.5, ke = 0.3, V = 10, ke0 = k, E0 = 5, Emax = 90, EC50 = 3)
  sim <- rxSolve(mod_effect, params_k, event %>% et(times))
  as.data.frame(sim) %>% mutate(ke0 = factor(paste0("ke0 = ", k)))
})

df_hyst <- bind_rows(hysteresis_data)

ggplot(df_hyst, aes(conc, effect, color = ke0)) +
  geom_path(linewidth = 0.8) +
  labs(title = "Hysteresis at Different ke0 Values",
       x = "Plasma Concentration (mg/L)", y = "Effect") +
  scale_color_manual(values = c("#D4A843", "#C44536", "#0B6E6B", "#1B365D")) +
  facet_wrap(~ke0, scales = "free_x") +
  theme_minimal()
```

---

## Indirect Response Models

In indirect response models, the drug does not directly cause the effect — it modulates the production ($k_{in}$) or loss ($k_{out}$) of an endogenous response mediator:

### Basic Indirect Response Model

$$\frac{dR}{dt} = k_{in} - k_{out} \cdot R$$

At baseline: $R_0 = k_{in} / k_{out}$

### Four Basic Indirect Response Models

| Model | Drug Action | Equation |
|-------|-------------|----------|
| Type I | Inhibits $k_{in}$ | $\frac{dR}{dt} = k_{in} \cdot (1 - \frac{I_{max}C}{IC_{50}+C}) - k_{out}R$ |
| Type II | Inhibits $k_{out}$ | $\frac{dR}{dt} = k_{in} - k_{out} \cdot (1 - \frac{I_{max}C}{IC_{50}+C}) \cdot R$ |
| Type III | Stimulates $k_{in}$ | $\frac{dR}{dt} = k_{in} \cdot (1 + \frac{S_{max}C}{SC_{50}+C}) - k_{out}R$ |
| Type IV | Stimulates $k_{out}$ | $\frac{dR}{dt} = k_{in} - k_{out} \cdot (1 + \frac{S_{max}C}{SC_{50}+C}) \cdot R$ |

```{r}
# Indirect Response Type I (inhibition of kin)
mod_idr1 <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - ke * central
  C <- central / V

  # Drug effect: inhibition of response production
  inhibition <- Imax * C / (IC50 + C)

  d/dt(R) <- kin * (1 - inhibition) - kout * R
  response <- R
})

# Parameters
params_idr <- c(ka = 1.5, ke = 0.3, V = 10,
                kin = 10, kout = 0.5,  # R0 = kin/kout = 20
                Imax = 0.9, IC50 = 2)

sim_idr <- rxSolve(mod_idr1, params_idr, event %>% et(times))
idr_df <- as.data.frame(sim_idr)

# Plot PK and response
ggplot(idr_df, aes(time)) +
  geom_line(aes(y = C, color = "Concentration"), linewidth = 1) +
  geom_line(aes(y = response, color = "Response"), linewidth = 1) +
  labs(title = "Indirect Response Model (Type I)",
       subtitle = "Inhibition of response production (kin)",
       x = "Time (hours)", y = "Value", color = "") +
  scale_color_manual(values = c("#0B6E6B", "#C44536")) +
  theme_minimal()
```

### Characteristic Profiles

```{r}
# Compare all four indirect response types
simulate_idr <- function(type, params, event, times) {
  mod <- switch(type,
    "I" = rxode2({
      d/dt(depot) <- -ka*depot; d/dt(central) <- ka*depot - ke*central
      C <- central/V; d/dt(R) <- kin*(1 - Imax*C/(IC50+C)) - kout*R; response <- R
    }),
    "II" = rxode2({
      d/dt(depot) <- -ka*depot; d/dt(central) <- ka*depot - ke*central
      C <- central/V; d/dt(R) <- kin - kout*(1 - Imax*C/(IC50+C))*R; response <- R
    }),
    "III" = rxode2({
      d/dt(depot) <- -ka*depot; d/dt(central) <- ka*depot - ke*central
      C <- central/V; d/dt(R) <- kin*(1 + Smax*C/(SC50+C)) - kout*R; response <- R
    }),
    "IV" = rxode2({
      d/dt(depot) <- -ka*depot; d/dt(central) <- ka*depot - ke*central
      C <- central/V; d/dt(R) <- kin - kout*(1 + Smax*C/(SC50+C))*R; response <- R
    })
  )
  sim <- rxSolve(mod, params, event %>% et(times))
  as.data.frame(sim) %>% mutate(Type = paste("Type", type))
}

types <- c("I", "II", "III", "IV")
idr_all <- bind_rows(lapply(types, function(t) {
  p <- c(ka = 1.5, ke = 0.3, V = 10, kin = 10, kout = 0.5,
         Imax = 0.9, IC50 = 2, Smax = 4, SC50 = 2)
  simulate_idr(t, p, event, times)
}))

ggplot(idr_all, aes(time, response, color = Type)) +
  geom_line(linewidth = 1) +
  facet_wrap(~Type, scales = "free_y") +
  labs(title = "Four Types of Indirect Response Models",
       x = "Time (hours)", y = "Response") +
  scale_color_manual(values = c("#0B6E6B", "#C44536", "#D4A843", "#1B365D")) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Disease Progression Models

Disease progression models extend PK/PD by coupling a third state variable — the disease state $D(t)$ — to the drug concentration and effect. This captures the reality that diseases evolve over time, and drugs modify that trajectory.

### The Basic Disease Progression ODE

A simple model for disease natural history:

$$\frac{dD}{dt} = \alpha - \beta \cdot D$$

where $\alpha$ is the disease progression rate and $\beta$ is the natural recovery/remission rate. At steady state without treatment: $D_{ss} = \alpha/\beta$.

Drug effect is coupled in through the effect $E(t)$:

$$\frac{dD}{dt} = \alpha - \beta \cdot D - \gamma \cdot E(t)$$

where $\gamma$ quantifies how strongly the drug effect translates into disease modification.

### Symptomatic vs Disease-Modifying Effects

| Effect Type | Mathematical Signature | Clinical Implication |
|---|---|---|
| **Symptomatic** | Effect on $D$ reverses when drug stops | Disease returns to natural trajectory |
| **Disease-modifying** | Effect persists or slows $\alpha$ permanently | Long-term benefit beyond drug exposure |

```{r}
library(rxode2)
library(ggplot2)
library(dplyr)

# Disease progression coupled with PK/PD
mod_disease <- rxode2({
  # PK: one-compartment oral
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - ke * central
  C <- central / V
  
  # PD: Emax effect driven by concentration
  E <- Emax * C / (EC50 + C)
  
  # Disease state
  d/dt(D) <- alpha - beta * D - gamma_drug * E
  
  # Outputs
  conc    <- C
  effect  <- E
  disease <- D
})

# Parameters
params_disease <- c(
  ka = 0.9, ke = 0.22, V = 28,         # PK
  Emax = -4, EC50 = 3.5,               # PD (negative = improvement)
  alpha = 0.35, beta = 0.035,          # Disease: steady state D = 10
  gamma_drug = 0.8                     # Drug effect coupling
)

times <- seq(0, 72, by = 0.2)
event <- et(time = 0, amt = 200) %>%
  et(seq(24, 72, 24), amt = rep(200, 3))  # daily dosing

sim_disease <- rxSolve(mod_disease, params_disease, event %>% et(times))
df_disease <- as.data.frame(sim_disease)

# Plot all three coupled states
ggplot(df_disease, aes(time)) +
  geom_line(aes(y = disease, color = "Disease State (D)"), linewidth = 1) +
  geom_line(aes(y = conc * 5, color = "Conc (×5)"), linewidth = 0.8, alpha = 0.6) +
  labs(title = "Disease Progression with PK/PD Coupling",
       subtitle = "Drug slows disease progression; symptomatic relief reverses when drug stops",
       x = "Time (hours)", y = "Disease Severity", color = "") +
  scale_color_manual(values = c("#C44536", "#0B6E6B")) +
  theme_minimal()
```

### The General Pattern: Coupled Multi-State ODE Systems

Once you have concentration $C(t)$ and effect $E(t)$ as ODE states, you can couple in additional states:

- **Enzyme level** (induction): $dE_{enz}/dt = k_{syn} \cdot (1 + I_{max}C/(IC_{50}+C)) - k_{deg} \cdot E_{enz}$
- **Second drug** (DDI): $dC_B/dt = \dots$ with combined or competitive clearance
- **Biomarker**: $dM/dt = k_{in} \cdot f(C) - k_{out} \cdot M$

The pattern is always the same: each new state's derivative is a function of the existing states, and the whole system is solved with a single `ode()` call.

::: {.warning-box}
### The Identifiability Risk

As you couple more states, the parameter count multiplies rapidly.
A PK+PD+disease model may have 10+ parameters. When the number of
informative data points doesn't grow proportionally, parameters become
unidentifiable — the model can produce the same fit with very different
parameter sets. **Fix parameters obtainable from independent sources**
(e.g., $V$ from NCA, $\alpha$ from natural history studies) before
estimating the rest.
:::

---

## Choosing a PK/PD Link Model

::: {.theory-box}
### Decision Flow

1. Plot concentration vs effect over time — are they in phase?
   - **Yes** → Direct link model
   - **No** → Check hysteresis direction

2. Plot concentration vs effect (hysteresis plot)
   - **No loop** → Direct link
   - **Counter-clockwise loop** → Effect compartment model
   - **Clockwise loop** → Tolerance or indirect response

3. Consider mechanism
   - Receptor-mediated, rapid onset → Direct link or effect compartment
   - Physiological mediator turnover → Indirect response
   - Gene expression, protein synthesis → Signal transduction cascade
:::

::: {.exercise-box}
### Exercises

1. Simulate an effect compartment model with $k_{e0}$ = 0.2, 0.5, 1.0, and 3.0. How does the peak effect time change?
2. Create a hysteresis plot (concentration vs effect) for an effect compartment model and identify the direction of the loop.
3. Simulate an indirect response Type I model. What happens to the response when $k_{out}$ is doubled?
4. Explain why a clockwise hysteresis loop suggests tolerance rather than effect compartment delay.
:::

::: {.mistakes-box}
### Common Mistakes

- **Assuming direct link by default** — always check for hysteresis
- **Misinterpreting hysteresis direction**: Counter-clockwise = distribution delay; Clockwise = tolerance
- **Using wrong indirect response type**: The four types produce different response profiles
- **Ignoring baseline turnover**: $R_0 = k_{in}/k_{out}$ must match observed baseline
:::

::: {.key-box}
### Key Takeaway

PK/PD link models bridge the gap between drug concentration and effect. Master the three main types: direct link (no delay), effect compartment (distributional delay), and indirect response (pharmacological delay). Always visualize hysteresis before choosing your model.
:::

