---
title: "Nonlinear and Saturable PK"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Understand Michaelis–Menten (capacity-limited) elimination
- Simulate saturable protein binding and its effect on PK
- Model target-mediated drug disposition (TMDD)
- Identify when linear PK assumptions fail
- Compare linear and nonlinear model predictions in R

---

## When Linearity Breaks Down

Most PK models assume **first-order (linear) elimination**: rate $\propto$ concentration. But many drugs show **saturable (nonlinear) PK** at therapeutic doses because:

1. **Metabolizing enzymes become saturated** (Michaelis–Menten kinetics)
2. **Protein binding is saturable** (only free drug is cleared)
3. **Target binding is saturable** (target-mediated drug disposition)
4. **Transporters become saturated** (renal secretion, biliary excretion)

::: {.key-box}
### Linear vs Nonlinear PK

| Property | Linear PK | Nonlinear PK |
|----------|-----------|--------------|
| AUC vs Dose | Proportional | More than proportional |
| $t_{1/2}$ | Constant | Dose-dependent |
| $CL$ | Constant | Dose-dependent |
| Superposition | Applies | Does NOT apply |
:::

---

## Michaelis–Menten Elimination

### Model Equation

$$\frac{dC}{dt} = -\frac{V_{max} \cdot C}{K_m + C}$$

For IV bolus: $$\frac{dC}{dt} = -\frac{V_{max} \cdot C}{K_m + C}, \quad C(0) = \frac{D}{V}$$

where:
- $V_{max}$: Maximum elimination rate (amount/time)
- $K_m$: Michaelis constant (concentration at half $V_{max}$)

```{r}
library(rxode2)
library(ggplot2)
library(dplyr)

# Michaelis-Menten elimination model
mod_mm <- rxode2({
  d/dt(central) <- -Vmax * central / (Km * V + central)
  conc <- central / V
})

# Simulate at different doses
doses <- c(50, 200, 800, 3200)
times <- seq(0, 48, by = 0.1)
params_mm <- c(Vmax = 50, Km = 5, V = 10)  # Vmax mg/h, Km mg/L, V L

sims_mm <- lapply(doses, function(d) {
  event <- et(time = 0, amt = d)
  sim <- rxSolve(mod_mm, params_mm, event %>% et(times))
  as.data.frame(sim) %>% mutate(Dose = factor(paste(d, "mg")))
})

df_mm <- bind_rows(sims_mm)

ggplot(df_mm, aes(time, conc, color = Dose)) +
  geom_line(linewidth = 1) +
  labs(title = "Michaelis–Menten Elimination: Effect of Dose",
       subtitle = "Higher doses show slower elimination (saturation)",
       x = "Time (hours)", y = "Concentration (mg/L)") +
  scale_color_manual(values = c("#0B6E6B", "#2E8B7E", "#D4A843", "#C44536")) +
  theme_minimal()
```

### When Does Nonlinearity Matter?

- When concentrations exceed $K_m$ (the Michaelis constant)
- At low concentrations ($C \ll K_m$): Elimination approaches first-order with $k \approx V_{max}/K_m$
- At high concentrations ($C \gg K_m$): Elimination approaches zero-order at rate $V_{max}$

```{r}
# Demonstrate the transition from first-order to zero-order
C <- seq(0.1, 100, length.out = 200)
Vmax <- 50
Km <- 5

elim_rate_mm <- Vmax * C / (Km + C)          # MM rate (amount/h)
elim_rate_1st <- (Vmax / Km) * C              # First-order approximation (low C)
elim_rate_0th <- rep(Vmax, length(C))          # Zero-order approximation (high C)

df_rates <- data.frame(
  C = rep(C, 3),
  Rate = c(elim_rate_mm, elim_rate_1st, elim_rate_0th),
  Model = rep(c("Michaelis-Menten", "First-order (Vmax/Km)", "Zero-order (Vmax)"),
              each = length(C))
)

ggplot(df_rates, aes(C, Rate, color = Model)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = Km, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = Km * 1.5, y = Vmax * 0.3, label = "Km", hjust = 0) +
  labs(title = "Michaelis–Menten: Transition from First- to Zero-Order",
       x = "Concentration (mg/L)", y = "Elimination Rate (mg/h)") +
  scale_color_manual(values = c("#0B6E6B", "#D4A843", "#C44536")) +
  theme_minimal()
```

---

## Saturable Protein Binding

Many drugs bind to plasma proteins (albumin, α1-acid glycoprotein). Only **unbound (free) drug** is pharmacologically active and available for elimination.

### Model for Saturable Binding

$$C_{bound} = \frac{B_{max} \cdot C_{free}}{K_D + C_{free}}$$

where $B_{max}$ is the maximum binding capacity and $K_D$ is the dissociation constant.

```{r}
# Saturable protein binding model
C_free <- seq(0.01, 200, length.out = 300)
Bmax <- 100  # maximum binding capacity (mg/L)
KD <- 5      # dissociation constant (mg/L)

C_bound <- Bmax * C_free / (KD + C_free)
C_total <- C_free + C_bound
fu <- C_free / C_total  # fraction unbound

df_binding <- data.frame(
  C_free = rep(C_free, 3),
  Value = c(C_free, C_bound, C_total),
  Component = rep(c("Free", "Bound", "Total"), each = length(C_free))
)

ggplot(df_binding, aes(C_free, Value, color = Component)) +
  geom_line(linewidth = 1) +
  labs(title = "Saturable Protein Binding",
       subtitle = paste0("Bmax = ", Bmax, " mg/L, KD = ", KD, " mg/L"),
       x = "Free Concentration (mg/L)", y = "Concentration (mg/L)") +
  scale_color_manual(values = c("#0B6E6B", "#C44536", "#1B365D")) +
  theme_minimal()

# Fraction unbound vs concentration
df_fu <- data.frame(C_free = C_free, fu = fu)

ggplot(df_fu, aes(C_free, fu)) +
  geom_line(color = "#C44536", linewidth = 1.2) +
  labs(title = "Fraction Unbound vs Concentration",
       subtitle = "fu increases as binding sites become saturated",
       x = "Total Concentration (mg/L)", y = "Fraction Unbound (fu)") +
  theme_minimal()
```

---

## Target-Mediated Drug Disposition (TMDD)

TMDD occurs when a significant fraction of the drug binds to its pharmacological target (e.g., receptor, enzyme). This binding contributes to both distribution and elimination.

### Minimal TMDD Model

```{r}
# Simplified TMDD model
mod_tmdd <- rxode2({
  # Free drug in central compartment
  d/dt(C) <- -ke * C - kon * C * R + koff * RC

  # Free receptor
  d/dt(R) <- ksyn - kdeg * R - kon * C * R + koff * RC

  # Drug-receptor complex
  d/dt(RC) <- kon * C * R - koff * RC - kint * RC

  conc_free <- C
  conc_total <- C + RC
})

params_tmdd <- c(
  ke = 0.1,    # elimination rate of free drug
  kon = 0.5,   # binding rate
  koff = 0.1,  # dissociation rate
  ksyn = 1.0,  # receptor synthesis rate
  kdeg = 0.2,  # receptor degradation rate
  kint = 0.3   # internalization rate of complex
)

V <- 5

# Initial conditions
inits_tmdd <- c(C = 100 / V, R = params_tmdd[["ksyn"]] / params_tmdd[["kdeg"]], RC = 0)

times <- seq(0, 72, by = 0.1)
event <- et(time = 0, amt = 100)

sim_tmdd <- rxSolve(mod_tmdd, c(params_tmdd, V = V), event %>% et(times), inits = inits_tmdd)
tmdd_df <- as.data.frame(sim_tmdd)

# Plot
ggplot(tmdd_df, aes(time)) +
  geom_line(aes(y = conc_free, color = "Free Drug"), linewidth = 1) +
  geom_line(aes(y = conc_total, color = "Total Drug"), linewidth = 1) +
  geom_line(aes(y = R * 10, color = "Receptor (×10)"), linewidth = 1, linetype = "dashed") +
  labs(title = "Target-Mediated Drug Disposition",
       subtitle = "Drug binding to receptor affects PK profile",
       x = "Time (hours)", y = "Concentration (mg/L)", color = "") +
  scale_color_manual(values = c("#0B6E6B", "#C44536", "#D4A843")) +
  theme_minimal()
```

---

## Detecting Nonlinear PK

### Methods to Identify Nonlinearity

1. **Dose proportionality check**: Plot AUC vs dose — should be linear
2. **Dose-normalized profiles**: Overlay $C/D$ vs time — should superimpose for linear PK
3. **Multiple-dose comparison**: Check if superposition principle holds

```{r}
# Check dose proportionality
calc_auc <- function(sim) {
  sum(diff(sim$time) * (sim$conc[-1] + sim$conc[-nrow(sim)]) / 2)
}

doses_check <- c(10, 30, 100, 300, 1000)
auc_values <- sapply(doses_check, function(d) {
  event <- et(time = 0, amt = d)
  sim <- rxSolve(mod_mm, params_mm, event %>% et(times))
  calc_auc(as.data.frame(sim))
})

df_auc <- data.frame(Dose = doses_check, AUC = auc_values) %>%
  mutate(Dose_Normalized_AUC = AUC / Dose)

knitr::kable(df_auc, caption = "Dose Proportionality Check",
             col.names = c("Dose (mg)", "AUC (mg·h/L)", "Dose-Normalized AUC"))

ggplot(df_auc, aes(Dose, Dose_Normalized_AUC)) +
  geom_point(size = 3, color = "#C44536") +
  geom_line(color = "#C44536") +
  labs(title = "Dose-Normalized AUC vs Dose",
       subtitle = "Non-horizontal line indicates nonlinear PK",
       x = "Dose (mg)", y = "AUC / Dose") +
  theme_minimal()
```

---

::: {.exercise-box}
### Exercises

1. Simulate Michaelis–Menten elimination at doses of 10, 50, 200, and 1000 mg. Compare $t_{1/2}$ (time to halve concentration) across doses.
2. At what concentration does the MM elimination rate reach 90% of $V_{max}$?
3. Simulate saturable protein binding and plot total vs free concentration. What happens to the fraction unbound at high concentrations?
4. For TMDD, explain why the PK profile appears nonlinear at low doses but approximately linear at very high doses.
:::

::: {.mistakes-box}
### Common Mistakes

- **Assuming linear PK without checking dose proportionality**
- **Using first-order models for drugs with known nonlinearity** (e.g., phenytoin, high-dose methotrexate)
- **Ignoring $K_m$ relative to therapeutic concentrations**: If $C_{therapeutic} \approx K_m$, nonlinearity is significant
- **Extrapolating low-dose linear PK to high doses** without checking for saturation
- **Confusing saturable protein binding with saturable elimination** — they have different PK signatures
:::

::: {.key-box}
### Key Takeaway

Nonlinear PK arises when elimination, binding, or distribution processes become saturated. Michaelis–Menten kinetics is the most common form. Always check for dose proportionality before assuming linear PK — especially for drugs with narrow therapeutic windows or high-dose regimens.
:::

