---
title: "Variability and Population PK"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Understand sources of variability in PK parameters
- Distinguish fixed effects from random effects
- Build a population PK model with covariates
- Use `nlmixr2` for nonlinear mixed-effects modeling
- Interpret population parameters, individual predictions, and shrinkage

---

## Why Population PK?

In classical PK, we analyze one subject at a time (individual PK). But in drug development, we need to:

1. Analyze **sparse data** (1–3 samples per subject from hundreds of patients)
2. Quantify **between-subject variability** (why does CL vary 5-fold?)
3. Identify **covariates** (weight, age, renal function) that explain variability
4. Support **dose individualization**

**Population PK** (PopPK) uses nonlinear mixed-effects models to analyze all subjects simultaneously.

---

## The Population Model Structure

### Hierarchical Model

**Level 1 — Structural (PK) model:**
$$C_{ij} = f(t_{ij}, \theta_i) + \varepsilon_{ij}$$

**Level 2 — Between-subject variability (BSV):**
$$\theta_i = \theta_{pop} \cdot e^{\eta_i}, \quad \eta_i \sim N(0, \omega^2)$$

**Level 3 — Covariate effects:**
$$\theta_i = \theta_{pop} \cdot \left(\frac{WT_i}{70}\right)^{\beta_{WT}} \cdot e^{\eta_i}$$

**Level 4 — Residual error:**
$$\varepsilon_{ij} \sim N(0, \sigma^2) \quad \text{or} \quad C_{ij} = f(t_{ij}, \theta_i) \cdot (1 + \varepsilon_{ij})$$

```{r}
library(ggplot2)
library(dplyr)

# Conceptual visualization: Between-subject variability
set.seed(123)
n_subj <- 50
times <- seq(0, 24, by = 0.5)

# Generate individual PK profiles with variability in CL and V
pop_CL <- 5     # L/h
pop_V  <- 50    # L

# Log-normal BSV: ω_CL = 0.3, ω_V = 0.2
set.seed(42)

sim_pop <- expand.grid(time = times, ID = 1:n_subj) %>%
  mutate(
    eta_CL = rep(rnorm(n_subj, 0, 0.3), each = length(times)),
    eta_V  = rep(rnorm(n_subj, 0, 0.2), each = length(times)),
    CL_ind = pop_CL * exp(eta_CL),
    V_ind  = pop_V * exp(eta_V),
    ke_ind = CL_ind / V_ind,
    dose   = 500,
    conc   = (dose / V_ind) * exp(-ke_ind * time),
    conc_obs = conc * exp(rnorm(n(), 0, 0.1))
  )

# Plot individual profiles
ggplot(sim_pop, aes(time, conc, group = ID)) +
  geom_line(alpha = 0.3, color = "#0B6E6B") +
  labs(title = "Population PK: 50 Simulated Subjects",
       subtitle = "Between-subject variability in CL (ω=0.3) and V (ω=0.2)",
       x = "Time (hours)", y = "Concentration (mg/L)") +
  theme_minimal()
```

---

## Population PK with nlmixr2

```{r}
#| eval: false
library(nlmixr2)

# Define a one-compartment model for nlmixr2
pop_model <- function() {
  ini({
    # Population parameters (fixed effects)
    tka <- log(1.0)     # log ka
    tcl <- log(5.0)     # log CL
    tv  <- log(50)      # log V

    # Between-subject variability (random effects)
    eta.ka ~ 0.3
    eta.cl ~ 0.3
    eta.v  ~ 0.2

    # Residual error
    add.err <- 1
    prop.err <- 0.1
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv + eta.v)

    d/dt(depot)   <- -ka * depot
    d/dt(central) <-  ka * depot - (cl/v) * central
    cp <- central / v
    cp ~ add(add.err) + prop(prop.err)
  })
}
```

```{r}
# Simulate sparse population PK data
set.seed(123)
n_subjects <- 100
sparse_times <- c(0.5, 2, 6, 24)  # Typical sparse sampling

pk_sparse <- expand.grid(ID = 1:n_subjects, time = sparse_times) %>%
  arrange(ID, time) %>%
  mutate(
    eta_CL = rep(rnorm(n_subjects, 0, 0.3), each = length(sparse_times)),
    eta_V  = rep(rnorm(n_subjects, 0, 0.2), each = length(sparse_times)),
    CL = 5 * exp(eta_CL),
    V  = 50 * exp(eta_V),
    ke = CL / V,
    DOSE = 500,
    conc = (DOSE / V) * exp(-ke * time),
    conc_obs = conc * exp(rnorm(n(), 0, 0.15)),  # 15% CV residual error
    CENS = 0  # Not censored
  ) %>%
  filter(conc_obs > 0.01)  # Remove BLOQ

head(pk_sparse, 10)
```

### Fitting the Population Model

```{r}
#| eval: false
# Fit the model (requires nlmixr2)
fit_pop <- nlmixr2(
  pop_model,
  data = pk_sparse,
  est = "saem"  # SAEM estimation method
)

# View results
print(fit_pop)

# Parameter estimates
summary(fit_pop)
```

---

## Covariate Modeling

### Common Covariates in PK

| Covariate | Typical Effect On | Functional Form |
|-----------|-------------------|----------------|
| Body weight (WT) | CL, V | Power: $(WT/70)^{0.75}$ for CL |
| Age | CL (maturation, decline) | Power or exponential |
| Renal function (eGFR) | CL (renal drugs) | Linear or power |
| Sex | V, CL | Categorical |
| Genotype (CYP450) | CL (metabolized drugs) | Categorical |
| Disease status | CL, V | Categorical |

```{r}
# Covariate model simulation: Effect of weight on CL
weights <- seq(40, 120, by = 5)
ref_wt <- 70
CL_ref <- 5.0
theta_wt <- 0.75  # allometric exponent for CL

CL_by_wt <- CL_ref * (weights / ref_wt)^theta_wt

df_cov <- data.frame(Weight = weights, CL = CL_by_wt)

ggplot(df_cov, aes(Weight, CL)) +
  geom_line(color = "#0B6E6B", linewidth = 1.2) +
  geom_point(size = 2, color = "#0B6E6B") +
  geom_vline(xintercept = ref_wt, linetype = "dashed", alpha = 0.5) +
  labs(title = "Allometric Scaling of Clearance with Body Weight",
       subtitle = expression(CL == CL[ref] %.% (WT/70)^0.75),
       x = "Body Weight (kg)", y = "Clearance (L/h)") +
  theme_minimal()
```

---

## Sparse vs Rich Sampling

| Design | Samples per Subject | Subjects | Analysis |
|--------|--------------------|----------|----------|
| **Rich** | 10–15 | 12–24 | Individual PK (standard two-stage) |
| **Sparse** | 1–6 | 100–1000 | Population PK (mixed effects) |
| **Hybrid** | Mix of rich and sparse | Variable | Population PK |

```{r}
# Visualize sparse vs rich sampling designs
rich_times <- c(0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 24)
sparse_times_viz <- c(0.5, 2, 6, 24)

sample_designs <- data.frame(
  time = c(rich_times, sparse_times_viz),
  conc = 10 * exp(-0.3 * c(rich_times, sparse_times_viz)),
  Design = rep(c("Rich (n=1–30)", "Sparse (n=100+)"),
               c(length(rich_times), length(sparse_times_viz)))
)

ggplot(sample_designs, aes(time, conc, color = Design, shape = Design)) +
  geom_point(size = 3) +
  scale_y_log10() +
  labs(title = "Sampling Design Comparison",
       x = "Time (hours)", y = "Concentration (log scale)") +
  scale_color_manual(values = c("#C44536", "#0B6E6B")) +
  theme_minimal()
```

---

## Model Diagnostics for Population PK

### Key Diagnostics

1. **Observed vs Predicted**: Individual predictions (IPRED) should follow identity
2. **Residuals vs Predicted**: No trends, symmetric around zero
3. **Distribution of random effects ($\eta$)**: Should be approximately normal
4. **Shrinkage**: High shrinkage ($\eta$-shrinkage > 20–30%) reduces reliability of individual estimates

```{r}
# Simulate diagnostic plots
set.seed(456)
n <- 200
pred <- runif(n, 0.1, 50)
obs <- pred * exp(rnorm(n, 0, 0.2))

diag_df <- data.frame(
  PRED = pred,
  IPRED = obs,
  CWRES = (log(obs) - log(pred)) / 0.2
)

# Observed vs Predicted
p1 <- ggplot(diag_df, aes(PRED, IPRED)) +
  geom_point(alpha = 0.5, color = "#0B6E6B") +
  geom_abline(slope = 1, intercept = 0, color = "#C44536", linetype = "dashed") +
  labs(title = "Observed vs Population Predicted", x = "PRED", y = "DV") +
  theme_minimal()

# Residuals vs Predicted
p2 <- ggplot(diag_df, aes(PRED, CWRES)) +
  geom_point(alpha = 0.5, color = "#C44536") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = c(-2, 2), linetype = "dotted", alpha = 0.5) +
  labs(title = "CWRES vs Predicted", x = "PRED", y = "CWRES") +
  theme_minimal()

library(patchwork)
p1 + p2
```

---

## Shrinkage

Shrinkage occurs when individual data are sparse and individual estimates "shrink" toward the population mean:

$$\eta\text{-shrinkage} = 1 - \frac{var(\hat{\eta}_i)}{\omega^2}$$

- **Low shrinkage (< 10%)**: Individual estimates are reliable
- **Moderate shrinkage (10–30%)**: Caution needed for covariate selection using EBEs
- **High shrinkage (> 30%)**: Individual estimates unreliable; use population predictions

---

::: {.exercise-box}
### Exercises

1. Simulate a population PK dataset with 200 subjects, 3 samples each. Fit using nlmixr2 and report $\omega_{CL}$ and $\omega_V$.
2. Add body weight as a covariate on CL. Does it reduce $\omega_{CL}$? By how much?
3. Generate diagnostic plots for a fitted model. What patterns would indicate model misspecification?
4. Explain why shrinkage matters for covariate model building.
5. Compare rich vs sparse sampling designs: which gives better estimates of $\omega^2$?
:::

::: {.mistakes-box}
### Common Mistakes

- **Using individual PK methods on sparse data**: Leads to biased, imprecise estimates
- **Ignoring shrinkage**: EBEs with high shrinkage should not be used for covariate screening
- **Overfitting covariates**: Each covariate consumes degrees of freedom; use scientific rationale
- **Assuming log-normal BSV for all parameters**: check ETA distributions
- **Not checking parameter correlations**: High correlations can destabilize the model
:::

::: {.key-box}
### Key Takeaway

Population PK uses mixed-effects models to analyze all subjects simultaneously, quantifying both population-typical parameters and between-subject variability. Covariates like weight and renal function explain variability and enable dose individualization. Always check diagnostics and shrinkage before interpreting individual estimates.
:::

