---
title: "Model Building and Diagnostics"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Choose an appropriate structural PK/PD model
- Select and evaluate residual error models
- Create goodness-of-fit plots for model evaluation
- Perform visual predictive checks (VPC)
- Use bootstrap for uncertainty assessment
- Apply objective function criteria for model selection

---

## The Model Building Workflow

```{r}
#| echo: false
library(ggplot2)

# Model building workflow as a diagram
workflow_text <- data.frame(
  step = 1:7,
  label = c("1. Exploratory Data Analysis",
            "2. Choose Structural Model",
            "3. Select Error Model",
            "4. Covariate Screening",
            "5. Full Model Building",
            "6. Model Refinement",
            "7. Model Validation"),
  y = 7:1
)

ggplot(workflow_text, aes(1, y, label = label)) +
  geom_text(size = 4, color = "#0B6E6B", fontface = "bold") +
  geom_segment(aes(x = 0.9, xend = 0.9, y = 6.5, yend = 0.5),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "#C44536", linewidth = 1) +
  xlim(0.5, 1.5) + ylim(0, 8) +
  labs(title = "PK/PD Model Building Workflow") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

---

## Choosing the Structural Model

### Decision Criteria

| Criterion | How to Assess |
|-----------|--------------|
| **Scientific plausibility** | Does the model reflect known physiology? |
| **Data support** | Can the data distinguish between models? |
| **Objective function (OFV)** | Lower OFV = better fit |
| **AIC/BIC** | Penalizes for extra parameters |
| **Goodness-of-fit plots** | Visual assessment of fit quality |
| **Parameter precision** | RSE% should be reasonable (< 30–50%) |

```{r}
library(dplyr)
library(ggplot2)

# Simulate data from a two-compartment model
set.seed(123)
sample_times <- c(0.05, 0.1, 0.25, 0.5, 1, 2, 4, 6, 8, 12, 24)
true_A <- 15; true_alpha <- 2.5
true_B <- 8;  true_beta  <- 0.25

obs <- data.frame(
  time = sample_times,
  conc = true_A * exp(-true_alpha * sample_times) +
         true_B * exp(-true_beta * sample_times)
)
obs$conc <- obs$conc * exp(rnorm(nrow(obs), 0, 0.1))

# Fit one-compartment
fit1 <- nls(conc ~ C0 * exp(-ke * time), data = obs,
            start = list(C0 = 20, ke = 0.3))

# Fit two-compartment
fit2 <- nls(conc ~ A * exp(-alpha * time) + B * exp(-beta * time),
            data = obs,
            start = list(A = 15, alpha = 2, B = 8, beta = 0.2),
            lower = c(0, 0, 0, 0), algorithm = "port")

# Model comparison
model_comp <- data.frame(
  Model = c("One-Compartment", "Two-Compartment"),
  AIC = round(c(AIC(fit1), AIC(fit2)), 1),
  BIC = round(c(AIC(fit1) + 2*log(nrow(obs)), AIC(fit2) + 4*log(nrow(obs))), 1),
  Parameters = c(2, 4)
)
knitr::kable(model_comp, caption = "Model Comparison")
```

---

## Residual Error Models

### Common Error Models

| Model | Equation | When to Use |
|-------|----------|-------------|
| **Additive** | $Y = F + \varepsilon$ | Constant error across concentrations |
| **Proportional** | $Y = F \cdot (1 + \varepsilon)$ | Error scales with concentration |
| **Combined** | $Y = F \cdot (1 + \varepsilon_1) + \varepsilon_2$ | Both components present |
| **Log-additive** | $\log(Y) = \log(F) + \varepsilon$ | Lognormal error, PK default |

```{r}
# Visualize different error models
conc_true <- 10^seq(-1, 2, length.out = 100)
set.seed(123)

add_err  <- 2
prop_err <- 0.2

df_errors <- data.frame(
  conc = rep(conc_true, 3),
  Error_Model = rep(c("Additive (σ=2)", "Proportional (σ=20%)", "Combined"), each = 100),
  SD = c(rep(add_err, 100),
         prop_err * conc_true,
         sqrt(add_err^2 + (prop_err * conc_true)^2))
)

ggplot(df_errors, aes(conc, SD, color = Error_Model)) +
  geom_line(linewidth = 1.2) +
  scale_x_log10() +
  labs(title = "Residual Error: Standard Deviation vs Concentration",
       x = "True Concentration (log scale)", y = "Residual SD") +
  scale_color_manual(values = c("#0B6E6B", "#C44536", "#D4A843")) +
  theme_minimal()
```

---

## Goodness-of-Fit Plots

### The Four Essential Plots

```{r}
# Generate example diagnostic data
set.seed(42)
n <- 100
pred <- runif(n, 0.5, 50)
obs <- pred * exp(rnorm(n, 0, 0.2))
ipred <- obs  # simplified
time <- 1:n

diag_data <- data.frame(
  DV = obs, PRED = pred, IPRED = ipred,
  CWRES = rnorm(n, 0, 1), TIME = time,
  IWRES = (log(obs) - log(pred)) / 0.2
)

# 1. DV vs PRED
p1 <- ggplot(diag_data, aes(PRED, DV)) +
  geom_point(alpha = 0.6, color = "#0B6E6B") +
  geom_abline(slope = 1, intercept = 0, color = "#C44536", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "#D4A843", linewidth = 1) +
  labs(title = "DV vs PRED", x = "Population Predicted", y = "Observed") +
  theme_minimal()

# 2. DV vs IPRED
p2 <- ggplot(diag_data, aes(IPRED, DV)) +
  geom_point(alpha = 0.6, color = "#0B6E6B") +
  geom_abline(slope = 1, intercept = 0, color = "#C44536", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "#D4A843", linewidth = 1) +
  labs(title = "DV vs IPRED", x = "Individual Predicted", y = "Observed") +
  theme_minimal()

# 3. CWRES vs PRED
p3 <- ggplot(diag_data, aes(PRED, CWRES)) +
  geom_point(alpha = 0.6, color = "#C44536") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = c(-2, 2), linetype = "dotted", alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "#0B6E6B", linewidth = 1) +
  labs(title = "CWRES vs PRED", x = "Population Predicted", y = "CWRES") +
  theme_minimal()

# 4. CWRES vs TIME
p4 <- ggplot(diag_data, aes(TIME, CWRES)) +
  geom_point(alpha = 0.6, color = "#C44536") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "#0B6E6B", linewidth = 1) +
  labs(title = "CWRES vs TIME", x = "Time", y = "CWRES") +
  theme_minimal()

library(patchwork)
(p1 + p2) / (p3 + p4)
```

---

## Visual Predictive Check (VPC)

The VPC simulates data from the final model and compares the observed data distribution to the simulated distribution:

```{r}
# Simulate a VPC
library(rxode2)

set.seed(42)
n_sim <- 500
times <- seq(0, 24, by = 0.5)

# Simulate from a reference model
mod_ref <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - (CL/V) * central
  conc <- central / V
})

# Population parameters with variability
pop_params <- c(ka = 1.0, CL = 5, V = 50)

vpc_data <- lapply(1:n_sim, function(i) {
  params <- c(
    ka = 1.0 * exp(rnorm(1, 0, 0.3)),
    CL = 5.0 * exp(rnorm(1, 0, 0.3)),
    V  = 50 * exp(rnorm(1, 0, 0.2))
  )
  event <- et(time = 0, amt = 500)
  sim <- rxSolve(mod_ref, params, event %>% et(times))
  as.data.frame(sim) %>% mutate(sim_id = i)
})

vpc_df <- bind_rows(vpc_data)

# Compute percentiles of simulated data
vpc_summary <- vpc_df %>%
  group_by(time) %>%
  summarise(
    p05 = quantile(conc, 0.05),
    p50 = quantile(conc, 0.50),
    p95 = quantile(conc, 0.95),
    .groups = "drop"
  )

# Generate "observed" data (a single trial)
set.seed(99)
obs_data <- data.frame(
  time = rep(c(0.5, 1, 2, 4, 6, 8, 12, 24), each = 5),
  ID = rep(1:5, 8)
) %>%
  mutate(
    conc = 500/50 * exp(-5/50 * time) * exp(rnorm(n(), 0, 0.2))
  )

# VPC plot
ggplot() +
  geom_ribbon(data = vpc_summary, aes(time, ymin = p05, ymax = p95),
              fill = "#0B6E6B", alpha = 0.15) +
  geom_line(data = vpc_summary, aes(time, p50),
            color = "#0B6E6B", linewidth = 1, linetype = "dashed") +
  geom_point(data = obs_data, aes(time, conc, group = ID),
             color = "#C44536", size = 1.5, alpha = 0.6) +
  scale_y_log10() +
  labs(title = "Visual Predictive Check (VPC)",
       subtitle = "Shaded area: 90% prediction interval; Dashed: median; Points: observed",
       x = "Time (hours)", y = "Concentration (mg/L, log scale)") +
  theme_minimal()
```

---

## Bootstrap for Uncertainty

Bootstrap resampling provides non-parametric confidence intervals for parameter estimates:

```{r}
#| eval: false
# Simplified bootstrap example
n_boot <- 200
bootstrap_params <- matrix(NA, nrow = n_boot, ncol = 3)

for (i in 1:n_boot) {
  # Resample subjects with replacement
  boot_ids <- sample(unique(pk_data$ID), replace = TRUE)

  boot_data <- do.call(rbind, lapply(seq_along(boot_ids), function(j) {
    orig_data <- pk_data[pk_data$ID == boot_ids[j], ]
    orig_data$ID <- j  # Rename to avoid duplicates
    orig_data
  }))

  # Fit model to bootstrap sample and store parameters
  # bootstrap_params[i, ] <- coef(fit_boot)
}
```

---

## Model Selection Criteria

| Criterion | Formula | Notes |
|-----------|---------|-------|
| **OFV** | $-2 \times \log(Likelihood)$ | Lower = better |
| **AIC** | $OFV + 2p$ | Penalizes for $p$ parameters |
| **BIC** | $OFV + p \cdot \ln(n)$ | Stronger penalty for large $n$ |
| **ΔOFV** | $OFV_1 - OFV_2$ | $\sim \chi^2$ with df = $p_2 - p_1$ |

For nested models, a drop in OFV > 3.84 (p < 0.05, 1 df) is considered significant.

---

## Model Selection Decision Tree

```{r}
#| echo: false
# Model selection decision tree as a flowchart

decision_steps <- data.frame(
  Step = c("Is clearance constant (linear PK)?",
            "Does effect lag concentration (hysteresis)?",
            "Multiple subjects with variability to explain?",
            "Need uncertainty quantification / priors?"),
  Yes = c("→ Analytical 1-/2-compartment (Ch 6-7)",
           "→ Add effect compartment (Ch 10)",
           "→ NLME / Covariate model (Ch 12, 16)",
           "→ Bayesian MCMC/Stan (Ch 18)"),
  No  = c("→ ODE with MM/induction/inhibition (Ch 11, 16)",
           "→ Direct Cp→E Emax/Hill model (Ch 9)",
           "→ Standard NLS/MLE (Ch 9, 13)",
           "→ Frequentist NLS/MLE/NLME suffices")
)

knitr::kable(decision_steps,
  caption = "Model Selection Decision Tree: Four Key Branching Points")
```

---

## Identifiability: Can Your Data Support Your Model?

**Identifiability** asks whether unique parameter values can be recovered from available data. An unidentifiable model can produce the same fit with very different parameter combinations — making estimates meaningless.

::: {.warning-box}
### Identifiability Red Flags

1. **Too many parameters for the data**: Estimating > 4-5 parameters simultaneously from one dataset without fixing any externally-derivable values (e.g., $V$ from NCA, $k_a$ from literature)

2. **Coupled multi-state systems**: PK+PD+disease or two-drug DDI models where the parameter count grows faster than informative data points. Each new state typically adds 2-4 parameters.

3. **Flat or ridge-shaped likelihood/posterior**: Check via **profile likelihood** (fix one parameter at a range of values, re-optimize the rest, watch OFV) or **posterior pairs plots** — if the OFV barely changes as a parameter varies over a wide range, that parameter is unidentifiable.

4. **High parameter correlations**: $|\rho| > 0.95$ between estimates (e.g., $CL$ and $V$ in oral dosing without IV data). The correlation matrix from `nls`/`nlme` output is your first diagnostic.

5. **$k_a$ and $k_e$ in oral data**: Without early sampling (< 30 min post-dose), these two rate constants are notoriously confounded — the "flip-flop" phenomenon.
:::

### The Sensitivity Triage Principle

**Run sensitivity analysis before estimation** to triage which parameters to fit vs fix. Parameters with first-order Sobol index $S_i < 0.05$ (explain < 5% of output variance) are strong candidates for fixing at literature values. This reduces the estimation burden and improves identifiability of the remaining parameters. See @sec-sensitivity for full methodology.

---

## The Estimation Ladder

PK/PD parameter estimation forms a ladder of increasing complexity and information:

| Rung | Method | R Function | Use When |
|---|---|---|---|
| 1 | **OLS** (linearized) | `lm()` | Quick initial estimates; log-transform C vs t for $k_e$ |
| 2 | **WLS** | `lm(weights = ...)` | Heteroscedastic residuals (variance not constant) |
| 3 | **NLS** | `nls()` / `nlsLM()` | Nonlinear model with additive error on natural scale |
| 4 | **MLE** | `stats4::mle()` | Need $\sigma$, asymptotic CIs, non-normal errors |
| 5 | **Bayesian** | Stan / `brms` | Priors needed, full posterior uncertainty, sparse data |

Each rung costs more computation but reduces bias (OLS linearization bias) or adds information (uncertainty quantification in Bayesian). Start low on the ladder for exploration, then climb as needed.

::: {.exercise-box}
### Exercises

1. Compare a one- vs two-compartment model using AIC and BIC. Which is preferred?
2. Create the four essential goodness-of-fit plots for a fitted model. What patterns indicate a good fit?
3. Generate a VPC for a PK model. Interpret what it means if >10% of observed points fall outside the 90% PI.
4. What does a funnel shape in CWRES vs PRED indicate about your residual error model?
5. Use bootstrap to compute 95% CI for CL from a population PK model.
:::

::: {.mistakes-box}
### Common Mistakes

- **Relying only on OFV**: Always check GOF plots, parameter precision, and scientific plausibility
- **Ignoring correlated residuals**: Time series structure in residuals suggests model misspecification
- **VPC with too few simulations**: Use at least 500–1000 simulations
- **Over-interpreting small OFV drops**: ΔOFV < 3.84 may be noise, not signal
- **Not checking parameter identifiability**: High RSE (> 50%) or correlations > 0.95 are red flags
:::

::: {.key-box}
### Key Takeaway

Model evaluation is as important as model fitting. Use a combination of OFV/AIC, goodness-of-fit plots, VPC, and bootstrap to build confidence in your model. A good model is not just statistically adequate — it must also be scientifically plausible and practically useful.
:::

