---
title: "Simulation and Dose Optimization"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Simulate dosing regimens for multiple-dose scenarios
- Predict accumulation and steady-state concentrations
- Select optimal dosing intervals based on PK/PD targets
- Compute probability of target attainment (PTA)
- Apply therapeutic drug monitoring concepts
- Perform individualized dose adjustments

---

## Why Simulate?

Clinical trial simulation allows you to explore "what-if" scenarios without exposing patients to risk:

1. **Dose selection**: What dose achieves the desired exposure?
2. **Dosing interval**: How often should the drug be given?
3. **Special populations**: How should we dose in renal impairment?
4. **Drug interactions**: What happens when co-administered with an inhibitor?
5. **Missed doses**: What if a patient misses a dose?

---

## Multiple Dosing and Accumulation

### The Accumulation Principle

With repeated dosing at interval $\tau$, drug accumulates until the amount eliminated during each interval equals the dose:

$$R_{acc} = \frac{1}{1 - e^{-k_e \tau}}$$

```{r}
library(rxode2)
library(ggplot2)
library(dplyr)

# Simulate multiple-dose IV bolus
mod_iv <- rxode2({
  d/dt(central) <- -ke * central
  conc <- central / V
})

ke <- 0.3; V <- 10; tau <- 6  # q6h dosing
n_doses <- 10
dose_times <- seq(0, (n_doses - 1) * tau, by = tau)
event <- et(time = dose_times, amt = rep(100, n_doses))
times <- seq(0, n_doses * tau, by = 0.05)

sim_multi <- rxSolve(mod_iv, c(ke = ke, V = V), event %>% et(times))

# Compute accumulation factor
R_acc <- 1 / (1 - exp(-ke * tau))

ggplot(as.data.frame(sim_multi), aes(time, conc)) +
  geom_line(color = "#0B6E6B", linewidth = 0.8) +
  labs(title = "Multiple IV Bolus Dosing (q6h)",
       subtitle = paste0("Accumulation factor = ", round(R_acc, 2),
                         "; 97% steady state by ~", round(5 * log(2)/ke, 1), " h"),
       x = "Time (hours)", y = "Concentration (mg/L)") +
  theme_minimal()
```

### Peak and Trough at Steady State

For IV bolus at steady state:

$$C_{ss,max} = \frac{D}{V} \cdot \frac{1}{1 - e^{-k_e \tau}}$$
$$C_{ss,min} = \frac{D}{V} \cdot \frac{e^{-k_e \tau}}{1 - e^{-k_e \tau}}$$

```{r}
# Steady-state peak and trough vs dosing interval
tau_seq <- seq(2, 24, by = 0.5)
D <- 100; V <- 10; ke <- 0.3

Css_max <- (D/V) / (1 - exp(-ke * tau_seq))
Css_min <- (D/V) * exp(-ke * tau_seq) / (1 - exp(-ke * tau_seq))

df_ss <- data.frame(
  tau = rep(tau_seq, 2),
  conc = c(Css_max, Css_min),
  Type = rep(c("Css,max", "Css,min"), each = length(tau_seq))
)

ggplot(df_ss, aes(tau, conc, color = Type)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Steady-State Peak and Trough vs Dosing Interval",
       x = "Dosing Interval (hours)", y = "Concentration (mg/L)") +
  scale_color_manual(values = c("#C44536", "#0B6E6B")) +
  theme_minimal()
```

---

## Oral Dosing at Steady State

### Average Steady-State Concentration

$$\bar{C}_{ss} = \frac{F \cdot D}{CL \cdot \tau}$$

This is dose proportional and independent of $V$ and $k_a$.

```{r}
# Simulate multiple oral doses
mod_oral <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - (CL/V) * central
  conc <- central / V
})

params <- c(ka = 1.0, CL = 5, V = 50)
tau_oral <- 12
dose_times <- seq(0, 5 * tau_oral, by = tau_oral)
event_oral <- et(time = dose_times, amt = rep(250, length(dose_times)))
times_oral <- seq(0, 6 * tau_oral, by = 0.1)

sim_ss <- rxSolve(mod_oral, params, event_oral %>% et(times_oral))

# Compute average Css
F_bio <- 1.0
Css_avg <- F_bio * 250 / (params["CL"] * tau_oral)

ggplot(as.data.frame(sim_ss), aes(time, conc)) +
  geom_line(color = "#0B6E6B", linewidth = 0.8) +
  geom_hline(yintercept = Css_avg, linetype = "dashed", color = "#C44536") +
  annotate("text", x = max(times_oral) * 0.7, y = Css_avg * 0.85,
           label = paste0("Avg Css = ", round(Css_avg, 1), " mg/L"),
           color = "#C44536") +
  labs(title = "Oral Dosing to Steady State (q12h)",
       x = "Time (hours)", y = "Concentration (mg/L)") +
  theme_minimal()
```

---

## Probability of Target Attainment (PTA)

PTA answers: *Given variability, what is the probability that a dosing regimen achieves the target?*

### Target Examples

| Target Type | Example | Definition |
|-------------|---------|-----------|
| $C_{min} > MIC$ | Antibiotics | Trough above minimum inhibitory concentration |
| $C_{max} < Toxicity$ | Safety | Peak below toxicity threshold |
| $AUC/MIC > 125$ | Fluoroquinolones | Exposure ratio |
| $T > MIC$ (40–60%) | β-lactams | Time above MIC |

```{r}
# PTA simulation: Probability of Cmin > MIC
set.seed(123)
n_sim <- 5000

# Population PK parameters
pop_CL <- 5.0; omega_CL <- 0.3  # L/h
pop_V  <- 50;  omega_V  <- 0.2  # L

# Regimen
dose <- 500  # mg
tau  <- 8    # hours

# Target
MIC <- 1.0   # mg/L

# Simulate
sim_CL <- pop_CL * exp(rnorm(n_sim, 0, omega_CL))
sim_V  <- pop_V  * exp(rnorm(n_sim, 0, omega_V))
sim_ke <- sim_CL / sim_V

# Cmin at steady state
Cmin_ss <- (dose / sim_V) * exp(-sim_ke * tau) / (1 - exp(-sim_ke * tau))

# PTA
pta <- mean(Cmin_ss > MIC)
cat("Probability of Target Attainment (Cmin > MIC):", round(pta * 100, 1), "%\n")
```

```{r}
# Visualize PTA distribution
ggplot(data.frame(Cmin = Cmin_ss), aes(Cmin)) +
  geom_histogram(fill = "#0B6E6B", alpha = 0.7, bins = 50) +
  geom_vline(xintercept = MIC, color = "#C44536", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = MIC * 1.5, y = 200,
           label = paste0("PTA = ", round(pta * 100, 1), "%\n",
                          "(Cmin > MIC = ", MIC, " mg/L)"),
           color = "#C44536", hjust = 0) +
  labs(title = "Probability of Target Attainment",
       subtitle = "Distribution of steady-state Cmin across 5000 simulated subjects",
       x = "Cmin,ss (mg/L)", y = "Count") +
  theme_minimal()
```

---

## Dose Optimization by Target

### PTA vs Dose

```{r}
# PTA across dose range
dose_range <- seq(100, 2000, by = 50)
pta_by_dose <- sapply(dose_range, function(d) {
  Cmin_ss <- (d / sim_V) * exp(-sim_ke * tau) / (1 - exp(-sim_ke * tau))
  mean(Cmin_ss > MIC)
})

df_pta <- data.frame(Dose = dose_range, PTA = pta_by_dose)

ggplot(df_pta, aes(Dose, PTA)) +
  geom_line(color = "#0B6E6B", linewidth = 1.2) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "#C44536", alpha = 0.7) +
  geom_vline(xintercept = dose_range[which.min(abs(pta_by_dose - 0.9))],
             linetype = "dashed", color = "#D4A843", alpha = 0.7) +
  labs(title = "PTA vs Dose",
       subtitle = "Dose needed for 90% PTA",
       x = "Dose (mg)", y = "Probability of Target Attainment") +
  theme_minimal()
```

---

## Therapeutic Drug Monitoring (TDM)

### Bayesian Dose Adjustment

After measuring a drug concentration in a patient, Bayesian methods combine the population prior with the individual observation to compute a posterior:

$$P(\theta_i \mid C_{obs}) \propto P(C_{obs} \mid \theta_i) \cdot P(\theta_i)$$

```{r}
# Simplified Bayesian TDM example
# Prior: CL ~ lognormal(log(5), 0.3)
# Observation: single concentration = 8 mg/L at t = 2h after 500 mg IV

# Prior density
CL_seq <- seq(1, 20, length.out = 200)
prior_CL <- dlnorm(CL_seq, log(5), 0.3)

# Likelihood (simplified: assuming V=50 fixed)
V_fixed <- 50
conc_obs <- 8; t_obs <- 2; dose <- 500
likelihood <- dnorm(conc_obs,
                    mean = (dose / V_fixed) * exp(-(CL_seq / V_fixed) * t_obs),
                    sd = 1.0)

# Posterior (unnormalized)
posterior_CL <- prior_CL * likelihood
posterior_CL <- posterior_CL / sum(posterior_CL) / diff(CL_seq[1:2])

df_tdm <- data.frame(
  CL = rep(CL_seq, 2),
  Density = c(prior_CL, posterior_CL),
  Distribution = rep(c("Prior", "Posterior"), each = length(CL_seq))
)

ggplot(df_tdm, aes(CL, Density, color = Distribution)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Bayesian TDM: Prior to Posterior Update",
       subtitle = paste0("Observed conc = ", conc_obs, " mg/L at t = ", t_obs, " h"),
       x = "Clearance (L/h)", y = "Density") +
  scale_color_manual(values = c("#0B6E6B", "#C44536")) +
  theme_minimal()
```

---

## Dose Individualization

### Loading Dose

When rapid attainment of therapeutic concentrations is needed:

$$D_{loading} = \frac{C_{target} \cdot V}{F}$$

### Maintenance Dose

$$D_{maintenance} = \frac{C_{target} \cdot CL \cdot \tau}{F}$$

```{r}
# Dose individualization example
calculate_dose <- function(target_conc, CL, V, tau, F = 1) {
  loading  <- target_conc * V / F
  maintenance <- target_conc * CL * tau / F

  data.frame(
    Parameter = c("Loading Dose", "Maintenance Dose", "Dosing Interval", "Target Css"),
    Value = c(paste0(round(loading, 0), " mg"),
              paste0(round(maintenance, 0), " mg"),
              paste0(tau, " h"),
              paste0(target_conc, " mg/L"))
  )
}

# Example: 70 kg patient with CL=5 L/h, V=50 L
calculate_dose(target_conc = 15, CL = 5, V = 50, tau = 8)
```

---

## Monte Carlo Population Simulation

When you need to explore how physiological variability affects drug exposure — but don't yet have data to fit a full NLME model — **Monte Carlo population simulation** provides a lightweight alternative.

### The Pattern

1. Sample PK parameters from log-normal distributions for $N$ virtual subjects
2. Simulate each subject's concentration-time profile
3. Aggregate results (e.g., compute percentiles of AUC, $C_{max}$)

This is distinct from full PopPK: distribution parameters are specified from literature rather than estimated from data.

```{r}
library(rxode2)
library(ggplot2)
library(dplyr)

# One-compartment oral model
mod_oral <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - (CL/V) * central
  conc <- central / V
})

set.seed(42)
n_virtual <- 200

# Sample parameters from literature-based distributions
# CL ~ LogNormal(8, 0.35), V ~ LogNormal(65, 0.25), ka ~ LogNormal(1.2, 0.35)
pop_sim <- data.frame(
  ID = 1:n_virtual,
  CL = rlnorm(n_virtual, log(8), 0.35),
  V  = rlnorm(n_virtual, log(65), 0.25),
  ka = rlnorm(n_virtual, log(1.2), 0.35)
)

# Simulate each subject
times <- seq(0, 24, by = 0.5)

sim_monte_carlo <- lapply(1:n_virtual, function(i) {
  params <- c(ka = pop_sim$ka[i], CL = pop_sim$CL[i], V = pop_sim$V[i])
  event <- et(time = 0, amt = 750)
  sim <- rxSolve(mod_oral, params, event %>% et(times))
  df <- as.data.frame(sim)
  df$ID <- i
  df
})

df_mc <- bind_rows(sim_monte_carlo)

# Compute population percentiles
mc_summary <- df_mc %>%
  group_by(time) %>%
  summarise(
    p05 = quantile(conc, 0.05),
    p25 = quantile(conc, 0.25),
    p50 = quantile(conc, 0.50),
    p75 = quantile(conc, 0.75),
    p95 = quantile(conc, 0.95),
    .groups = "drop"
  )

ggplot() +
  geom_ribbon(data = mc_summary, aes(time, ymin = p05, ymax = p95),
              fill = "#0B6E6B", alpha = 0.1) +
  geom_ribbon(data = mc_summary, aes(time, ymin = p25, ymax = p75),
              fill = "#0B6E6B", alpha = 0.2) +
  geom_line(data = mc_summary, aes(time, p50),
            color = "#0B6E6B", linewidth = 1.2) +
  labs(title = "Monte Carlo Population Simulation",
       subtitle = paste0(n_virtual, " virtual subjects; dark band: IQR, light band: 90% PI"),
       x = "Time (hours)", y = "Concentration (mg/L)") +
  theme_minimal()
```

### When to Use This vs Full NLME

| Scenario | Monte Carlo | Full NLME |
|---|---|---|
| No observed data yet (trial planning) | ✓ Fast, literature-based | ✗ Nothing to fit |
| Rich sampling, single subject | ✗ Unnecessary | ✗ Overkill — use NLS |
| Sparse data, many subjects | ✗ Distributions are guesswork | ✓ Estimate $\omega$, $\sigma$ from data |
| Exploring "what-if" sensitivity to BSV magnitude | ✓ Vary $\omega$ and observe impact | ✗ Requires re-fitting |

::: {.key-box}
### Ground Your Distributions

Monte Carlo population simulation is fast and useful for trial/sampling-design decisions, but the distribution parameters ("arbitrary for demonstration") should be **grounded in literature or existing PopPK estimates** before use in real design decisions.
:::

---

::: {.exercise-box}
### Exercises

1. Compute the accumulation factor for a drug with $t_{1/2}$ = 12 h dosed q12h, q24h, and q48h.
2. Simulate oral dosing to steady state and compute $C_{ss,avg}$, $C_{ss,max}$, and $C_{ss,min}$.
3. Perform a PTA analysis: target $AUC/MIC > 125$, $MIC = 2$ mg/L, dose = 500 mg q12h. Vary CL from 3 to 10 L/h.
4. A patient has CL = 3 L/h (low clearance). Adjust the maintenance dose to achieve the same $C_{ss,avg}$ as a typical patient with CL = 5 L/h.
5. Simulate a loading dose + maintenance regimen and compare to maintenance-only in time to reach therapeutic range.
:::

::: {.mistakes-box}
### Common Mistakes

- **Forgetting that time to steady state depends on $t_{1/2}$, not dosing interval**
- **Using $C_{avg}$ when $C_{min}$ or $C_{max}$ is the actual target**
- **Assuming PTA from limited simulations without checking convergence**
- **Not accounting for between-occasion variability in TDM**
- **Using loading dose without checking for toxicity at peak**
:::

::: {.key-box}
### Key Takeaway

Simulation is the bridge from PK/PD models to clinical decisions. Master accumulation, steady state, PTA, and TDM concepts. The ultimate goal is to select the right dose for the right patient at the right time — model-informed precision dosing.
:::

