---
title: "Exposure–Response and Clinical Outcomes"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Link drug exposure (PK) to clinical efficacy and safety endpoints
- Model biomarker trajectories and surrogate endpoints
- Analyze time-to-event outcomes in relation to exposure
- Assess covariate effects on exposure-response relationships
- Translate modeling results into clinical decisions and benefit-risk assessments

---

## What Is Exposure-Response Analysis?

Exposure-response (E-R) analysis relates a measure of drug exposure (e.g., AUC, $C_{avg}$, $C_{trough}$) to a clinical outcome (efficacy or safety). It answers:

- Is higher exposure associated with better efficacy?
- Is higher exposure associated with more toxicity?
- What exposure range provides the optimal benefit-risk balance?

---

## Types of Clinical Endpoints

| Endpoint Type | Example | Statistical Model |
|---------------|---------|------------------|
| **Continuous** | Blood pressure change (mmHg) | Linear/Emax regression |
| **Binary** | Response (yes/no) | Logistic regression |
| **Categorical** | AE grade (0–4) | Proportional odds |
| **Time-to-event** | Time to progression/death | Cox PH, parametric survival |
| **Count** | Number of seizures | Poisson/negative binomial |
| **Longitudinal** | Repeated biomarker measurements | Mixed-effects models |

---

## Continuous Efficacy Endpoints

```{r}
library(ggplot2)
library(dplyr)

# Simulate exposure-response data for a continuous endpoint
set.seed(123)
n_patients <- 150

er_data <- data.frame(
  AUC = rlnorm(n_patients, log(500), 0.4),  # AUC in mg·h/L
  Emax = -15, EC50 = 800, E0 = 0
) %>%
  mutate(
    Effect_true = E0 + (Emax * AUC) / (EC50 + AUC),
    Effect_obs  = Effect_true + rnorm(n(), 0, 3)
  )

# Fit Emax model
fit_er <- nls(Effect_obs ~ E0 + (Emax * AUC) / (EC50 + AUC),
              data = er_data,
              start = list(E0 = 0, Emax = -10, EC50 = 500))
summary(fit_er)
```

```{r}
# Plot E-R relationship
auc_pred <- seq(min(er_data$AUC), max(er_data$AUC), length.out = 200)
pred_effect <- predict(fit_er, newdata = data.frame(AUC = auc_pred))

ggplot(er_data, aes(AUC, Effect_obs)) +
  geom_point(alpha = 0.5, color = "gray50") +
  geom_line(data = data.frame(AUC = auc_pred, Effect = pred_effect),
            aes(AUC, Effect), color = "#C44536", linewidth = 1.2) +
  geom_hline(yintercept = -5, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 1500, y = -5, label = "Target effect", hjust = 0, size = 3) +
  labs(title = "Exposure-Response: Continuous Efficacy Endpoint",
       x = "AUC (mg·h/L)", y = "Change from Baseline (mmHg)") +
  theme_minimal()
```

---

## Binary Endpoints: Response Rate

```{r}
# Simulate binary response data
er_binary <- er_data %>%
  mutate(
    # Probability of response increases with AUC
    prob_response = plogis(-2 + 0.004 * AUC),
    Response = rbinom(n(), 1, prob_response)
  )

# Logistic regression
fit_logistic <- glm(Response ~ AUC, data = er_binary, family = binomial)
summary(fit_logistic)

# Predict response rate vs AUC
auc_seq <- seq(100, 3000, length.out = 200)
pred_prob <- predict(fit_logistic, newdata = data.frame(AUC = auc_seq),
                     type = "response")

ggplot() +
  # Observed (binned)
  geom_point(data = er_binary %>%
    mutate(AUC_bin = cut(AUC, breaks = 5)) %>%
    group_by(AUC_bin) %>%
    summarise(
      AUC_mid = mean(AUC),
      Resp_Rate = mean(Response),
      N = n(),
      .groups = "drop"
    ),
    aes(AUC_mid, Resp_Rate, size = N), color = "#0B6E6B") +
  # Model fit
  geom_line(data = data.frame(AUC = auc_seq, Prob = pred_prob),
            aes(AUC, Prob), color = "#C44536", linewidth = 1.2) +
  labs(title = "Exposure-Response: Binary Response Endpoint",
       x = "AUC (mg·h/L)", y = "Probability of Response", size = "N") +
  theme_minimal()
```

---

## Time-to-Event Analysis

```{r}
library(survival)

# Simulate time-to-event data (progression-free survival)
set.seed(456)
n_pfs <- 200

pfs_data <- data.frame(
  ID = 1:n_pfs,
  AUC = rlnorm(n_pfs, log(500), 0.4),
  # Higher AUC = longer PFS
  log_HR = -0.002,  # log hazard ratio per unit AUC
  baseline_hazard = 0.02
) %>%
  mutate(
    # Simulate event times using exponential distribution
    lambda = baseline_hazard * exp(log_HR * AUC),
    Time = rexp(n(), lambda),
    Censor_Time = runif(n(), 3, 24),  # random censoring
    Status = ifelse(Time <= Censor_Time, 1, 0),
    Obs_Time = pmin(Time, Censor_Time)
  )

# Cox proportional hazards model
fit_cox <- coxph(Surv(Obs_Time, Status) ~ AUC, data = pfs_data)
summary(fit_cox)
```

```{r}
# Kaplan-Meier by AUC quartiles
pfs_data <- pfs_data %>%
  mutate(AUC_quartile = factor(ntile(AUC, 4),
                                labels = c("Q1 (Low)", "Q2", "Q3", "Q4 (High)")))

fit_km <- survfit(Surv(Obs_Time, Status) ~ AUC_quartile, data = pfs_data)

library(survminer)
ggsurvplot(fit_km, data = pfs_data,
           palette = c("#0B6E6B", "#2E8B7E", "#D4A843", "#C44536"),
           risk.table = TRUE,
           xlab = "Time (months)", ylab = "Progression-Free Survival",
           title = "PFS by AUC Quartile",
           legend.title = "Exposure")
```

---

## Biomarkers and Surrogate Endpoints

A **biomarker** is a measurable indicator of a biological process. A **surrogate endpoint** is a biomarker intended to substitute for a clinical endpoint.

### Modeling a Biomarker Trajectory

```{r}
# Simulate longitudinal biomarker data
set.seed(789)
n_subj <- 30
times <- c(0, 1, 2, 4, 8, 12, 24)  # weeks

biomarker_data <- expand.grid(ID = 1:n_subj, Time = times) %>%
  mutate(
    AUC = rep(rlnorm(n_subj, log(500), 0.3), each = length(times)),
    # Biomarker decreases with time, rate depends on AUC
    Biomarker = 100 * exp(-0.1 * exp(log(0.002) + 0.003 * AUC) * Time) +
                rnorm(n(), 0, 5)
  )

ggplot(biomarker_data, aes(Time, Biomarker, group = ID)) +
  geom_line(alpha = 0.3, color = "#0B6E6B") +
  labs(title = "Biomarker Trajectories by Subject",
       x = "Time (weeks)", y = "Biomarker Level") +
  theme_minimal()
```

---

## Covariate Effects on Exposure-Response

```{r}
# Forest plot: covariate effects on E-R parameters
covariate_effects <- data.frame(
  Covariate = c("Age > 65", "Weight > 90 kg", "Male sex", "eGFR < 60", "CYP2D6 PM"),
  Effect_on_EC50 = c(1.35, 0.85, 1.10, 1.50, 2.20),
  Lower_CI = c(1.05, 0.65, 0.90, 1.20, 1.60),
  Upper_CI = c(1.75, 1.10, 1.35, 1.90, 3.00)
)

ggplot(covariate_effects, aes(x = Effect_on_EC50, y = reorder(Covariate, Effect_on_EC50))) +
  geom_point(color = "#0B6E6B", size = 3) +
  geom_errorbarh(aes(xmin = Lower_CI, xmax = Upper_CI),
                 height = 0.2, color = "#0B6E6B", linewidth = 1) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "#C44536") +
  labs(title = "Covariate Effects on EC50",
       subtitle = "Values > 1 indicate reduced potency (higher EC50)",
       x = "Fold Change in EC50 (95% CI)", y = "") +
  theme_minimal()
```

---

## Benefit-Risk Assessment

The ultimate goal is to balance efficacy and safety:

```{r}
# Simulate efficacy and safety vs exposure
exposure_seq <- seq(100, 3000, length.out = 200)

# Efficacy: Emax model (higher = better)
efficacy <- 0 + (80 * exposure_seq) / (800 + exposure_seq)

# Safety (adverse event rate): logistic model
safety <- plogis(-3 + 0.002 * exposure_seq)

# Therapeutic window
df_benefit_risk <- data.frame(
  Exposure = rep(exposure_seq, 2),
  Probability = c(efficacy / 100, safety),
  Outcome = rep(c("Efficacy (Response Rate)", "Safety (AE Rate)"), each = length(exposure_seq))
)

ggplot(df_benefit_risk, aes(Exposure, Probability, color = Outcome)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = c(500, 1500), linetype = "dashed", alpha = 0.5) +
  annotate("rect", xmin = 500, xmax = 1500, ymin = 0, ymax = 1,
           fill = "#0B6E6B", alpha = 0.05) +
  annotate("text", x = 1000, y = 0.5, label = "Therapeutic\nWindow", size = 4) +
  labs(title = "Benefit-Risk Assessment",
       x = "Drug Exposure (AUC)", y = "Probability") +
  scale_color_manual(values = c("#0B6E6B", "#C44536")) +
  theme_minimal()
```

---

::: {.exercise-box}
### Exercises

1. Fit an Emax model to relate AUC to a continuous efficacy endpoint. What is the estimated $EC_{50}$?
2. Use logistic regression to model response rate vs AUC. Report the odds ratio per 100-unit increase in AUC.
3. Perform a Cox regression for time-to-event vs AUC. Interpret the hazard ratio.
4. Plot a forest plot showing covariate effects on $EC_{50}$ from a population E-R analysis.
5. Determine the therapeutic window where response rate > 60% and AE rate < 10%.
:::

::: {.mistakes-box}
### Common Mistakes

- **Confusing correlation with causation**: E-R relationships from observational data may be confounded
- **Extrapolating beyond observed exposure range**: Never predict outside the data
- **Ignoring time-dependency**: Exposure at week 1 may not predict outcome at week 24
- **Using a single exposure metric**: $C_{trough}$, $C_{avg}$, and AUC may give different E-R relationships
- **Not adjusting for baseline**: Always include baseline severity as a covariate
:::

::: {.key-box}
### Key Takeaway

Exposure-response analysis is the critical step that translates PK/PD modeling into clinical decision-making. Master the relationship between exposure metrics (AUC, $C_{trough}$) and clinical endpoints (continuous, binary, time-to-event). The ultimate goal is to identify the exposure range that maximizes benefit while minimizing risk.
:::

