---
title: "Physiologically Based Pharmacokinetic Modeling"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Understand the principles of whole-body PBPK modeling
- Describe organ compartments and physiological parameters
- Simulate tissue drug concentrations using PBPK models in R
- Understand tissue partitioning and permeability concepts
- Identify when PBPK is preferable to classical compartmental models

---

## What Is PBPK?

**Physiologically based pharmacokinetic (PBPK)** models describe drug disposition using anatomical and physiological compartments connected by blood flow. Unlike classical compartmental models (which are empirical), PBPK models are **mechanistic** — they incorporate:

- Organ volumes and blood flows
- Tissue:plasma partition coefficients
- Protein binding in plasma and tissues
- Metabolic clearance in specific organs (liver, kidney)

PBPK is widely used for:
- First-in-human dose prediction
- Pediatric extrapolation
- Drug-drug interaction predictions
- Formulation development
- Special population extrapolation

---

## PBPK Model Structure

```{r}
#| echo: false
library(ggplot2)

# Conceptual PBPK diagram as text
pbpk_organs <- data.frame(
  Organ = c("Lung", "Brain", "Heart", "Liver\n(Metabolism)", "Kidney\n(Excretion)",
            "Muscle", "Adipose", "Skin", "Bone", "Gut\n(Absorption)"),
  Flow_pct = c(100, 15, 4, 25, 20, 15, 5, 5, 5, 16),
  x = c(0.5, 0.25, 0.35, 0.65, 0.75, 0.15, 0.85, 0.45, 0.55, 0.6),
  y = c(0.95, 0.7, 0.5, 0.3, 0.3, 0.1, 0.1, 0.5, 0.7, 0.1)
)

ggplot(pbpk_organs, aes(x, y, label = Organ, size = Flow_pct)) +
  geom_text(color = "#0B6E6B", fontface = "bold") +
  scale_size(range = c(3, 7), guide = "none") +
  xlim(0, 1) + ylim(0, 1.1) +
  labs(title = "Whole-Body PBPK Model Structure",
       subtitle = "Font size ~ relative blood flow") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 9))
```

---

## Mathematical Structure

For each non-eliminating tissue:

$$V_T \cdot \frac{dC_T}{dt} = Q_T \cdot \left(C_{art} - \frac{C_T}{Kp_T}\right)$$

For eliminating tissues (liver, kidney):

$$V_T \cdot \frac{dC_T}{dt} = Q_T \cdot \left(C_{art} - \frac{C_T}{Kp_T}\right) - CL_{int,T} \cdot C_{u,T}$$

where:
- $V_T$: tissue volume (L)
- $Q_T$: blood flow to tissue (L/h)
- $C_{art}$: arterial concentration
- $C_T$: tissue concentration
- $Kp_T$: tissue:plasma partition coefficient
- $CL_{int,T}$: intrinsic clearance in tissue

---

## Building a Minimal PBPK Model in R

```{r}
library(rxode2)
library(ggplot2)
library(dplyr)

# Simplified PBPK model (lumped tissues: richly perfused + poorly perfused)
mod_pbpk <- rxode2({
  # Concentrations (computed first from compartment amounts)
  C_lu <- A_lu / V_lu
  C_rp <- A_rp / V_rp
  C_pp <- A_pp / V_pp
  C_li <- A_li / V_li

  # Arterial blood (lung output)
  C_art <- C_lu / Kp_lu

  # Venous blood (mixed return from tissues)
  C_ven <- (Q_rp * C_rp/Kp_rp + Q_pp * C_pp/Kp_pp + Q_li * C_li/Kp_li) / QCO

  # Lung
  d/dt(A_lu) <- QCO * (C_ven - C_lu / Kp_lu)

  # Richly perfused tissues (lumped)
  d/dt(A_rp) <- Q_rp * (C_art - C_rp / Kp_rp)

  # Poorly perfused tissues (lumped)
  d/dt(A_pp) <- Q_pp * (C_art - C_pp / Kp_pp)

  # Liver (eliminating organ)
  d/dt(A_li) <- Q_li * (C_art - C_li / Kp_li) - CL_int * (C_li / Kp_li)

  # Output: plasma concentration (approximated as arterial)
  C_plasma <- C_art
})

# Physiological parameters (70 kg human)
QCO <- 5.4 * 60     # Cardiac output: L/h
params_pbpk <- c(
  # Blood flows (L/h)
  QCO = QCO, Q_rp = 0.51 * QCO, Q_pp = 0.19 * QCO, Q_li = 0.25 * QCO,

  # Volumes (L)
  V_lu = 1.0, V_rp = 3.0, V_pp = 40.0, V_li = 1.8,

  # Partition coefficients
  Kp_lu = 1.5, Kp_rp = 2.0, Kp_pp = 0.8, Kp_li = 3.0,

  # Intrinsic clearance (L/h)
  CL_int = 20
)

times_pbpk <- seq(0, 24, by = 0.1)

# IV bolus dose (µg) into the richly perfused/blood pool compartment (A_rp)
event_pbpk <- et(time = 0, amt = 10000, cmt = "A_rp") %>% et(times_pbpk)

sim_pbpk <- rxSolve(mod_pbpk, params_pbpk, event_pbpk)
pbpk_df <- as.data.frame(sim_pbpk)

# Plot plasma and tissue concentrations
ggplot(pbpk_df, aes(time)) +
  geom_line(aes(y = C_plasma, color = "Plasma"), linewidth = 1) +
  geom_line(aes(y = C_rp / params_pbpk[["Kp_rp"]], color = "Richly Perfused"), linewidth = 1) +
  geom_line(aes(y = C_pp / params_pbpk[["Kp_pp"]], color = "Poorly Perfused"), linewidth = 1) +
  scale_y_log10() +
  labs(title = "Minimal PBPK Model: Tissue Concentration Profiles",
       subtitle = "IV bolus, 70 kg human",
       x = "Time (hours)", y = "Concentration (log scale)",
       color = "Compartment") +
  scale_color_manual(values = c("#0B6E6B", "#C44536", "#D4A843")) +
  theme_minimal()
```

---

## Tissue Partition Coefficients

$Kp$ values can be predicted from drug physicochemical properties:

### Tissue Composition Models

Tissues are composed of water, neutral lipids, and phospholipids. Drug partitioning depends on:

- **Lipophilicity** (logP, logD)
- **pKa** (ionization)
- **Protein binding** in plasma and tissues

```{r}
# Example: predicted Kp values for a hypothetical drug
Kp_predictions <- data.frame(
  Tissue = c("Lung", "Brain", "Heart", "Liver", "Kidney",
             "Muscle", "Adipose", "Skin", "Bone", "Gut"),
  Kp = c(1.5, 2.8, 1.2, 3.5, 2.0, 0.8, 10.0, 1.5, 0.5, 2.0)
)

ggplot(Kp_predictions, aes(x = reorder(Tissue, Kp), y = Kp)) +
  geom_col(fill = "#0B6E6B", alpha = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
  coord_flip() +
  labs(title = "Predicted Tissue:Plasma Partition Coefficients (Kp)",
       x = "", y = "Kp (Tissue:Plasma)") +
  theme_minimal()
```

---

## Scaling from In Vitro to In Vivo

### IVIVE (In Vitro to In Vivo Extrapolation)

$$CL_{int,in\ vivo} = CL_{int,in\ vitro} \cdot \text{Scaling Factor}$$

Scaling factors depend on:
- Microsomal protein per gram of liver
- Liver weight
- Hepatocyte density

```{r}
# IVIVE example: Hepatocyte CL_int → In vivo CL
CL_int_hepatocytes <- 5  # µL/min/million cells
hepatocytes_per_g_liver <- 120e6  # cells/g liver
liver_weight_g <- 1500  # g

# Scale to whole liver
CL_int_liver <- CL_int_hepatocytes * hepatocytes_per_g_liver * liver_weight_g
# Convert µL/min to L/h
CL_int_liver_Lh <- CL_int_liver * 60 / 1e6

# Well-stirred liver model
Q_h <- 1500 / 60  # L/h, hepatic blood flow
fu <- 0.3  # fraction unbound in plasma

CL_hepatic <- (Q_h * fu * CL_int_liver_Lh) / (Q_h + fu * CL_int_liver_Lh)

cat("In vitro CL_int (hepatocytes):", CL_int_hepatocytes, "µL/min/10^6 cells\n")
cat("Scaled CL_int (whole liver):", round(CL_int_liver_Lh, 1), "L/h\n")
cat("Predicted in vivo CL (well-stirred):", round(CL_hepatic, 1), "L/h\n")
```

---

## When to Use PBPK vs Compartmental Models

| Criterion | Classical Compartmental | PBPK |
|-----------|------------------------|------|
| **Data requirement** | Concentration-time data only | Rich physiological data |
| **Mechanistic insight** | Limited | High |
| **Extrapolation** | Within observed range only | To new populations, routes |
| **DDI prediction** | Requires clinical DDI data | Can predict from in vitro |
| **Regulatory acceptance** | Standard for NCA/PopPK | Growing for special populations |
| **Complexity** | Low–moderate | High |

---

## IVIVE: From In Vitro to In Vivo

PBPK models need clearance parameters for each organ. **In Vitro-In Vivo Extrapolation (IVIVE)** provides these from laboratory measurements — translating microsome or hepatocyte assays into whole-body predictions.

### The IVIVE Scaling Chain

The standard 4-step pipeline:

**Step 1 — Intrinsic clearance from in vitro kinetics:**

$$CL_{int} = \frac{V_{max}}{K_m}$$

Measured in microsomes or hepatocytes at sub-saturating concentrations ($C \ll K_m$).

**Step 2 — Scale to whole-organ via the well-stirred model:**

$$CL_{hepatic} = \frac{Q_{hep} \cdot f_u \cdot CL_{int}}{Q_{hep} + f_u \cdot CL_{int}}$$

where $Q_{hep}$ is hepatic blood flow (~90 L/h in a 70 kg adult) and $f_u$ is the unbound fraction in plasma.

**Step 3 — Adjust for protein binding:**

$$CL_{effective} = CL_{hepatic} \cdot f_u$$

Only unbound drug is available for metabolism.

**Step 4 — Feed into PBPK mass-balance ODEs:**

$$V_T \cdot \frac{dC_T}{dt} = Q_T \cdot \left(C_{art} - \frac{C_T}{Kp_T}\right) - CL_{met,T} \cdot C_{u,T}$$

### Flow-Limited vs Capacity-Limited Clearance

| Condition | $CL_{hepatic} \approx$ | Implication |
|---|---|---|
| $f_u \cdot CL_{int} \gg Q_{hep}$ | $Q_{hep}$ (flow-limited) | $CL_{int}$ errors barely matter — clearance is determined by blood delivery |
| $f_u \cdot CL_{int} \ll Q_{hep}$ | $f_u \cdot CL_{int}$ (capacity-limited) | $CL_{int}$ errors propagate linearly — accurate in vitro data is critical |

::: {.key-box}
### Know Your Regime

Before performing IVIVE, check which regime your drug is in. For high-extraction drugs (flow-limited, $E \approx 1$), $CL_{hep} \approx Q_{hep}$ regardless of $CL_{int}$ precision. For low-extraction drugs (capacity-limited), every error in $V_{max}$ or $K_m$ propagates to the in vivo prediction.
:::

### Allometric Scaling for Interspecies Extrapolation

$$Y = a \cdot \left(\frac{W}{W_{ref}}\right)^b$$

Default exponents: $\beta \approx 0.75$ for clearance, $\beta \approx 1.0$ for volume of distribution. These emerge from allometric principles (Kleiber's law for metabolic rate) and are the starting point for first-in-human dose prediction.

---

::: {.exercise-box}
### Exercises

1. Explain why PBPK models are preferred for first-in-human dose prediction from preclinical data.
2. Build a minimal PBPK model with 4 tissue compartments. Simulate and compare tissue concentrations.
3. How does increasing the adipose $Kp$ affect the terminal half-life? Why?
4. Use IVIVE to predict hepatic clearance from microsomal stability data: $CL_{int}$ = 50 µL/min/mg protein, MPPGL = 40 mg/g, liver weight = 1500 g.
:::

::: {.mistakes-box}
### Common Mistakes

- **Using PBPK when a simple compartmental model suffices**: PBPK requires many parameters — if you don't have them, don't guess
- **Ignoring blood:plasma ratio**: PBPK models track blood concentrations, not plasma
- **Forgetting permeability-limited distribution**: Not all tissues are perfusion-limited
- **Over-reliance on predicted Kp values**: Validate with observed data whenever possible
:::

::: {.key-box}
### Key Takeaway

PBPK models are the most mechanistically detailed PK models, incorporating anatomy, physiology, and drug-specific properties. They excel at extrapolation — predicting PK in untested populations, routes, and DDI scenarios. Use them when mechanistic understanding and extrapolation are needed; use simpler models when they suffice.
:::

