---
title: "Bayesian PK/PD and Advanced Modeling"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Understand Bayesian principles applied to PK/PD estimation
- Implement Bayesian parameter estimation with prior information
- Interpret MCMC output and posterior distributions in PK/PD context
- Choose appropriate priors for PK/PD parameters
- Compare Bayesian and frequentist (maximum likelihood) approaches
- Use Stan/brms for Bayesian PK/PD modeling in R

---

## Why Bayesian PK/PD?

Bayesian methods are increasingly used in pharmacometrics because they:

1. **Incorporate prior knowledge**: Leverage previous studies, preclinical data, or physiological constraints
2. **Handle sparse data naturally**: Individual Bayesian posteriors can be stable with 1–2 samples
3. **Quantify uncertainty fully**: Posterior distributions show all parameter uncertainty
4. **Enable TDM**: Bayesian forecasting for individualized dosing
5. **Avoid asymptotic approximations**: Exact finite-sample inference

---

## Bayes' Theorem in PK/PD

$$p(\theta \mid y) = \frac{p(y \mid \theta) \cdot p(\theta)}{p(y)} \propto p(y \mid \theta) \cdot p(\theta)$$

where:
- $p(\theta)$: Prior distribution (what we believe about PK parameters before seeing data)
- $p(y \mid \theta)$: Likelihood (probability of observed data given parameters)
- $p(\theta \mid y)$: Posterior distribution (updated beliefs after seeing data)

---

## A Bayesian PK Example: One-Compartment IV

```{r}
library(ggplot2)
library(dplyr)

# Simulate PK data for one subject
set.seed(42)
dose <- 500  # mg
true_CL <- 5.0  # L/h
true_V  <- 45   # L
true_ke <- true_CL / true_V

sample_times <- c(0.25, 0.5, 1, 2, 4, 8, 12, 24)
obs_data <- data.frame(
  time = sample_times,
  conc = (dose / true_V) * exp(-true_ke * sample_times)
)
obs_data$conc <- obs_data$conc * exp(rnorm(nrow(obs_data), 0, 0.15))

# Visualize the data
ggplot(obs_data, aes(time, conc)) +
  geom_point(size = 3, color = "#C44536") +
  scale_y_log10() +
  labs(title = "Observed PK Data (Single Subject)",
       x = "Time (hours)", y = "Concentration (mg/L, log scale)") +
  theme_minimal()
```

### Prior Distributions

For PK parameters, we typically use log-normal priors because parameters are positive:

$$CL \sim \text{LogNormal}(\log(4), 0.5)$$
$$V \sim \text{LogNormal}(\log(40), 0.3)$$
$$\sigma \sim \text{HalfNormal}(0, 2)$$

```{r}
# Visualize priors
CL_seq <- seq(0.5, 20, length.out = 200)
V_seq  <- seq(10, 100, length.out = 200)

prior_CL <- dlnorm(CL_seq, log(4), 0.5)
prior_V  <- dlnorm(V_seq, log(40), 0.3)

df_priors <- bind_rows(
  data.frame(value = CL_seq, density = prior_CL, Parameter = "CL (L/h)"),
  data.frame(value = V_seq,  density = prior_V,  Parameter = "V (L)")
)

ggplot(df_priors, aes(value, density, fill = Parameter)) +
  geom_area(alpha = 0.5) +
  geom_vline(data = data.frame(Parameter = c("CL (L/h)", "V (L)"),
                                true = c(true_CL, true_V)),
             aes(xintercept = true), color = "#C44536", linewidth = 1, linetype = "dashed") +
  annotate("text", x = c(true_CL, true_V), y = max(c(prior_CL, prior_V)) * 0.9,
           label = "True value", color = "#C44536", hjust = -0.2, size = 3) +
  facet_wrap(~Parameter, scales = "free") +
  labs(title = "Prior Distributions for PK Parameters",
       x = "Parameter Value", y = "Density") +
  scale_fill_manual(values = c("#0B6E6B", "#D4A843")) +
  theme_minimal()
```

---

## MCMC Sampling for PK Parameters

### Using Stan via cmdstanr

```{r}
#| eval: false
library(cmdstanr)

# Stan model for one-compartment IV bolus
stan_code <- "
data {
  int<lower=0> N;
  array[N] real time;
  array[N] real conc;
  real dose;
}
parameters {
  real<lower=0> CL;
  real<lower=0> V;
  real<lower=0> sigma;
}
transformed parameters {
  real ke = CL / V;
  array[N] real conc_pred;
  for (i in 1:N) {
    conc_pred[i] = (dose / V) * exp(-ke * time[i]);
  }
}
model {
  // Priors
  CL ~ lognormal(log(4), 0.5);
  V  ~ lognormal(log(40), 0.3);
  sigma ~ normal(0, 2);

  // Likelihood
  conc ~ lognormal(log(conc_pred), sigma);
}
"

# Compile and sample
# mod <- cmdstan_model(write_stan_file(stan_code))
# fit <- mod$sample(
#   data = list(N = nrow(obs_data), time = obs_data$time,
#               conc = obs_data$conc, dose = dose),
#   seed = 123, chains = 4, parallel_chains = 4,
#   iter_warmup = 1000, iter_sampling = 2000
# )
# fit$summary()
```

---

## Simulating MCMC Results

```{r}
# Simulate posterior samples (normally we get these from MCMC)
set.seed(123)
n_post <- 4000

posterior_samples <- data.frame(
  CL = rlnorm(n_post, log(true_CL), 0.15),
  V  = rlnorm(n_post, log(true_V), 0.10),
  sigma = abs(rnorm(n_post, 0.2, 0.05))
) %>%
  mutate(ke = CL / V)

# Posterior summary
post_summary <- posterior_samples %>%
  summarise(
    across(everything(),
           list(mean = ~mean(.x), sd = ~sd(.x),
                q025 = ~quantile(.x, 0.025), q975 = ~quantile(.x, 0.975)))
  )

knitr::kable(
  data.frame(
    Parameter = c("CL (L/h)", "V (L)", "ke (1/h)", "sigma"),
    Mean = round(c(mean(posterior_samples$CL), mean(posterior_samples$V),
                   mean(posterior_samples$ke), mean(posterior_samples$sigma)), 2),
    SD = round(c(sd(posterior_samples$CL), sd(posterior_samples$V),
                  sd(posterior_samples$ke), sd(posterior_samples$sigma)), 2),
    `2.5%` = round(c(quantile(posterior_samples$CL, 0.025),
                      quantile(posterior_samples$V, 0.025),
                      quantile(posterior_samples$ke, 0.025),
                      quantile(posterior_samples$sigma, 0.025)), 2),
    `97.5%` = round(c(quantile(posterior_samples$CL, 0.975),
                       quantile(posterior_samples$V, 0.975),
                       quantile(posterior_samples$ke, 0.975),
                       quantile(posterior_samples$sigma, 0.975)), 2)
  ),
  caption = "Posterior Summary (Simulated MCMC)"
)
```

---

## Visualizing the Posterior

```{r}
# Posterior distributions
df_post <- posterior_samples %>%
  tidyr::pivot_longer(c(CL, V, ke), names_to = "Parameter", values_to = "Value")

ggplot(df_post, aes(Value, after_stat(density), fill = Parameter)) +
  geom_histogram(alpha = 0.5, bins = 50, position = "identity") +
  geom_vline(data = data.frame(Parameter = c("CL", "V", "ke"),
                                true = c(true_CL, true_V, true_ke)),
             aes(xintercept = true), color = "#C44536", linewidth = 1, linetype = "dashed") +
  facet_wrap(~Parameter, scales = "free") +
  labs(title = "Posterior Distributions of PK Parameters",
       subtitle = "Dashed line = true value used to simulate data",
       x = "Parameter Value", y = "Density") +
  scale_fill_manual(values = c("#0B6E6B", "#C44536", "#D4A843")) +
  theme_minimal()
```

---

## Posterior Predictive Check

```{r}
# Generate posterior predictive samples
n_ppc <- 200
ppc_times <- seq(0, 24, by = 0.2)
ppc_idx <- sample(n_post, n_ppc)

ppc_df <- lapply(seq_along(ppc_idx), function(i) {
  idx <- ppc_idx[i]
  data.frame(
    time = ppc_times,
    conc = (dose / posterior_samples$V[idx]) *
           exp(-posterior_samples$ke[idx] * ppc_times),
    draw = i
  )
}) %>% bind_rows()

# Compute prediction intervals
ppc_summary <- ppc_df %>%
  group_by(time) %>%
  summarise(
    p05 = quantile(conc, 0.05),
    p50 = quantile(conc, 0.50),
    p95 = quantile(conc, 0.95),
    .groups = "drop"
  )

ggplot() +
  geom_ribbon(data = ppc_summary, aes(time, ymin = p05, ymax = p95),
              fill = "#0B6E6B", alpha = 0.2) +
  geom_line(data = ppc_summary, aes(time, p50),
            color = "#0B6E6B", linewidth = 1) +
  geom_point(data = obs_data, aes(time, conc),
             color = "#C44536", size = 3) +
  scale_y_log10() +
  labs(title = "Posterior Predictive Check",
       subtitle = "Shaded: 90% prediction interval; Dashed: median; Points: observed",
       x = "Time (hours)", y = "Concentration (mg/L, log scale)") +
  theme_minimal()
```

---

## Bayesian vs Frequentist: Key Differences

| Aspect | Frequentist (MLE) | Bayesian |
|--------|------------------|----------|
| **Parameter interpretation** | Fixed, unknown constant | Random variable with distribution |
| **Uncertainty** | Confidence intervals (asymptotic) | Credible intervals (direct probability) |
| **Prior information** | Not used | Formally incorporated |
| **Sparse data** | May be unstable | Priors stabilize estimates |
| **Computation** | Optimization (fast) | MCMC sampling (slower) |
| **P-value** | Yes | No (posterior probabilities instead) |
| **TDM application** | MAP estimation | Full posterior / MAP |

---

## MCMC Convergence Diagnostics

A non-converged MCMC posterior is meaningless — regardless of how plausible the estimates look. Always check these diagnostics before interpreting results:

### The Convergence Checklist

| Diagnostic | Threshold | Action If Failed |
|---|---|---|
| **$\hat{R}$ (Gelman-Rubin)** | $\hat{R} < 1.1$ for all parameters | Run more iterations; reparameterize (non-centered) |
| **Effective Sample Size (ESS)** | ESS > 400 per chain (bulk & tail) | Thin less aggressively; increase total iterations |
| **Acceptance rate (MH)** | 20-50% | < 20%: decrease step size; > 50%: increase step size |
| **Trace plots** | Stationary, well-mixed "hairy caterpillar" | Drift/trend: discard more burn-in; check model specification |
| **Divergent transitions (HMC)** | 0 (or < 1%) | Use non-centered parameterization; increase `adapt_delta` |
| **Pairs plots** | No funnel shapes or extreme correlations | Funnel at small $\omega$: non-centered $\eta \sim N(0,1)$, $\theta = \exp(\mu + \omega \cdot \eta)$ |

### The Funnel Problem

In hierarchical models, when the between-subject SD $\omega$ is near zero, the posterior develops a "funnel" shape — very narrow at small $\omega$, wide at large $\omega$. Standard samplers struggle with this geometry.

**The fix: Non-centered parameterization.** Instead of sampling $\theta_i \sim \text{LogNormal}(\mu, \omega)$, sample:

$$\eta_i \sim N(0, 1), \quad \theta_i = \exp(\mu + \omega \cdot \eta_i)$$

This removes the dependence of the geometry on $\omega$ and dramatically improves sampling.

```{r}
# Visualize the difference between centered and non-centered parameterization
set.seed(123)
n_samples <- 2000

# Centered: direct sampling of theta (creates funnel)
mu <- 1; omega <- 0.1
theta_centered <- rlnorm(n_samples, mu, omega)

# Non-centered: sample eta ~ N(0,1), then transform
eta <- rnorm(n_samples, 0, 1)
theta_noncentered <- exp(mu + omega * eta)

# Both give the same marginal distribution
# But for MCMC, the non-centered form is far more efficient

df_param <- data.frame(
  theta = c(theta_centered, theta_noncentered),
  Method = rep(c("Centered", "Non-centered"), each = n_samples)
)

ggplot(df_param, aes(theta, after_stat(density), fill = Method)) +
  geom_histogram(alpha = 0.5, bins = 50, position = "identity") +
  labs(title = "Centered vs Non-Centered Parameterization",
       subtitle = "Both give identical marginal distributions, but non-centered is MCMC-friendly",
       x = expression(theta), y = "Density") +
  scale_fill_manual(values = c("#C44536", "#0B6E6B")) +
  theme_minimal()
```

::: {.key-box}
### Default Rules of Thumb for MCMC

- Always use **log-scale parameterization** for strictly-positive PK parameters ($CL$, $V$, $k_a$, $k_e$): sample $\log(CL)$ not $CL$
- For hierarchical (population) models: **always start with non-centered parameterization**
- Target acceptance rate: **~20-50%** for random-walk MH; **> 80%** for HMC/NUTS (`adapt_delta = 0.9` or higher)
- Burn-in: discard at minimum the **first 50%** of warmup iterations; inspect trace plots to confirm
- **Never use a single chain** — run at least 4 chains from different starting values and compare $\hat{R}$
:::

---

## Choosing Priors

### Types of Priors

| Prior Type | Purpose | Example |
|-----------|---------|---------|
| **Informative** | Strong prior knowledge | CL ~ LogNormal(log(5), 0.1) |
| **Weakly informative** | Regularization without strong influence | CL ~ LogNormal(log(5), 1.0) |
| **Non-informative** | Let data dominate | CL ~ Uniform(0, 1000) |
| **Physiological** | Based on known constraints | $V_{plasma}$ ≈ 3 L |

```{r}
# Compare informative vs weakly informative priors
CL_seq <- seq(0.1, 20, length.out = 200)

df_prior_comp <- data.frame(
  CL = rep(CL_seq, 2),
  Density = c(dlnorm(CL_seq, log(5), 0.15),   # Informative
              dlnorm(CL_seq, log(5), 1.0)),    # Weakly informative
  Prior = rep(c("Informative (σ=0.15)", "Weakly Informative (σ=1.0)"),
              each = length(CL_seq))
)

ggplot(df_prior_comp, aes(CL, Density, color = Prior)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Informative vs Weakly Informative Priors",
       x = "Clearance (L/h)", y = "Density") +
  scale_color_manual(values = c("#C44536", "#0B6E6B")) +
  theme_minimal()
```

---

::: {.exercise-box}
### Exercises

1. Run a Bayesian PK analysis for one-compartment oral data. Use Stan or brms to estimate $k_a$, $CL$, and $V$.
2. Compare posterior means and 95% credible intervals to MLE estimates. Are they different? When and why?
3. Perform a prior sensitivity analysis: how do different prior choices affect the posterior for CL?
4. Generate a posterior predictive check and interpret whether the model adequately captures the data.
:::

::: {.mistakes-box}
### Common Mistakes

- **Using overly informative priors**: May dominate the data and bias results
- **Not checking MCMC convergence**: R-hat > 1.01 indicates problems
- **Confusing credible intervals with confidence intervals**: They have different interpretations
- **Ignoring prior sensitivity analysis**: Always check that conclusions aren't driven by prior choice
- **Using flat priors on unbounded parameters**: Can cause improper posteriors
:::

::: {.key-box}
### Key Takeaway

Bayesian methods provide a natural framework for PK/PD modeling, especially when prior information is available or data are sparse. The posterior distribution fully quantifies parameter uncertainty, enabling robust TDM and dose individualization. Always check MCMC convergence and prior sensitivity.
:::

