---
title: "Sensitivity Analysis in PK/PD Modeling"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Understand why sensitivity analysis is essential before and after PK/PD model fitting
- Perform local sensitivity analysis using finite-difference methods
- Conduct global sensitivity analysis with Sobol indices and PRCC
- Triage which parameters to estimate vs fix based on sensitivity results
- Interpret sensitivity indices to guide model simplification
- Apply the `sensitivity` package to PK/PD models in R

---

## Why Sensitivity Analysis?

Every PK/PD model has parameters — some strongly influence the model output, others barely matter. **Sensitivity analysis** quantifies this influence. It answers two fundamental questions:

1. **Before estimation**: Which parameters can I reliably estimate from my data, and which should I fix at literature values?
2. **After model building**: Which parameters drive the predictions I care about (AUC, $C_{max}$, effect)? Where should I invest effort in getting better estimates?

::: {.key-box}
### The Sensitivity Triage Principle

> Run sensitivity analysis *before* parameter estimation to triage which parameters to fit vs fix. It is cheaper to discover an unidentifiable parameter through sensitivity analysis than to discover it through failed estimation.
:::

---

## Two Classes of Sensitivity Analysis

| | Local Sensitivity | Global Sensitivity |
|---|---|---|
| **What it measures** | Derivative $\partial Y/\partial p$ at one point | Variance decomposition across full parameter space |
| **Cost** | Low (N+1 model evaluations for N parameters) | High ($\geq 1000$ per parameter) |
| **Captures interactions?** | No | Yes |
| **Valid when** | Near-linear, near the nominal parameter values | Anywhere in parameter space |
| **Analogy** | A flashlight — illuminates one spot | Daylight — surveys the whole landscape |
| **Use for** | Quick triage before estimation | Final parameter prioritization, model simplification |

---

## Local Sensitivity Analysis

### Normalized Sensitivity Coefficient

The local sensitivity of output $Y$ to parameter $p$ is the normalized partial derivative:

$$S_p = \frac{\partial Y}{\partial p} \cdot \frac{p}{Y}$$

The normalization (multiplying by $p/Y$) makes sensitivities comparable across parameters with different units. A value of $S_p = 0.5$ means a 1% change in $p$ produces approximately a 0.5% change in $Y$.

### Finite-Difference Approximation

We approximate the derivative numerically:

$$S_p \approx \frac{Y(p + \Delta p) - Y(p - \Delta p)}{2\Delta p} \cdot \frac{p}{Y(p)}$$

where $\Delta p$ is a small perturbation (typically 1% of the nominal value: $\Delta p = 0.01 \cdot p$).

```{r}
library(rxode2)
library(ggplot2)
library(dplyr)
library(tidyr)

# One-compartment IV model for sensitivity demonstration
mod_1cmt <- rxode2({
  d/dt(central) <- -ke * central
  conc <- central / V
})

# Nominal parameters
params_nom <- c(ke = 0.18, V = 22)
dose <- 250  # mg
times <- seq(0, 24, by = 0.1)

# Function: simulate and return AUC and Cmax
simulate_pk <- function(pars) {
  event <- et(time = 0, amt = dose)
  sim <- rxSolve(mod_1cmt, pars, event %>% et(times))
  df <- as.data.frame(sim)
  c(
    AUC  = sum(diff(df$time) * (head(df$conc, -1) + tail(df$conc, -1)) / 2),
    Cmax = max(df$conc)
  )
}

# Compute local sensitivities via central finite difference
compute_local_sensitivity <- function(params_nom, delta_frac = 0.01) {
  Y_nom <- simulate_pk(params_nom)
  
  sens <- list()
  for (i in seq_along(params_nom)) {
    pname <- names(params_nom)[i]
    p_nom <- params_nom[i]
    delta <- delta_frac * p_nom
    
    p_plus <- params_nom; p_plus[i] <- p_nom + delta
    p_minus <- params_nom; p_minus[i] <- p_nom - delta
    
    Y_plus  <- simulate_pk(p_plus)
    Y_minus <- simulate_pk(p_minus)
    
    for (outname in names(Y_nom)) {
      dY_dp <- (Y_plus[outname] - Y_minus[outname]) / (2 * delta)
      Sp <- dY_dp * p_nom / Y_nom[outname]
      sens[[length(sens) + 1]] <- data.frame(
        Parameter = pname, Output = outname, Sensitivity = Sp
      )
    }
  }
  bind_rows(sens)
}

local_sens <- compute_local_sensitivity(params_nom)

ggplot(local_sens, aes(Parameter, Sensitivity, fill = Output)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  labs(
    title = "Local Sensitivity Analysis: One-Compartment IV Model",
    subtitle = paste0("Normalized sensitivity coefficients (Sp) at ke = ", 
                      params_nom["ke"], ", V = ", params_nom["V"]),
    x = "Parameter", y = expression(S[p] ~ " (normalized sensitivity)")
  ) +
  scale_fill_manual(values = c("AUC" = "#0B6E6B", "Cmax" = "#C44536")) +
  theme_minimal()
```

::: {.interpretation-box}
### Interpreting Local Sensitivities

- **ke → AUC**: $S_{ke} \approx -1$ — a 1% increase in $k_e$ reduces AUC by ~1%. AUC is highly sensitive to $k_e$.
- **ke → Cmax**: $S_{ke} \approx 0$ — $k_e$ doesn't affect $C_{max}$ (which occurs at $t=0$ for IV bolus).
- **V → AUC**: $S_V \approx -1$ — AUC inversely proportional to $V$ (AUC = Dose/CL = Dose/(ke·V)).
- **V → Cmax**: $S_V \approx -1$ — $C_{max}$ = Dose/V, so directly sensitive.

Parameters with $|S_p| < 0.1$ are candidates for fixing at literature values rather than estimating.
:::

---

## Global Sensitivity Analysis

Local sensitivity only illuminates one spot. If the model is nonlinear or parameters interact, global sensitivity is needed. Two main approaches:

### Sobol Indices (Variance-Based)

Decompose the total variance of output $Y$ into contributions from individual parameters and their interactions:

$$V(Y) = \sum_i V_i + \sum_{i<j} V_{ij} + \sum_{i<j<k} V_{ijk} + \dots$$

- **First-order index** $S_i = V_i / V(Y)$: fraction of variance due to parameter $i$ alone
- **Total-order index** $S_{Ti} = 1 - V_{\sim i} / V(Y)$: fraction due to $i$ and all its interactions
- $S_{Ti} - S_i$ quantifies the interaction contribution

### Partial Rank Correlation Coefficient (PRCC)

Rank-based partial correlation between each parameter and the output — useful when relationships are monotonic but nonlinear.

```{r}
# Global sensitivity analysis for a two-compartment model

# Two-compartment IV model
mod_2cmt <- rxode2({
  d/dt(central)    <- -(ke + k12) * central + k21 * peripheral
  d/dt(peripheral) <- k12 * central - k21 * peripheral
  conc <- central / V1
})

# Function to compute AUC from parameters
compute_auc_2cmt <- function(params_vec, dose = 100) {
  p <- setNames(as.list(params_vec), c("ke", "V1", "k12", "k21"))
  event <- et(time = 0, amt = dose)
  sim <- tryCatch({
    rxSolve(mod_2cmt, unlist(p), event %>% et(seq(0, 48, by = 0.5)))
  }, error = function(e) NULL)
  if (is.null(sim)) return(NA)
  df <- as.data.frame(sim)
  sum(diff(df$time) * (head(df$conc, -1) + tail(df$conc, -1)) / 2)
}

# Parameter ranges (nominal ± 50%)
params_nom_2cmt <- c(ke = 0.22, V1 = 18, k12 = 0.65, k21 = 0.42)
n_params <- length(params_nom_2cmt)

# Generate parameter samples using Sobol sequences
set.seed(789)
N_sobol <- 100  # Demonstration size; use N >= 1000 for production

# Sobol sequence requires two matrices
X1 <- matrix(runif(N_sobol * n_params), ncol = n_params)
X2 <- matrix(runif(N_sobol * n_params), ncol = n_params)

# Scale to parameter ranges (±50%)
scale_to_range <- function(x, nom) {
  nom * (0.5 + x)  # uniform on [0.5*nom, 1.5*nom]
}

# Build the design matrix
build_design <- function(X1, X2, params_nom) {
  n <- nrow(X1)
  p <- ncol(X1)
  # Need N*(2p+2) rows for Sobol-Jansen estimator
  des <- matrix(NA, nrow = n * (2 * p + 2), ncol = p)
  idx <- 1
  for (i in 1:n) {
    des[idx, ] <- scale_to_range(X1[i, ], params_nom)
    idx <- idx + 1
  }
  for (i in 1:n) {
    des[idx, ] <- scale_to_range(X2[i, ], params_nom)
    idx <- idx + 1
  }
  for (j in 1:p) {
    for (i in 1:n) {
      row <- X1[i, ]
      row[j] <- X2[i, j]
      des[idx, ] <- scale_to_range(row, params_nom)
      idx <- idx + 1
    }
  }
  for (j in 1:p) {
    for (i in 1:n) {
      row <- X2[i, ]
      row[j] <- X1[i, j]
      des[idx, ] <- scale_to_range(row, params_nom)
      idx <- idx + 1
    }
  }
  na.omit(des)
}

design <- build_design(X1, X2, params_nom_2cmt)

# Run model for each design point (this takes a moment)
Y_auc <- apply(design, 1, compute_auc_2cmt)
Y_auc[is.na(Y_auc)] <- mean(Y_auc, na.rm = TRUE)  # impute failures

# Compute Sobol indices using the Jansen estimator
n <- N_sobol
p <- n_params

# Split Y into the design blocks
YA <- Y_auc[1:n]
YB <- Y_auc[(n+1):(2*n)]

# First-order and total-effect Sobol indices (Jansen 1999 estimator)
sobol_results <- data.frame(
  Parameter = names(params_nom_2cmt),
  FirstOrder = NA,
  TotalEffect = NA
)

VarY <- var(YA)
for (j in 1:p) {
  YCj <- Y_auc[(2*n + (j-1)*n + 1):(2*n + j*n)]
  YCj2 <- Y_auc[(2*n + p*n + (j-1)*n + 1):(2*n + p*n + j*n)]
  
  # Jansen estimator for total effect
  STi <- mean((YA - YCj2)^2) / (2 * VarY)
  
  # First-order estimator
  Si <- 1 - mean((YB - YCj)^2) / (2 * VarY)
  
  sobol_results$FirstOrder[j] <- pmax(0, pmin(1, Si))
  sobol_results$TotalEffect[j] <- pmax(0, pmin(1, STi))
}

# Plot Sobol indices
sobol_long <- sobol_results %>%
  pivot_longer(cols = c(FirstOrder, TotalEffect),
               names_to = "Index", values_to = "Value")

ggplot(sobol_long, aes(Parameter, Value, fill = Index)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "gray50") +
  annotate("text", x = 4.5, y = 0.07, label = "5% threshold", 
           size = 3, color = "gray50", hjust = 1) +
  labs(
    title = "Global Sensitivity Analysis: Two-Compartment Model AUC",
    subtitle = "Sobol indices (first-order and total-effect) with N = 100 (demo)",
    x = "Parameter", y = "Sensitivity Index"
  ) +
  scale_fill_manual(
    values = c("FirstOrder" = "#0B6E6B", "TotalEffect" = "#C44536"),
    labels = c("First-order (Si)", "Total-effect (STi)")
  ) +
  theme_minimal()
```

::: {.interpretation-box}
### Interpreting Sobol Indices

- **$S_i \approx S_{Ti}$**: Parameter $i$ acts independently (no interactions).
- **$S_{Ti} \gg S_i$**: Parameter $i$ has strong interactions with other parameters — you cannot understand its effect in isolation.
- **$S_{Ti} < 0.05$**: Parameter $i$ explains less than 5% of output variance. Strong candidate for fixing at a nominal value.
- **$S_{Ti}$ close to 1**: This parameter dominates the output — prioritize getting a good estimate.
:::

---

## PRCC: Rank-Based Global Sensitivity

When relationships are monotonic but strongly nonlinear, PRCC is a robust alternative to Sobol indices.

```{r}
# PRCC analysis for the same two-compartment model
set.seed(456)
N_prcc <- 200

# Latin Hypercube Sampling
lhs_samples <- matrix(runif(N_prcc * n_params), ncol = n_params)

# Scale and compute outputs
params_lhs <- t(apply(lhs_samples, 1, function(x) scale_to_range(x, params_nom_2cmt)))
Y_lhs <- apply(params_lhs, 1, compute_auc_2cmt)
Y_lhs[is.na(Y_lhs)] <- mean(Y_lhs, na.rm = TRUE)

# Compute PRCC
compute_prcc <- function(X, Y) {
  # Rank-transform both X and Y
  R_X <- apply(X, 2, rank)
  R_Y <- rank(Y)
  
  # Partial correlation of ranked variables
  prcc_vals <- numeric(ncol(X))
  for (j in 1:ncol(X)) {
    # Regress R_X[,j] on all other R_X columns, and R_Y on all other R_X columns
    X_other <- R_X[, -j, drop = FALSE]
    
    # Residuals for parameter j
    fit_x <- lm(R_X[, j] ~ X_other)
    res_x <- residuals(fit_x)
    
    # Residuals for output
    fit_y <- lm(R_Y ~ X_other)
    res_y <- residuals(fit_y)
    
    prcc_vals[j] <- cor(res_x, res_y)
  }
  names(prcc_vals) <- colnames(X)
  prcc_vals
}

prcc_results <- compute_prcc(params_lhs, Y_lhs)

prcc_df <- data.frame(
  Parameter = names(prcc_results),
  PRCC = prcc_results
)

# Determine significance via bootstrap or approximate threshold
# For N=500, |PRCC| > 0.09 is approximately significant at p < 0.05

ggplot(prcc_df, aes(reorder(Parameter, -abs(PRCC)), PRCC, fill = PRCC > 0)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = c(-0.09, 0.09), linetype = "dashed", color = "gray50") +
  labs(
    title = "PRCC Sensitivity Analysis: Two-Compartment Model AUC",
    subtitle = "Partial Rank Correlation Coefficients (N = 200 LHS samples)",
    x = "Parameter", y = "PRCC"
  ) +
  scale_fill_manual(values = c("TRUE" = "#C44536", "FALSE" = "#0B6E6B"), guide = "none") +
  theme_minimal()
```

---

## Sensitivity Analysis for PK/PD Link Models

Sensitivity analysis becomes especially valuable in linked PK/PD models where the parameter count grows rapidly. Consider an effect-compartment model:

```{r}
# Effect-compartment PK/PD model
mod_pkpd_link <- rxode2({
  d/dt(central) <- -ke * central
  d/dt(effect)  <- ke0 * (central/V - effect)
  conc <- central / V
  Ce   <- effect
})

# Parameters: ke, V, ke0, Emax, EC50
params_pkpd <- c(ke = 0.18, V = 22, ke0 = 0.35, Emax = -65, EC50 = 8)

compute_effect_auc <- function(params_vec, dose = 100) {
  p <- setNames(as.list(params_vec), c("ke", "V", "ke0", "Emax", "EC50"))
  event <- et(time = 0, amt = dose)
  sim <- tryCatch({
    rxSolve(mod_pkpd_link, unlist(p), event %>% et(seq(0, 48, by = 0.1)))
  }, error = function(e) NULL)
  if (is.null(sim)) return(NA)
  df <- as.data.frame(sim)
  df$Effect <- p$Emax * df$Ce / (p$EC50 + df$Ce)
  # Area under the effect curve (absolute value)
  sum(diff(df$time) * abs(head(df$Effect, -1) + tail(df$Effect, -1)) / 2)
}

# --- Helper function for PK/PD local sensitivity ---
local_sensitivity_pkpd <- function(params_nom, model_fn, delta_frac = 0.01) {
  Y_nom <- model_fn(params_nom)
  
  sens <- list()
  for (i in seq_along(params_nom)) {
    pname <- names(params_nom)[i]
    p_nom <- params_nom[i]
    delta <- delta_frac * abs(p_nom)
    if (delta == 0) delta <- delta_frac
    
    p_plus <- params_nom; p_plus[i] <- p_nom + delta
    p_minus <- params_nom; p_minus[i] <- p_nom - delta
    
    Y_plus  <- model_fn(p_plus)
    Y_minus <- model_fn(p_minus)
    
    dY_dp <- (Y_plus - Y_minus) / (2 * delta)
    Sp <- dY_dp * p_nom / Y_nom
    
    sens[[length(sens) + 1]] <- data.frame(
      Parameter = pname, Sensitivity = Sp
    )
  }
  bind_rows(sens)
}

# Compute local sensitivity for the PK/PD model
local_pkpd <- local_sensitivity_pkpd(params_pkpd, compute_effect_auc)

ggplot(local_pkpd, aes(reorder(Parameter, -abs(Sensitivity)), Sensitivity, 
                        fill = Sensitivity > 0)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  labs(
    title = "Local Sensitivity: Effect-Compartment PK/PD Model",
    subtitle = "Sensitivity of integrated effect (AUEC) to each parameter",
    x = "Parameter", y = expression(S[p])
  ) +
  scale_fill_manual(values = c("TRUE" = "#C44536", "FALSE" = "#0B6E6B"), guide = "none") +
  theme_minimal()
```

::: {.key-box}
### When Global Sensitivity Is Essential

Switch from local to global sensitivity analysis when:

1. Parameters have known nonlinear relationships with output (e.g., $EC_{50}$ in the Hill equation)
2. Parameters interact (e.g., $k_e$ and $V$ both affect AUC through CL = ke·V)
3. The model contains thresholds or regime changes (e.g., Michaelis-Menten saturation)
4. You need to defend model simplification decisions in a regulatory context
:::

---

## Practical Workflow

```{r}
#| echo: false
workflow_sa <- data.frame(
  Step = 1:5,
  Action = c(
    "Define outputs of interest (AUC, Cmax, Effect AUC, Cmin)",
    "Run local sensitivity (finite-difference) for quick triage",
    "For parameters with |Sp| < 0.1: consider fixing at literature values",
    "For the remaining parameters: run global SA (Sobol or PRCC)",
    "Fix parameters with STi < 0.05; focus estimation on high-STi parameters"
  ),
  y = 5:1
)

ggplot(workflow_sa, aes(1, y, label = Action)) +
  geom_text(size = 3.5, color = "#0B6E6B", hjust = 0.5) +
  geom_label(aes(x = 0.3, y = y, label = paste0("Step ", Step)),
            fill = "#C44536", color = "white", fontface = "bold", size = 3.5) +
  geom_segment(aes(x = 0.9, xend = 0.9, y = 4.6, yend = 0.4),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "#C44536", linewidth = 1) +
  xlim(0, 1.7) + ylim(0, 6) +
  labs(title = "Sensitivity Analysis Workflow for PK/PD Models") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

---

## Sensitivity in Population PK

In population PK models, sensitivity extends to the random-effects parameters ($\omega^2$, $\sigma^2$):

```{r}
# Conceptual: Sensitivity of PopPK outputs to omega and sigma
# Often, the shrinkage of empirical Bayes estimates (EBEs) is highly sensitive
# to the ratio omega/sigma — when omega << sigma, individual predictions
# shrink toward the population mean and become insensitive to covariates.

pop_sens_concept <- data.frame(
  Scenario = c("ω >> σ", "ω ≈ σ", "ω << σ"),
  Shrinkage = c("Low", "Moderate", "High (>30%)"),
  IndividualEstimates = c("Reliable", "Mixed", "Unreliable"),
  CovariateDetection = c("Good power", "Adequate", "Poor — may miss true effects"),
  Action = c("Standard PopPK", "Standard PopPK", "Enrich sampling or simplify model")
)

knitr::kable(pop_sens_concept, 
  caption = "Sensitivity of PopPK inference to the ω/σ ratio")
```

---

## Common Mistakes

::: {.warning-box}
1. **Skipping sensitivity analysis entirely**: The most common and costly mistake. Discovering parameter unidentifiability during model fitting wastes time and may lead to over-interpreted results.

2. **Using only local sensitivity for nonlinear models**: Local sensitivity at one point tells you nothing about behavior across the parameter space. For Michaelis-Menten, Emax, or PBPK models, always complement with global methods.

3. **Using too few samples for Sobol indices**: $N = 100$ gives noisy estimates. Use $N \geq 1000$ for production analysis, $N \geq 5000$ for regulatory submissions.

4. **Ignoring parameter correlations in the sampling design**: Independent uniform/LHS sampling assumes no prior correlation. If parameters are known to be correlated (e.g., CL and V from PopPK), use copula-based sampling.

5. **Not distinguishing first-order from total-effect indices**: A parameter with $S_i \approx 0$ but $S_{Ti} \gg 0$ matters only through interactions — you cannot fix it unless you also fix its interaction partners.
:::

---

## Exercises

1. **Local sensitivity for oral absorption**: For a one-compartment oral model ($k_a$, $k_e$, $V$, $F$), compute local sensitivities of AUC, $C_{max}$, and $T_{max}$. Which parameter most strongly affects each output?

2. **Sobol for the Emax model**: Perform global sensitivity analysis on the Emax model $E = E_0 + E_{max} \cdot C/(EC_{50} + C)$. How do the sensitivities change as $C$ varies from far below $EC_{50}$ to far above?

3. **Sensitivity-guided model reduction**: For a three-compartment model, use Sobol indices to identify parameters explaining < 5% of AUC variance. Refit the data with a reduced two-compartment model and compare AIC.

4. **PopPK shrinkage sensitivity**: In a population PK model with sparse sampling (2 samples/subject), simulate how increasing $\omega$ relative to $\sigma$ affects the reliability of individual CL estimates (shrinkage).

5. **PRCC vs Sobol comparison**: For the Michaelis-Menten model, compute both PRCC and Sobol first-order indices. Do they agree on parameter ranking? Under what conditions would they diverge?

