---
title: "Reporting, Reproducibility, and Regulatory Thinking"
---


## Learning Objectives

By the end of this chapter you will be able to:

- Structure a PK/PD analysis report for regulatory or internal use
- Create publication-quality tables and figures for PK/PD results
- Build reproducible analysis pipelines with Quarto
- Apply common reporting standards in pharmacometrics
- Communicate uncertainty clearly to stakeholders
- Build a reusable analysis code library

---

## The PK/PD Report Structure

A standard PK/PD analysis report typically includes:

```{r}
#| echo: false
library(ggplot2)

report_sections <- data.frame(
  Section = c("1. Objectives", "2. Data Description", "3. Exploratory Analysis",
              "4. PK Analysis (NCA)", "5. PK Modeling", "6. PD Analysis",
              "7. PK/PD Modeling", "8. Covariate Analysis", "9. Simulation",
              "10. Conclusions"),
  y = 10:1
)

ggplot(report_sections, aes(1, y, label = Section)) +
  geom_text(size = 4, color = "#0B6E6B", fontface = "bold", hjust = 0) +
  xlim(1, 2) + ylim(0, 11) +
  labs(title = "Standard PK/PD Report Structure") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

---

## Creating Publication-Quality Tables

### Parameter Estimate Table

```{r}
library(dplyr)
library(knitr)

# Example: PK parameter summary table
pk_params <- data.frame(
  Parameter = c("ka (1/h)", "CL (L/h)", "V (L)", "t½ (h)",
                "ω_CL (%CV)", "ω_V (%CV)", "σ_prop (%CV)", "σ_add (mg/L)"),
  Estimate = c(1.23, 5.67, 48.2, 5.89, 28.5, 18.2, 15.3, 0.52),
  RSE_pct = c(8.5, 5.2, 6.8, 7.1, 15.3, 22.1, 12.4, 35.6),
  CI95_Lower = c(1.03, 5.10, 42.1, 5.12, 20.1, 11.8, 11.2, 0.18),
  CI95_Upper = c(1.47, 6.31, 55.2, 6.78, 39.5, 28.1, 21.0, 1.51),
  Shrinkage = c(NA, NA, NA, NA, 12.3, 8.7, NA, NA)
)

knitr::kable(pk_params,
  caption = "Final Population PK Parameter Estimates",
  digits = c(0, 2, 1, 1, 1, 1, 1, 1, 1, 1),
  col.names = c("Parameter", "Estimate", "RSE (%)", "95% CI Lower",
                "95% CI Upper", "Shrinkage (%)"),
  align = "lccccc"
)
```

---

## Creating Publication-Quality Figures

### Concentration-Time Profile (Standard PK)

```{r}
library(rxode2)

# Generate example data for figure
mod_fig <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <-  ka * depot - (CL/V) * central
  conc <- central / V
})

set.seed(42)
n_fig <- 8
fig_times <- seq(0, 24, by = 0.2)

fig_data <- lapply(1:n_fig, function(i) {
  params <- c(ka = 1.0 * exp(rnorm(1, 0, 0.3)),
              CL = 5.0 * exp(rnorm(1, 0, 0.25)),
              V  = 50 * exp(rnorm(1, 0, 0.15)))
  event <- et(time = 0, amt = 500)
  sim <- rxSolve(mod_fig, params, event %>% et(fig_times))
  as.data.frame(sim) %>% mutate(ID = factor(i))
}) %>% bind_rows()

# Summary statistics
fig_summary <- fig_data %>%
  group_by(time) %>%
  summarise(
    Mean = mean(conc),
    SD = sd(conc),
    Median = median(conc),
    .groups = "drop"
  )

# Publication-quality figure
ggplot() +
  # Individual profiles (semi-transparent)
  geom_line(data = fig_data, aes(time, conc, group = ID),
            color = "gray70", alpha = 0.4, linewidth = 0.5) +
  # Mean ± SD ribbon
  geom_ribbon(data = fig_summary,
              aes(time, ymin = Mean - SD, ymax = Mean + SD),
              fill = "#0B6E6B", alpha = 0.15) +
  # Mean line
  geom_line(data = fig_summary, aes(time, Mean),
            color = "#0B6E6B", linewidth = 1.2) +
  # Median line
  geom_line(data = fig_summary, aes(time, Median),
            color = "#C44536", linewidth = 1, linetype = "dashed") +
  scale_y_log10(
    breaks = c(0.1, 0.3, 1, 3, 10, 30),
    labels = c("0.1", "0.3", "1", "3", "10", "30")
  ) +
  labs(
    title = "Mean (±SD) Plasma Concentration-Time Profile",
    subtitle = "Semi-logarithmic scale; n = 8 subjects",
    x = "Time (hours)",
    y = expression("Concentration (mg/L)"),
    caption = "Solid line: arithmetic mean; Shaded band: ±1 SD; Dashed: median"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "gray50", size = 9),
    panel.grid.minor = element_blank()
  )
```

---

## Reproducible Analysis with Quarto

### Example Quarto Document Structure

````markdown
---
title: "PK Analysis Report — Study ABC-001"
author: "Pharmacometrics Team"
date: "2024-06-01"
format: html
execute:
  echo: false
  warning: false
---

```r setup
library(PKNCA)
library(ggplot2)
library(dplyr)
library(knitr)

# Load data
pk_data <- read.csv("data/pk_concentrations.csv")
```

## Noncompartmental Analysis

```r nca
# NCA computation
# ...
```

## Population PK Model

```r poppk
# Model fitting
# ...
```
````

---

## Common Reporting Standards

### Key Elements to Include

1. **Study design**: Sampling schedule, dose levels, population
2. **Software**: R version, package versions (use `sessionInfo()`)
3. **Data handling**: BLOQ treatment, outlier rules, exclusions
4. **Model development**: Steps, decisions, and rationale
5. **Parameter estimates**: With RSE%, CI, and shrinkage
6. **Diagnostics**: GOF, VPC, bootstrap
7. **Covariate effects**: Forest plots with clinical relevance
8. **Simulation details**: Number of replicates, random seed

```{r}
# Capture and report session information
# sessionInfo()
```

---

## Communicating Uncertainty

### Types of Uncertainty to Report

| Source | How to Report |
|--------|--------------|
| **Parameter uncertainty** | RSE%, confidence/credible intervals |
| **Model uncertainty** | Compare alternative models |
| **Prediction uncertainty** | Prediction intervals from simulation |
| **Assay uncertainty** | Report LLOQ, QC data |
| **Sampling uncertainty** | Note sparse vs rich design limitations |

```{r}
# Example: Clear communication of dose recommendation with uncertainty
cat("
Recommended dose: 500 mg q12h

Expected Css,avg: 10.4 mg/L (90% PI: 6.5–16.8 mg/L)
PTA (Cmin > 2 mg/L): 94.3% (95% CI: 91.2–96.7%)
PTA (Cmax < 50 mg/L): 99.1% (95% CI: 98.0–99.7%)

Note: Dose adjustment recommended for eGFR < 30 mL/min.
Simulation based on 2000 virtual subjects with PK variability
from the final population model.
")
```

---

## Building a Reusable Code Library

```{r}
# Template for a reusable PK function library
# Save as R/pk_functions.R

#' Calculate elimination rate constant from two concentration-time points
#'
#' @param c1 Concentration at time 1 (mg/L)
#' @param c2 Concentration at time 2 (mg/L)
#' @param t1 Time 1 (hours)
#' @param t2 Time 2 (hours)
#' @return Elimination rate constant (1/h)
#' @export
calc_ke <- function(c1, c2, t1, t2) {
  if (c2 >= c1) stop("c2 must be less than c1 (declining phase)")
  if (t2 <= t1) stop("t2 must be greater than t1")
  (log(c1) - log(c2)) / (t2 - t1)
}

#' Calculate half-life from ke
#'
#' @param ke Elimination rate constant (1/h)
#' @return Half-life (hours)
#' @export
calc_half_life <- function(ke) {
  log(2) / ke
}

#' Calculate accumulation factor
#'
#' @param ke Elimination rate constant (1/h)
#' @param tau Dosing interval (hours)
#' @return Accumulation factor
#' @export
calc_accumulation <- function(ke, tau) {
  1 / (1 - exp(-ke * tau))
}

# Example usage
# ke_est <- calc_ke(8.5, 2.1, 2, 8)
# t12 <- calc_half_life(ke_est)
```

---

## Regulatory Considerations

::: {.theory-box}
### FDA/EMA Expectations for PK/PD Analyses

- **Data standards**: SDTM/ADaM formats in submissions
- **Software**: R is accepted; document version and packages
- **Code submission**: Increasingly expected for reproducibility
- **Model documentation**: Complete model code, decision rationale
- **Sensitivity analyses**: Demonstrate robustness of conclusions
- **Define.xml**: Metadata for all datasets and variables
:::

---

::: {.exercise-box}
### Exercises

1. Create a Quarto template for a standard PK report with the sections listed above.
2. Generate a publication-quality parameter table with estimates, RSE%, and 95% CI.
3. Write a function library with utility functions for PK calculations (ke, half-life, AUC, accumulation).
4. Document a complete PK analysis workflow with commented R code that a colleague could reproduce.
:::

::: {.mistakes-box}
### Common Mistakes

- **Not documenting data exclusions**: Every excluded data point needs justification
- **Reporting only point estimates**: Always include uncertainty measures
- **Over-polishing figures before model is final**: Iterate on exploratory plots first
- **Not version-controlling analysis scripts**: Git + meaningful commit messages
- **Assuming readers understand your model**: Write for a clinical audience, not just modelers
:::

::: {.key-box}
### Key Takeaway

A good PK/PD analysis is only as valuable as its communication. Structure your reports clearly, quantify uncertainty honestly, and make every analysis reproducible. The best model is useless if nobody understands what it means for patients.
:::

