---
title: "Practice Datasets and Exercises with Solutions"
---


This appendix provides practice datasets and exercises with solutions for each major topic area.

---

## Dataset 1: Single-Dose Oral PK

```{r}
library(dplyr)

# Generate the dataset
set.seed(42)
practice1 <- data.frame(
  TIME = c(0, 0.25, 0.5, 1, 2, 3, 4, 6, 8, 12, 24),
  CONC = c(0, 1.2, 3.8, 7.5, 9.2, 8.1, 6.3, 4.1, 2.5, 1.1, 0.3)
)
```

### Exercise A1: NCA
Compute $C_{max}$, $T_{max}$, $AUC_{0 \to last}$, $AUC_{0 \to \infty}$, and $t_{1/2}$.

<details>
<summary>Solution</summary>

```{r}
nca <- practice1 %>%
  mutate(log_conc = log(CONC))

# Cmax
cmax <- max(nca$CONC)
tmax <- nca$TIME[which.max(nca$CONC)]

# AUC (linear trapezoidal)
auc_last <- sum(diff(nca$TIME) * (nca$CONC[-1] + nca$CONC[-nrow(nca)]) / 2)

# Terminal slope (last 4 points)
term_idx <- 8:11
lm_term <- lm(log_conc ~ TIME, data = nca[term_idx, ])
ke <- -coef(lm_term)[2]
t_half <- log(2) / ke

# AUC∞
auc_inf <- auc_last + nca$CONC[nrow(nca)] / ke

c(Cmax = cmax, Tmax = tmax, AUC_last = auc_last,
  AUC_inf = auc_inf, t_half = t_half, ke = ke)
```
</details>

---

## Dataset 2: IV Bolus with Noise

```{r}
set.seed(789)
practice2 <- data.frame(
  TIME = c(0.08, 0.25, 0.5, 1, 2, 4, 8, 12, 24),
  CONC = c(19.8, 16.2, 13.5, 10.1, 7.3, 4.5, 2.1, 1.1, 0.3)
)
```

### Exercise A2: One-Compartment Fit
Fit a one-compartment IV bolus model. Estimate $C_0$, $V$, $CL$.

<details>
<summary>Solution</summary>

```{r}
fit <- nls(CONC ~ C0 * exp(-ke * TIME), data = practice2,
           start = list(C0 = 20, ke = 0.3))
coef(fit)

# Assuming dose = 100 mg
C0 <- coef(fit)["C0"]
ke <- coef(fit)["ke"]
V <- 100 / C0
CL <- ke * V
c(V = V, CL = CL, t_half = log(2)/ke)
```
</details>

---

## Dataset 3: Dose-Response

```{r}
practice3 <- data.frame(
  DOSE = rep(c(0.1, 0.3, 1, 3, 10, 30, 100), each = 3),
  EFFECT = c(2, 4, 3, 8, 10, 12, 25, 28, 30, 48, 52, 50,
             72, 70, 75, 85, 88, 82, 95, 92, 98)
)
```

### Exercise A3: Emax Fit
Fit an Emax model and estimate $E_{max}$ and $EC_{50}$.

<details>
<summary>Solution</summary>

```{r}
fit_emax <- nls(EFFECT ~ E0 + (Emax * DOSE) / (EC50 + DOSE),
                data = practice3,
                start = list(E0 = 2, Emax = 90, EC50 = 5))
summary(fit_emax)
```
</details>

---

## Dataset 4: Population PK

```{r}
set.seed(555)
n_subj <- 8
id_times <- c(0.5, 1, 2, 4, 8, 12, 24)

practice4 <- expand.grid(ID = 1:n_subj, TIME = id_times) %>%
  mutate(
    DOSE = 500,
    CL = 5 * exp(rep(rnorm(n_subj, 0, 0.25), each = length(id_times))),
    V  = 50 * exp(rep(rnorm(n_subj, 0, 0.15), each = length(id_times))),
    ke = CL / V,
    CONC = (DOSE / V) * exp(-ke * TIME),
    CONC = CONC * exp(rnorm(n(), 0, 0.1))
  ) %>%
  arrange(ID, TIME)
```

### Exercise A4: Population Analysis
Compute individual $k_e$ via log-linear regression for each subject. Summarize mean and SD of $k_e$ and $V$.

<details>
<summary>Solution</summary>

```{r}
indiv_params <- practice4 %>%
  group_by(ID) %>%
  summarise(
    ke = -coef(lm(log(CONC) ~ TIME))[2],
    C0 = exp(coef(lm(log(CONC) ~ TIME))[1]),
    V = 500 / C0,
    .groups = "drop"
  )

indiv_params %>%
  summarise(
    ke_mean = mean(ke), ke_sd = sd(ke),
    V_mean = mean(V), V_sd = sd(V)
  )
```
</details>

---

## Exercise: Dose Optimization

Using a population model with $CL$ = 5 L/h (CV = 25%), $V$ = 50 L (CV = 15%), determine the dose needed for 90% PTA with target $C_{min} > 1$ mg/L at q12h dosing.

<details>
<summary>Solution</summary>

```{r}
set.seed(999)
n_sim <- 5000
CL_sim <- 5 * exp(rnorm(n_sim, 0, 0.25))
V_sim  <- 50 * exp(rnorm(n_sim, 0, 0.15))
ke_sim <- CL_sim / V_sim
tau <- 12
target <- 1

# Try different doses
for (d in seq(200, 2000, by = 10)) {
  Cmin <- (d / V_sim) * exp(-ke_sim * tau) / (1 - exp(-ke_sim * tau))
  pta <- mean(Cmin > target)
  if (pta >= 0.9) {
    cat("Dose for 90% PTA:", d, "mg\n")
    break
  }
}
```
</details>

---

## Additional Practice Problems

1. **Data cleaning**: Given a dataset with BLOQ values, missing times, and implausible concentrations, write R code to clean it.
2. **Model comparison**: Fit one- and two-compartment models. Compare with AIC and VPC.
3. **Covariate modeling**: Test whether body weight is a significant covariate on CL.
4. **DDI simulation**: Predict the PK change when a CYP3A4 inhibitor (Ki = 0.1 µM, I = 2 µM) is co-administered with a drug that is 90% metabolized by CYP3A4.
5. **TDM case**: A patient has a concentration of 3 mg/L at 6 h after a 500 mg IV dose. The population prior is CL ~ LN(log(5), 0.3), V ~ LN(log(50), 0.2). Estimate the patient's individual CL and V.

